---
title: "3.1 Descriptive Results for Risk-Set Matched Cohorts"
author: "Nemo Zhou"
date: "2026-07-18"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  3.1_descriptive_riskset_cohorts.Rmd
Author:  Nemo Zhou
Date started:      2026-06-29
Date last updated: 2026-07-18 (added primary pre-GLME QC gate)

Purpose:
  Matching-cohort descriptive EDA report for the risk-set matched analytic
  cohort families created by scripts 2.1-2.5:
    1. High- vs low/moderate-burden cancers
    2. Smoking-related cancers versus cancer-free controls
    3. Obesity-related cancers versus cancer-free controls
    4. Survival <=5 vs >5 years after diagnosis
    5. Overall incident cancer vs matched controls

  The matched data are long repeated-measures files, so descriptive tables are
  built at the matched-assignment level: one row per Analysis, Cohort,
  match_set, participant id, and role. This avoids inflating Table 1 by the
  number of repeated FI observations. Controls can still appear in multiple
  assignments because the risk-set design matches with replacement; those
  assignments are intentionally retained.

  This R Markdown report writes and displays:
    - cohort/group size and reuse summaries
    - FI observation support summaries before and after index
    - Table 1 comparing cancer cases and controls within each Analysis-Cohort
      in both tidy-long and side-by-side wide CSV formats
    - mean observed FI by relative-time bin for descriptive plotting
    - a faceted crude FI trajectory plot
    - overall matched case/control FI trajectory relative to assigned index time
    - overall matched case/control index-age overlap plot

Outputs:
  Results/cancer/visuals/3.1_descriptive_riskset_cohorts.html
  Results/cancer/data/3.1_riskset_assignment_counts.csv
  Results/cancer/data/3.1_riskset_reuse_summary.csv
  Results/cancer/data/3.1_riskset_fi_observation_summary.csv
  Results/cancer/data/3.1_riskset_table1_cases_vs_controls.csv
  Results/cancer/data/3.1_riskset_table1_cases_vs_controls_wide.csv
  Results/cancer/data/3.1_riskset_mean_fi_by_relative_time_bin.csv
  Results/cancer/data/3.1_overall_matched_fi_relative_to_index.csv
  Results/cancer/data/3.1_overall_index_age_distribution.csv
  Results/cancer/data/3.1_overall_pre_glme_qc_gate.csv
-->

```{r build-results, include=FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)


# -----------------------------------------------------------------------------
# 1. Paths and inputs
# -----------------------------------------------------------------------------

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
data_dir    <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "Results", "cancer", "data")
raw_path    <- file.path(data_dir, "FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")
overall_matched_path <- file.path(data_dir, "riskset_matched_overall_long.rds")

matched_files <- tibble::tribble(
  ~Analysis, ~matched_path,
  "High vs Low/Moderate Burden", file.path(data_dir, "riskset_matched_analysis_long.rds"),
  "Smoking-Related Cancer",      file.path(data_dir, "riskset_matched_smoking_long.rds"),
  "Obesity-Related Cancer",      file.path(data_dir, "riskset_matched_obesity_long.rds"),
  "Survival After Diagnosis",    file.path(data_dir, "riskset_matched_survival_long.rds"),
  "Overall Incident Cancer",     overall_matched_path
)

required_matched_cols <- c(
  "id", "match_set", "index_date", "role", "Cohort", "cycle",
  "age_at_cycle", "fi_score_nocancer", "base_race", "base_marital",
  "base_living", "base_pckgr", "index_age", "Age_Centered", "Post", "Group",
  "post_own_cancer"
)

if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)


# -----------------------------------------------------------------------------
# 2. Helpers
# -----------------------------------------------------------------------------

read_matched <- function(analysis, matched_path) {
  if (!file.exists(matched_path)) {
    stop("Matched dataset not found for ", analysis, ": ", matched_path)
  }
  dat <- readRDS(matched_path)
  missing_cols <- setdiff(required_matched_cols, names(dat))
  if (length(missing_cols) > 0) {
    stop(
      "Matched dataset for ", analysis, " is missing required columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }

  dat %>%
    mutate(
      Analysis = analysis,
      id = as.character(id),
      match_set = as.character(match_set),
      role = as.character(role),
      Group = factor(as.character(Group), levels = c("Control", "Cancer Case")),
      Cohort = as.character(Cohort),
      cycle = as.character(cycle),
      fi_score_nocancer = as.numeric(fi_score_nocancer),
      Age_Centered = as.numeric(Age_Centered),
      index_age = as.numeric(index_age),
      post_own_cancer = as.logical(post_own_cancer)
    )
}

first_nonmissing <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

last_nonmissing <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[length(x)]
}

mean_sd <- function(x, digits = 2) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  paste0(round(mean(x), digits), " (", round(stats::sd(x), digits), ")")
}

median_iqr <- function(x, digits = 2) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  q <- stats::quantile(x, probs = c(0.25, 0.5, 0.75), na.rm = TRUE, names = FALSE)
  paste0(round(q[2], digits), " [", round(q[1], digits), ", ", round(q[3], digits), "]")
}

format_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}

continuous_p <- function(x, group) {
  keep <- !is.na(x) & !is.na(group)
  x <- x[keep]
  group <- droplevels(factor(group[keep]))
  if (length(unique(group)) != 2 || length(x) < 3) return(NA_real_)
  tryCatch(stats::t.test(x ~ group)$p.value, error = function(e) NA_real_)
}

categorical_p <- function(x, group) {
  keep <- !is.na(x) & !is.na(group)
  x <- factor(x[keep])
  group <- droplevels(factor(group[keep]))
  if (length(unique(group)) != 2 || nlevels(x) < 2) return(NA_real_)
  tab <- table(x, group)
  tryCatch({
    if (any(tab < 5)) stats::fisher.test(tab)$p.value else stats::chisq.test(tab)$p.value
  }, error = function(e) NA_real_)
}

smd_continuous <- function(x, group) {
  keep <- !is.na(x) & !is.na(group)
  x <- x[keep]
  group <- droplevels(factor(group[keep], levels = c("Control", "Cancer Case")))
  if (length(unique(group)) != 2) return(NA_real_)
  x0 <- x[group == "Control"]
  x1 <- x[group == "Cancer Case"]
  pooled_sd <- sqrt((stats::var(x0) + stats::var(x1)) / 2)
  if (is.na(pooled_sd) || pooled_sd == 0) return(NA_real_)
  (mean(x1) - mean(x0)) / pooled_sd
}

summarize_continuous <- function(dat, var, label) {
  x <- dat[[var]]
  p <- continuous_p(x, dat$Group)
  smd <- smd_continuous(x, dat$Group)
  dat %>%
    group_by(Group) %>%
    summarize(
      Statistic = mean_sd(.data[[var]]),
      Median_IQR = median_iqr(.data[[var]]),
      Missing = sum(is.na(.data[[var]])),
      .groups = "drop"
    ) %>%
    mutate(
      Variable = label,
      Level = NA_character_,
      P_Value = format_p(p),
      SMD = round(smd, 3)
    )
}

summarize_categorical <- function(dat, var, label) {
  p <- categorical_p(dat[[var]], dat$Group)
  denom <- dat %>%
    group_by(Group) %>%
    summarize(Denom = sum(!is.na(.data[[var]])), .groups = "drop")

  dat %>%
    filter(!is.na(.data[[var]])) %>%
    count(Group, Level = .data[[var]], name = "n") %>%
    left_join(denom, by = "Group") %>%
    mutate(
      Level = as.character(Level),
      Statistic = paste0(n, " (", round(100 * n / Denom, 1), "%)"),
      Median_IQR = NA_character_,
      Missing = NA_integer_,
      Variable = label,
      P_Value = format_p(p),
      SMD = NA_real_
    ) %>%
    select(Group, Statistic, Median_IQR, Missing, Variable, Level, P_Value, SMD)
}

make_table1_one <- function(dat, continuous_vars, categorical_vars) {
  cont <- bind_rows(lapply(names(continuous_vars), function(v) {
    summarize_continuous(dat, v, continuous_vars[[v]])
  }))
  catg <- bind_rows(lapply(names(categorical_vars), function(v) {
    summarize_categorical(dat, v, categorical_vars[[v]])
  }))

  bind_rows(cont, catg) %>%
    select(Variable, Level, Group, Statistic, Median_IQR, Missing, P_Value, SMD)
}


# -----------------------------------------------------------------------------
# 3. Load matched datasets and optional 1986 baseline covariates
# -----------------------------------------------------------------------------

matched_long_raw <- bind_rows(lapply(seq_len(nrow(matched_files)), function(i) {
  read_matched(matched_files$Analysis[i], matched_files$matched_path[i])
}))

overall_pre_glme_qc_gate <- matched_long_raw %>%
  filter(Analysis == "Overall Incident Cancer", Cohort == "All Cancer Cohort") %>%
  group_by(Group) %>%
  summarize(
    n_assignments_before_censoring = n_distinct(trajectory_id),
    n_distinct_persons = n_distinct(id),
    n_rows_before_censoring = n(),
    n_rows_removed_post_own_cancer = sum(post_own_cancer),
    n_assignments_with_rows_removed = n_distinct(trajectory_id[post_own_cancer]),
    n_rows_after_primary_censoring = sum(!post_own_cancer),
    .groups = "drop"
  )

overall_yield_path <- file.path(results_dir, "matching_diagnostics",
                                 "riskset_matched_overall_long_yield.csv")
if (file.exists(overall_yield_path)) {
  overall_yield <- read.csv(overall_yield_path, stringsAsFactors = FALSE)
  if (nrow(overall_yield) == 1) {
    overall_pre_glme_qc_gate <- overall_pre_glme_qc_gate %>%
      mutate(cases_with_controls_from_yield = if_else(
        Group == "Cancer Case", overall_yield$cases_with_controls[[1]], NA_integer_
      ))
  }
}

matched_long <- matched_long_raw %>%
  filter(!post_own_cancer)

baseline_extra <- NULL
if (file.exists(raw_path)) {
  raw <- readRDS(raw_path)
  available_extra <- intersect(c("bmi", "smoke", "pckyr", "act", "calor", "alco"), names(raw))
  baseline_extra <- raw %>%
    mutate(id = as.character(id), cycle = as.character(cycle)) %>%
    filter(cycle == "86") %>%
    distinct(id, .keep_all = TRUE) %>%
    transmute(
      id,
      base_bmi = if ("bmi" %in% available_extra) as.numeric(as.character(bmi)) else NA_real_,
      base_smoke = if ("smoke" %in% available_extra) as.character(smoke) else NA_character_,
      base_pack_years = if ("pckyr" %in% available_extra) as.numeric(as.character(pckyr)) else NA_real_,
      base_activity = if ("act" %in% available_extra) as.numeric(as.character(act)) else NA_real_,
      base_calories = if ("calor" %in% available_extra) as.numeric(as.character(calor)) else NA_real_,
      base_alcohol = if ("alco" %in% available_extra) as.numeric(as.character(alco)) else NA_real_
    )
} else {
  warning("Raw FI dataset not found; Table 1 will use matched-dataset covariates only.")
}


# -----------------------------------------------------------------------------
# 4. Assignment-level analytic summaries
# -----------------------------------------------------------------------------

assignment_summary <- matched_long %>%
  arrange(Analysis, Cohort, match_set, id, role, Age_Centered) %>%
  group_by(Analysis, Cohort, match_set, id, role, Group) %>%
  summarize(
    index_date = first(index_date),
    index_age = first(index_age),
    n_fi_observations = n(),
    n_pre_index_fi = sum(Age_Centered < 0),
    n_post_index_fi = sum(Age_Centered >= 0),
    min_relative_time = min(Age_Centered, na.rm = TRUE),
    max_relative_time = max(Age_Centered, na.rm = TRUE),
    fi_first_observed = first_nonmissing(fi_score_nocancer),
    fi_last_pre_index = last_nonmissing(fi_score_nocancer[Age_Centered < 0]),
    fi_first_post_index = first_nonmissing(fi_score_nocancer[Age_Centered >= 0]),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    .groups = "drop"
  ) %>%
  mutate(
    has_pre_index_fi = n_pre_index_fi > 0,
    has_post_index_fi = n_post_index_fi > 0
  )

if (!is.null(baseline_extra)) {
  assignment_summary <- assignment_summary %>%
    left_join(baseline_extra, by = "id")
}


# -----------------------------------------------------------------------------
# 5. Cohort sizes, reuse, and FI observation support
# -----------------------------------------------------------------------------

assignment_counts <- assignment_summary %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    n_assignments = n(),
    n_distinct_persons = n_distinct(id),
    n_match_sets = n_distinct(match_set),
    mean_index_age = mean(index_age, na.rm = TRUE),
    sd_index_age = stats::sd(index_age, na.rm = TRUE),
    .groups = "drop"
  )

reuse_summary <- assignment_summary %>%
  group_by(Analysis, Cohort, Group, id) %>%
  summarize(n_assignments_for_person = n(), .groups = "drop") %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    n_persons = n(),
    mean_assignments_per_person = mean(n_assignments_for_person),
    max_assignments_per_person = max(n_assignments_for_person),
    persons_reused_more_than_once = sum(n_assignments_for_person > 1),
    pct_persons_reused_more_than_once = 100 * mean(n_assignments_for_person > 1),
    .groups = "drop"
  )

fi_observation_summary <- assignment_summary %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    mean_fi_observations = mean(n_fi_observations),
    median_fi_observations = stats::median(n_fi_observations),
    mean_pre_index_fi = mean(n_pre_index_fi),
    mean_post_index_fi = mean(n_post_index_fi),
    pct_with_pre_index_fi = 100 * mean(has_pre_index_fi),
    pct_with_post_index_fi = 100 * mean(has_post_index_fi),
    mean_min_relative_time = mean(min_relative_time),
    mean_max_relative_time = mean(max_relative_time),
    .groups = "drop"
  )


# -----------------------------------------------------------------------------
# 6. Table 1 comparing cases and controls within each Analysis-Cohort
# -----------------------------------------------------------------------------

continuous_vars <- c(
  index_age = "Index age, years",
  fi_first_observed = "FI, first observed analytic cycle",
  fi_last_pre_index = "FI, last pre-index observation",
  fi_first_post_index = "FI, first post-index observation",
  n_fi_observations = "Number of observed FI measurements",
  n_pre_index_fi = "Number of pre-index FI measurements",
  n_post_index_fi = "Number of post-index FI measurements"
)

if ("base_bmi" %in% names(assignment_summary)) {
  continuous_vars <- c(
    continuous_vars,
    base_bmi = "Baseline BMI",
    base_pack_years = "Baseline pack-years",
    base_activity = "Baseline physical activity",
    base_calories = "Baseline calories",
    base_alcohol = "Baseline alcohol"
  )
}

categorical_vars <- c(
  base_race = "Baseline race",
  base_marital = "Baseline marital status",
  base_living = "Baseline living arrangement",
  base_pckgr = "Baseline pack-year group"
)

if ("base_smoke" %in% names(assignment_summary)) {
  categorical_vars <- c(categorical_vars, base_smoke = "Baseline smoking status")
}

table1 <- assignment_summary %>%
  group_by(Analysis, Cohort) %>%
  group_modify(~ make_table1_one(.x, continuous_vars, categorical_vars)) %>%
  ungroup() %>%
  arrange(Analysis, Cohort, Variable, is.na(Level), Level, Group)

table1_wide <- table1 %>%
  mutate(
    Group = as.character(Group),
    Group = if_else(Group == "Cancer Case", "Cancer_Case", Group)
  ) %>%
  pivot_wider(
    id_cols = c(Analysis, Cohort, Variable, Level, P_Value, SMD),
    names_from = Group,
    values_from = c(Statistic, Median_IQR, Missing),
    names_glue = "{Group}_{.value}"
  ) %>%
  arrange(Analysis, Cohort, Variable, is.na(Level), Level)


# -----------------------------------------------------------------------------
# 7. Descriptive observed FI trajectories by relative-time bin
# -----------------------------------------------------------------------------

fi_by_time_bin <- matched_long %>%
  mutate(
    relative_time_bin = cut(
      Age_Centered,
      breaks = c(-Inf, seq(-20, 20, by = 2), Inf),
      labels = c("<= -20", paste(seq(-20, 18, by = 2), seq(-18, 20, by = 2), sep = " to "), "> 20"),
      right = FALSE
    ),
    relative_time_bin_label = as.character(relative_time_bin),
    relative_time_mid = case_when(
      relative_time_bin_label == "<= -20" ~ -21,
      relative_time_bin_label == "> 20" ~ 21,
      TRUE ~ suppressWarnings(as.numeric(sub(" to .*", "", relative_time_bin_label))) + 1
    )
  ) %>%
  group_by(Analysis, Cohort, Group, relative_time_bin, relative_time_mid) %>%
  summarize(
    n_observations = n(),
    n_assignments = n_distinct(paste(match_set, id, role, sep = "|")),
    n_persons = n_distinct(id),
    mean_fi = mean(fi_score_nocancer, na.rm = TRUE),
    sd_fi = stats::sd(fi_score_nocancer, na.rm = TRUE),
    se_fi = sd_fi / sqrt(n_observations),
    .groups = "drop"
  )

group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")

p_fi <- ggplot(fi_by_time_bin, aes(x = relative_time_mid, y = mean_fi, color = Group)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  facet_grid(Analysis ~ Cohort, scales = "free_y") +
  scale_color_manual(values = group_cols) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Observed Mean Frailty Index by Relative-Time Bin",
    x = "Years relative to assigned index time",
    y = "Mean no-cancer frailty index",
    color = NULL
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 8)
  )


# -----------------------------------------------------------------------------
# 8. Overall matched index-time EDA
# -----------------------------------------------------------------------------

overall_matched_long <- matched_long %>%
  filter(Analysis == "Overall Incident Cancer") %>%
  mutate(
    rel_year_bin = round(Age_Centered / 2) * 2
  ) %>%
  filter(rel_year_bin >= -20, rel_year_bin <= 20)

overall_matched_relative_fi <- overall_matched_long %>%
  group_by(Group, rel_year_bin) %>%
  summarize(
    n = sum(!is.na(fi_score_nocancer)),
    n_participants = n_distinct(id),
    n_assignments = n_distinct(paste(match_set, id, role, sep = "__")),
    mean_fi = mean(fi_score_nocancer, na.rm = TRUE),
    sd_fi = stats::sd(fi_score_nocancer, na.rm = TRUE),
    se_fi = sd_fi / sqrt(n),
    lcl = mean_fi - 1.96 * se_fi,
    ucl = mean_fi + 1.96 * se_fi,
    median_fi = stats::median(fi_score_nocancer, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Group, rel_year_bin)

p_overall_matched_relative_fi <- ggplot(
  overall_matched_relative_fi,
  aes(x = rel_year_bin, y = mean_fi, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray35") +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Overall Matched FI Trajectories Relative to Assigned Index Time",
    x = "Years relative to assigned index time",
    y = "Mean no-cancer frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme(legend.position = "bottom")

overall_index_age_distribution <- matched_long %>%
  filter(Analysis == "Overall Incident Cancer") %>%
  mutate(
    assignment_id = paste(match_set, id, role, sep = "__")
  ) %>%
  distinct(assignment_id, Group, index_age) %>%
  filter(!is.na(index_age))

overall_index_age_summary <- overall_index_age_distribution %>%
  group_by(Group) %>%
  summarize(
    n_assignments = n(),
    mean_index_age = mean(index_age),
    sd_index_age = stats::sd(index_age),
    median_index_age = stats::median(index_age),
    q1_index_age = stats::quantile(index_age, 0.25),
    q3_index_age = stats::quantile(index_age, 0.75),
    min_index_age = min(index_age),
    max_index_age = max(index_age),
    .groups = "drop"
  )

p_overall_index_age <- ggplot(
  overall_index_age_distribution,
  aes(x = index_age, fill = Group, color = Group)
) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = group_cols) +
  scale_color_manual(values = group_cols) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Overall Matched Assigned Index-Age Distribution",
    x = "Assigned index age, years",
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme(legend.position = "bottom")


# -----------------------------------------------------------------------------
# 9. Save outputs
# -----------------------------------------------------------------------------

write.csv(assignment_counts,
          file.path(results_dir, "3.1_riskset_assignment_counts.csv"),
          row.names = FALSE)
write.csv(reuse_summary,
          file.path(results_dir, "3.1_riskset_reuse_summary.csv"),
          row.names = FALSE)
write.csv(fi_observation_summary,
          file.path(results_dir, "3.1_riskset_fi_observation_summary.csv"),
          row.names = FALSE)
write.csv(table1,
          file.path(results_dir, "3.1_riskset_table1_cases_vs_controls.csv"),
          row.names = FALSE)
write.csv(table1_wide,
          file.path(results_dir, "3.1_riskset_table1_cases_vs_controls_wide.csv"),
          row.names = FALSE)
write.csv(fi_by_time_bin,
          file.path(results_dir, "3.1_riskset_mean_fi_by_relative_time_bin.csv"),
          row.names = FALSE)
write.csv(overall_matched_relative_fi,
          file.path(results_dir, "3.1_overall_matched_fi_relative_to_index.csv"),
          row.names = FALSE)
write.csv(overall_index_age_summary,
          file.path(results_dir, "3.1_overall_index_age_distribution.csv"),
          row.names = FALSE)
write.csv(overall_pre_glme_qc_gate,
          file.path(results_dir, "3.1_overall_pre_glme_qc_gate.csv"),
          row.names = FALSE)
cat("\nDescriptive analysis complete.\n")
cat("Assignment-level rows:", nrow(assignment_summary), "\n")
cat("Saved visual-data tables to:", results_dir, "\n")
```

# Assignment Counts

```{r assignment-counts, echo=FALSE}
knitr::kable(assignment_counts)
```

# Control Reuse Summary

```{r reuse-summary, echo=FALSE}
knitr::kable(reuse_summary)
```

# FI Observation Support

```{r fi-observation-summary, echo=FALSE}
knitr::kable(fi_observation_summary)
```

# Overall Pre-GLME QC Gate

The primary display censors a control's measurements at or after that control's
own later cancer diagnosis. This does not resolve informative death or loss to
follow-up; post-index estimates still require IPCW and joint-model sensitivity
interpretation.

```{r overall-pre-glme-qc, echo=FALSE}
knitr::kable(overall_pre_glme_qc_gate)
```

# Table 1: Cases Versus Controls

```{r table1-wide, echo=FALSE}
knitr::kable(table1_wide)
```

# Mean FI by Relative-Time Bin

```{r fi-by-time-bin-table, echo=FALSE}
knitr::kable(fi_by_time_bin)
```

```{r crude-fi-by-time-bin-plot, echo=FALSE, fig.width=14, fig.height=10}
p_fi
```

# Overall Matched FI Relative to Index Time

```{r overall-relative-fi-table, echo=FALSE}
knitr::kable(overall_matched_relative_fi)
```

```{r overall-relative-fi-plot, echo=FALSE, fig.width=8.5, fig.height=5.5}
p_overall_matched_relative_fi
```

# Overall Assigned Index-Age Distribution

```{r overall-index-age-table, echo=FALSE}
knitr::kable(overall_index_age_summary)
```

```{r overall-index-age-plot, echo=FALSE, fig.width=8.5, fig.height=5.5}
p_overall_index_age
```

# Output Inventory

```{r output-inventory, echo=FALSE}
outputs <- c(
  "3.1_riskset_assignment_counts.csv",
  "3.1_riskset_reuse_summary.csv",
  "3.1_riskset_fi_observation_summary.csv",
  "3.1_riskset_table1_cases_vs_controls.csv",
  "3.1_riskset_table1_cases_vs_controls_wide.csv",
  "3.1_riskset_mean_fi_by_relative_time_bin.csv",
  "3.1_overall_matched_fi_relative_to_index.csv",
  "3.1_overall_index_age_distribution.csv",
  "3.1_overall_pre_glme_qc_gate.csv"
)

knitr::kable(
  tibble::tibble(
    output = outputs,
    path = file.path(results_dir, outputs),
    exists = file.exists(file.path(results_dir, outputs))
  )
)
```
