---
title: "3.3 Matched Cohort Availability, Frailty Distributions, and Attrition"
author: "Nemo Zhou"
date: "2026-07-27"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  3.3_matched_cohort_availability_attrition.Rmd
Author:  Nemo Zhou
Date started:      2026-07-27
Date last updated: 2026-07-27 (separated death, later own-cancer,
                   loss-to-follow-up, and administrative censoring)

Purpose:
  Descriptive report for the four primary risk-set matched cancer cohort
  families:
    1. High-burden and Low/Moderate-Burden cancer cohorts
    2. Smoking-related cancer cohort
    3. Obesity-related cancer cohort
    4. Overall incident-cancer cohort

  This report recreates the assignment-count and control-reuse definitions
  reported by 3.1 from the current matched inputs, then combines them with
  matched-assignment-level FI availability, FI distributions, and
  distinct follow-up endpoint summaries. The availability and distribution
  sections rebuild a complete matched-assignment x 4-year-cycle grid from the
  canonical FI data, because the long matched RDS files retain observed FI
  rows and therefore cannot by themselves describe FI non-observation.

  Follow-up endpoints are summarized as mutually exclusive categories:
  death before the administrative endpoint, later own-cancer censoring among
  controls, loss to follow-up/nonresponse before 2020, and administrative
  censoring with an observed 2020 FI. Endpoint timing is displayed relative to
  assigned index age. Reused controls remain repeated assignments, consistent
  with the risk-set design.

Inputs:
  Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds
  Data/riskset_matched_analysis_long.rds
  Data/riskset_matched_smoking_long.rds
  Data/riskset_matched_obesity_long.rds
  Data/riskset_matched_overall_long.rds
  The current matched RDS files are authoritative for the 3.1-style
  assignment/reuse summaries; older 3.1 CSVs are not read because they may
  predate a matched-input rebuild.

Outputs:
  Results/cancer/data/3.3_fi_availability_by_cycle.csv
  Results/cancer/data/3.3_active_participants_vs_usable_fi.csv
  Results/cancer/data/3.3_assignment_counts.csv
  Results/cancer/data/3.3_reuse_summary.csv
  Results/cancer/data/3.3_fi_distribution_by_cycle.csv
  Results/cancer/data/3.3_fi_distribution_by_age_group.csv
  Results/cancer/data/3.3_death_dropout_relative_to_index_age.csv
  Results/cancer/data/3.3_attrition_event_summary.csv
  Results/cancer/data/3.3_censoring_category_summary.csv
  Results/cancer/data/3.3_censoring_category_by_relative_index_age.csv
  Results/cancer/visuals/3.3_matched_cohort_availability_attrition.html

Interpretation notes:
  The FI availability and distribution summaries are assignment-level
  summaries. Thus, a reused control contributes once per matched assignment.
  Percentages are descriptive properties of the matched assignment design and
  are not independent-person estimates. Controls' rows at or after a later
  own-cancer diagnosis are excluded, matching the primary censoring rule used
  in 3.1 and the downstream trajectory analyses.

  The availability table reports broad FI non-observation among all analytic
  assignments. The separate pre-matching plot compares all active participants
  with participants having a valid overall FI score.

  Age groups are <60, 60-69, 70-79, and 80+ years at the analytic FI cycle.
  Censoring plots use -20 through +20 years relative to assigned index age,
  matching the main descriptive relative-time support window; values beyond
  20 years in either direction are not displayed in the cumulative plot.
  Administrative censoring is not interpreted as loss to follow-up.
-->

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = file.path(tempdir(), "3.3_matched_cohort_figures", "")
)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
data_dir <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "Results", "cancer", "data")

raw_path <- file.path(data_dir, "FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

matched_specs <- tibble::tribble(
  ~Analysis, ~matched_path,
  "High vs Low/Moderate Burden", file.path(data_dir, "riskset_matched_analysis_long.rds"),
  "Smoking-Related Cancer",      file.path(data_dir, "riskset_matched_smoking_long.rds"),
  "Obesity-Related Cancer",      file.path(data_dir, "riskset_matched_obesity_long.rds"),
  "Overall Incident Cancer",     file.path(data_dir, "riskset_matched_overall_long.rds")
)

analytic_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
cycle_lookup <- tibble(
  cycle = analytic_cycles,
  cycle_year = c(1988, 1992, 1996, 2000, 2004, 2008, 2012, 2016, 2020)
)

cohort_levels <- c(
  "High Burden Cohort",
  "Low/Moderate Burden Cohort",
  "Smoking-Related Cancer Cohort",
  "Obesity-Related Cancer Cohort",
  "All Cancer Cohort"
)

group_levels <- c("Control", "Cancer Case")
group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")
censoring_category_levels <- c(
  "Death before administrative end",
  "Later own-cancer censoring",
  "Loss to follow-up/nonresponse before 2020",
  "Administrative censoring at 2020",
  "No observed FI"
)
censoring_category_cols <- c(
  "Death before administrative end" = "#7A0177",
  "Later own-cancer censoring" = "#D95F0E",
  "Loss to follow-up/nonresponse before 2020" = "#3182BD",
  "Administrative censoring at 2020" = "#31A354",
  "No observed FI" = "#969696"
)
theme_set(theme_minimal(base_size = 11))
```

```{r helpers}
first_nonmissing <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

write_csv_base <- function(x, path) {
  utils::write.csv(x, path, row.names = FALSE, na = "")
  invisible(x)
}

read_matched <- function(analysis, path) {
  if (!file.exists(path)) stop("Matched dataset not found for ", analysis, ": ", path)
  dat <- readRDS(path)
  required_cols <- c(
    "id", "match_set", "trajectory_id", "role", "Cohort", "Group",
    "index_date", "index_age", "post_own_cancer"
  )
  missing_cols <- setdiff(required_cols, names(dat))
  if (length(missing_cols) > 0) {
    stop(
      "Matched dataset for ", analysis, " is missing required columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }

  dat %>%
    transmute(
      Analysis = analysis,
      id = as.character(id),
      match_set = as.character(match_set),
      trajectory_id = as.character(trajectory_id),
      role = as.character(role),
      Cohort = as.character(Cohort),
      Group = as.character(Group),
      index_date = as.numeric(index_date),
      index_age = as.numeric(index_age),
      post_own_cancer = as.logical(post_own_cancer)
    )
}

age_group_levels <- c("<60", "60-69", "70-79", "80+")
```

# 3.1 Assignment Counts and Control Reuse

The following tables reproduce the assignment-level definitions from `3.1_descriptive_riskset_cohorts.Rmd` using the current matched RDS inputs. This avoids mixing a rebuilt matched input with older saved 3.1 CSV summaries. The survival-stratified cohort is intentionally excluded.

```{r load-31-summaries}
matched_long <- bind_rows(lapply(seq_len(nrow(matched_specs)), function(i) {
  read_matched(matched_specs$Analysis[i], matched_specs$matched_path[i])
}))

assignment_summary <- matched_long %>%
  filter(!post_own_cancer) %>%
  distinct(Analysis, Cohort, match_set, id, role, Group, index_date, index_age) %>%
  mutate(
    assignment_id = paste(Analysis, Cohort, match_set, id, role, sep = "__")
  )

assignment_counts <- assignment_summary %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    n_assignments = n(),
    n_distinct_persons = n_distinct(id),
    n_match_sets = n_distinct(match_set),
    mean_index_age = mean(index_age, na.rm = TRUE),
    sd_index_age = stats::sd(index_age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Analysis = factor(Analysis, levels = matched_specs$Analysis),
    Cohort = factor(Cohort, levels = cohort_levels),
    Group = factor(Group, levels = group_levels)
  ) %>%
  arrange(Analysis, Cohort, Group)

reuse_summary <- assignment_summary %>%
  group_by(Analysis, Cohort, Group, id) %>%
  summarize(n_assignments_for_person = n(), .groups = "drop") %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    n_persons = n(),
    mean_assignments_per_person = mean(n_assignments_for_person),
    max_assignments_per_person = max(n_assignments_for_person),
    persons_reused_more_than_once = sum(n_assignments_for_person > 1),
    pct_persons_reused_more_than_once = 100 * mean(n_assignments_for_person > 1),
    .groups = "drop"
  ) %>%
  mutate(
    Analysis = factor(Analysis, levels = matched_specs$Analysis),
    Cohort = factor(Cohort, levels = cohort_levels),
    Group = factor(Group, levels = group_levels)
  ) %>%
  arrange(Analysis, Cohort, Group)

write_csv_base(assignment_counts,
               file.path(results_dir, "3.3_assignment_counts.csv"))
write_csv_base(reuse_summary,
               file.path(results_dir, "3.3_reuse_summary.csv"))

assignment_counts_display <- assignment_counts %>%
  group_by(Analysis, Cohort) %>%
  mutate(
    `Assignments, n (%)` = paste0(
      n_assignments, " (", round(100 * n_assignments / sum(n_assignments), 1), "%)"
    )
  ) %>%
  ungroup() %>%
  transmute(
    Analysis, Cohort, Group,
    `Assignments, n (%)`,
    `Distinct persons, n` = n_distinct_persons,
    `Match sets, n` = n_match_sets,
    `Mean index age` = round(mean_index_age, 1),
    `SD index age` = round(sd_index_age, 1)
  )

knitr::kable(assignment_counts_display, digits = 2)
```

```{r reuse-table, echo=FALSE}
reuse_summary_display <- reuse_summary %>%
  transmute(
    Analysis, Cohort, Group,
    `Persons, n` = n_persons,
    `Mean assignments per person` = round(mean_assignments_per_person, 2),
    `Maximum assignments per person` = max_assignments_per_person,
    `Reused >1 assignment, n (%)` = paste0(
      persons_reused_more_than_once, " (",
      round(pct_persons_reused_more_than_once, 1), "%)"
    )
  )

knitr::kable(reuse_summary_display, digits = 2)
```

# Construct the Matched Assignment-by-Cycle Panel

The matched RDS files are used only as existing analytic inputs; this report does not rebuild any risk sets. A complete 1988-2020 four-year cycle grid is created by joining each distinct matched assignment to the canonical FI data. This is necessary to count FI non-observation after participants stop contributing observed FI values.

```{r load-and-construct-panel}
assignment_meta <- matched_long %>%
  filter(!post_own_cancer) %>%
  distinct(Analysis, Cohort, match_set, id, role, Group, index_date, index_age) %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_levels),
    Group = factor(Group, levels = group_levels),
    assignment_id = paste(Analysis, Cohort, match_set, id, role, sep = "__")
  )

if (!file.exists(raw_path)) stop("Canonical FI dataset not found: ", raw_path)
raw_fi <- readRDS(raw_path)

required_raw_cols <- c(
  "id", "cycle", "participated", "worked_rtmnyr", "dbmy09", "fi_score_nocancer",
  "n_answered_nocancer", "age_at_cycle", "cancer_index_dateca", "dtdth",
  "age_at_death", "death_status"
)
missing_raw_cols <- setdiff(required_raw_cols, names(raw_fi))
if (length(missing_raw_cols) > 0) {
  stop("Canonical FI dataset is missing required columns: ",
       paste(missing_raw_cols, collapse = ", "))
}

raw_4yr <- raw_fi %>%
  mutate(
    id = as.character(id),
    cycle = as.character(cycle),
    participated = as.numeric(participated),
    worked_rtmnyr = as.numeric(worked_rtmnyr),
    dbmy09 = as.numeric(dbmy09),
    fi_score_nocancer = as.numeric(fi_score_nocancer),
    n_answered_nocancer = as.numeric(n_answered_nocancer),
    age_at_cycle = as.numeric(age_at_cycle),
    dtdth = as.numeric(dtdth)
  ) %>%
  filter(cycle %in% analytic_cycles) %>%
  distinct(id, cycle, .keep_all = TRUE) %>%
  select(
    id, cycle, participated, worked_rtmnyr, dbmy09, fi_score_nocancer,
    n_answered_nocancer, age_at_cycle, dtdth
  )

participant_cancer <- raw_fi %>%
  transmute(
    id = as.character(id),
    own_cancer_month = as.numeric(cancer_index_dateca)
  ) %>%
  group_by(id) %>%
  summarize(
    own_cancer_month = first_nonmissing(own_cancer_month),
    .groups = "drop"
  )

cycle_timing <- raw_fi %>%
  transmute(
    cycle = as.character(cycle),
    participated = as.numeric(participated),
    worked_rtmnyr = as.numeric(worked_rtmnyr)
  ) %>%
  filter(cycle %in% analytic_cycles, participated == 1, !is.na(worked_rtmnyr)) %>%
  group_by(cycle) %>%
  summarize(cycle_end_month = max(worked_rtmnyr), .groups = "drop")

cycle_lookup <- cycle_lookup %>%
  left_join(cycle_timing, by = "cycle")

if (anyNA(cycle_lookup$cycle_end_month)) {
  stop("At least one analytic cycle lacks a valid administrative end month.")
}
administrative_end_year <- max(cycle_lookup$cycle_year)
administrative_end_month <- max(cycle_lookup$cycle_end_month)

matched_assignment_panel <- assignment_meta %>%
  tidyr::crossing(cycle_lookup) %>%
  left_join(raw_4yr, by = c("id", "cycle")) %>%
  left_join(participant_cancer, by = "id") %>%
  mutate(
    cycle_year = as.numeric(cycle_year),
    Age_Centered = age_at_cycle - index_age,
    analytic_time_month = coalesce(worked_rtmnyr, cycle_end_month),
    post_own_cancer = as.character(Group) == "Control" &
      !is.na(own_cancer_month) & !is.na(index_date) &
      own_cancer_month > index_date &
      analytic_time_month >= own_cancer_month,
    analytic_row = !post_own_cancer,
    source_row = analytic_row & !is.na(participated) & participated == 1,
    age_group = cut(
      age_at_cycle,
      breaks = c(-Inf, 60, 70, 80, Inf),
      right = FALSE,
      labels = age_group_levels
    )
  )

if (anyDuplicated(matched_assignment_panel[c("assignment_id", "cycle")])) {
  stop("The reconstructed panel has duplicate assignment-by-cycle rows.")
}

censoring_reentry <- matched_assignment_panel %>%
  filter(as.character(Group) == "Control") %>%
  arrange(Analysis, Cohort, assignment_id, cycle_year) %>%
  group_by(Analysis, Cohort, assignment_id) %>%
  summarize(
    post_cancer_reentry = any(diff(as.integer(post_own_cancer)) < 0),
    .groups = "drop"
  )

if (any(censoring_reentry$post_cancer_reentry)) {
  stop("Later own-cancer censoring is not absorbing for at least one control assignment.")
}
```

# FI Availability and Missingness by Analytic Cycle

The denominator for `FI not observed` is the number of non-censored matched assignments represented at that cycle. It includes assignments with an observed FI score, a missing FI score, or no usable cycle-level source row. Therefore, the percentage measures non-observation among all analytic assignments, not missingness among respondents only. The separate pre-matching plot below compares all participants who participated in the cycle (`participated == 1`) with those having a valid overall FI score. Missing individual FI components do not count as unusable FI when the overall score is still calculable.

```{r fi-availability}
fi_availability <- matched_assignment_panel %>%
  group_by(Analysis, Cohort, Group, cycle, cycle_year) %>%
  summarize(
    n_assignments_at_cycle = sum(analytic_row),
    n_distinct_persons_at_cycle = n_distinct(id[analytic_row]),
    n_no_source_rows = sum(analytic_row & !source_row),
    n_fi_observed = sum(source_row & !is.na(fi_score_nocancer)),
    n_fi_not_observed = sum(analytic_row & is.na(fi_score_nocancer)),
    percent_fi_not_observed = if_else(
      n_assignments_at_cycle > 0,
      100 * n_fi_not_observed / n_assignments_at_cycle,
      NA_real_
    ),
    median_n_answered = stats::median(n_answered_nocancer[source_row], na.rm = TRUE),
    q1_n_answered = stats::quantile(
      n_answered_nocancer[source_row], 0.25, na.rm = TRUE, names = FALSE
    ),
    q3_n_answered = stats::quantile(
      n_answered_nocancer[source_row], 0.75, na.rm = TRUE, names = FALSE
    ),
    .groups = "drop"
  ) %>%
  mutate(
    median_n_answered = if_else(is.nan(median_n_answered), NA_real_, median_n_answered),
    q1_n_answered = if_else(is.nan(q1_n_answered), NA_real_, q1_n_answered),
    q3_n_answered = if_else(is.nan(q3_n_answered), NA_real_, q3_n_answered)
  ) %>%
  arrange(Analysis, Cohort, Group, cycle_year)

write_csv_base(fi_availability,
               file.path(results_dir, "3.3_fi_availability_by_cycle.csv"))

fi_availability_display <- fi_availability %>%
  mutate(
    `No source row, n (%)` = paste0(
      n_no_source_rows, " (", round(100 * n_no_source_rows / n_assignments_at_cycle, 1), "%)"
    ),
    `FI observed, n (%)` = paste0(
      n_fi_observed, " (", round(100 * n_fi_observed / n_assignments_at_cycle, 1), "%)"
    ),
    `FI not observed, n (%)` = paste0(
      n_fi_not_observed, " (", round(100 * n_fi_not_observed / n_assignments_at_cycle, 1), "%)"
    ),
    `Median n FI items reported (IQR)` = paste0(
      round(median_n_answered, 1), " (", round(q1_n_answered, 1), "-", round(q3_n_answered, 1), ")"
    )
  ) %>%
  select(
    Analysis, Cohort, Group, cycle_year,
    `No source row, n (%)`,
    `FI observed, n (%)`, `FI not observed, n (%)`,
    `Median n FI items reported (IQR)`
  )

knitr::kable(fi_availability_display, digits = 2)
```

```{r active-participants-vs-usable-fi, echo=FALSE, fig.width=9, fig.height=5.5}
active_participants_vs_usable_fi <- raw_fi %>%
  mutate(
    id = as.character(id),
    cycle = as.character(cycle),
    participated = as.numeric(participated),
    fi_score_nocancer = as.numeric(fi_score_nocancer)
  ) %>%
  filter(cycle %in% analytic_cycles) %>%
  distinct(id, cycle, .keep_all = TRUE) %>%
  left_join(cycle_lookup, by = "cycle") %>%
  group_by(cycle, cycle_year) %>%
  summarize(
    n_total_active_participants = n_distinct(id[participated == 1]),
    n_valid_fi_score = n_distinct(id[participated == 1 & !is.na(fi_score_nocancer)]),
    .groups = "drop"
  ) %>%
  arrange(cycle_year)

write_csv_base(
  active_participants_vs_usable_fi,
  file.path(results_dir, "3.3_active_participants_vs_usable_fi.csv")
)

active_participants_vs_usable_fi_long <- active_participants_vs_usable_fi %>%
  tidyr::pivot_longer(
    cols = c(n_total_active_participants, n_valid_fi_score),
    names_to = "measure",
    values_to = "count"
  ) %>%
  mutate(
    measure = recode(
      measure,
      n_total_active_participants = "Total Active Participants",
      n_valid_fi_score = "Valid FI Score"
    ),
    measure = factor(
      measure,
      levels = c("Total Active Participants", "Valid FI Score")
    )
  )

active_participants_vs_usable_fi_labels <- active_participants_vs_usable_fi_long %>%
  mutate(
    label_x = cycle_year + if_else(
      measure == "Total Active Participants", -0.28, 0.28
    ),
    label_hjust = if_else(
      measure == "Total Active Participants", 1, 0
    )
  )

active_participants_plot <- ggplot(
  active_participants_vs_usable_fi_long,
  aes(x = cycle_year, y = count, fill = measure)
) +
  geom_col(position = position_dodge(width = 0.9), width = 0.42) +
  geom_text(
    data = active_participants_vs_usable_fi_labels,
    aes(
      x = label_x,
      y = count,
      label = format(count, big.mark = ",", trim = TRUE),
      hjust = label_hjust
    ),
    inherit.aes = FALSE,
    vjust = -0.25,
    size = 2.8,
    fontface = "bold"
  ) +
  scale_x_continuous(breaks = active_participants_vs_usable_fi$cycle_year) +
  scale_fill_manual(
    values = c("Total Active Participants" = "#4C78A8", "Valid FI Score" = "#E45756")
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Active Participants vs. Usable Frailty Data",
    subtitle = "Four-year analytic cycles; valid FI score means non-missing overall FI excluding cancer",
    x = "Study year",
    y = "Count",
    fill = NULL
  ) +
  theme(legend.position = "top")
active_participants_plot
```

# FI Distribution by Four-Year Analytic Cycle

```{r fi-cycle-distribution}
observed_fi_panel <- matched_assignment_panel %>%
  filter(analytic_row, !is.na(fi_score_nocancer)) %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_levels),
    Group = factor(Group, levels = group_levels)
  )

fi_distribution_by_cycle <- observed_fi_panel %>%
  group_by(Analysis, Cohort, Group, cycle, cycle_year) %>%
  summarize(
    n_observations = n(),
    n_assignments = n_distinct(assignment_id),
    n_distinct_persons = n_distinct(id),
    mean_fi = mean(fi_score_nocancer),
    sd_fi = stats::sd(fi_score_nocancer),
    median_fi = stats::median(fi_score_nocancer),
    q1_fi = stats::quantile(fi_score_nocancer, 0.25, names = FALSE),
    q3_fi = stats::quantile(fi_score_nocancer, 0.75, names = FALSE),
    .groups = "drop"
  ) %>%
  arrange(Analysis, Cohort, Group, cycle_year)

write_csv_base(fi_distribution_by_cycle,
               file.path(results_dir, "3.3_fi_distribution_by_cycle.csv"))
```

```{r fi-cycle-boxplot, echo=FALSE, fig.width=13, fig.height=12}
fi_cycle_boxplot <- observed_fi_panel %>%
  mutate(cycle_year_factor = factor(cycle_year, levels = unique(cycle_lookup$cycle_year))) %>%
  ggplot(aes(x = cycle_year_factor, y = fi_score_nocancer, fill = Group)) +
  geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.35, width = 0.75,
               position = position_dodge(width = 0.8)) +
  facet_wrap(~ Cohort, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = group_cols) +
  labs(
    title = "FI distribution by four-year analytic cycle",
    x = "Analytic cycle year",
    y = "FI excluding cancer",
    fill = NULL
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
fi_cycle_boxplot
```

# FI Distribution by Age Group

Age groups are defined using age at the analytic FI cycle, not assigned index age.

```{r fi-age-distribution}
fi_distribution_by_age_group <- observed_fi_panel %>%
  filter(!is.na(age_group)) %>%
  group_by(Analysis, Cohort, Group, age_group) %>%
  summarize(
    n_observations = n(),
    n_assignments = n_distinct(assignment_id),
    n_distinct_persons = n_distinct(id),
    mean_fi = mean(fi_score_nocancer),
    sd_fi = stats::sd(fi_score_nocancer),
    median_fi = stats::median(fi_score_nocancer),
    q1_fi = stats::quantile(fi_score_nocancer, 0.25, names = FALSE),
    q3_fi = stats::quantile(fi_score_nocancer, 0.75, names = FALSE),
    .groups = "drop"
  ) %>%
  arrange(Analysis, Cohort, Group, age_group)

write_csv_base(fi_distribution_by_age_group,
               file.path(results_dir, "3.3_fi_distribution_by_age_group.csv"))
```

```{r fi-age-boxplot, echo=FALSE, fig.width=13, fig.height=12}
fi_age_boxplot <- observed_fi_panel %>%
  filter(!is.na(age_group)) %>%
  mutate(age_group = factor(age_group, levels = age_group_levels)) %>%
  ggplot(aes(x = age_group, y = fi_score_nocancer, fill = Group)) +
  geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.35, width = 0.75,
               position = position_dodge(width = 0.8)) +
  facet_wrap(~ Cohort, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = group_cols) +
  labs(
    title = "FI distribution by age group",
    x = "Age at analytic FI cycle, years",
    y = "FI excluding cancer",
    fill = NULL
  ) +
  theme(legend.position = "bottom")
fi_age_boxplot
```

# Distinct Censoring Categories Relative to Assigned Index Age

Each matched assignment is placed in exactly one follow-up endpoint category.
Death and a control's later own cancer are ordered by their recorded month, so
the earlier event determines the category. Among assignments without either
event before the administrative endpoint, an observed 2020 FI is administrative
censoring; an earlier final FI is classified as loss to follow-up/nonresponse.
Administrative censoring is therefore not counted as dropout. These categories
describe the analytic follow-up endpoint and should not be interpreted as
causal reasons for questionnaire nonresponse.

```{r censoring-categories}
participant_death <- raw_fi %>%
  mutate(
    id = as.character(id),
    dbmy09 = as.numeric(dbmy09),
    dtdth = as.numeric(dtdth),
    age_at_death = as.numeric(age_at_death),
    death_status = as.numeric(death_status)
  ) %>%
  group_by(id) %>%
  summarize(
    birth_month = first_nonmissing(dbmy09),
    death_month = first_nonmissing(dtdth),
    age_at_death = first_nonmissing(age_at_death),
    death_status = max(death_status, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    death_status = if_else(is.infinite(death_status), NA_real_, death_status),
    death_month = if_else(
      is.na(death_month) & !is.na(age_at_death) & !is.na(birth_month),
      birth_month + 12 * age_at_death,
      death_month
    ),
    age_at_death = if_else(
      is.na(age_at_death) & !is.na(death_month) & !is.na(birth_month),
      (death_month - birth_month) / 12,
      age_at_death
    )
  )

assignment_history <- matched_assignment_panel %>%
  filter(analytic_row) %>%
  group_by(Analysis, Cohort, Group, assignment_id, id, index_date, index_age) %>%
  summarize(
    last_observed_age = if (any(!is.na(fi_score_nocancer))) {
      max(age_at_cycle[!is.na(fi_score_nocancer)], na.rm = TRUE)
    } else NA_real_,
    last_observed_month = if (any(!is.na(fi_score_nocancer) &
                                  !is.na(worked_rtmnyr))) {
      max(worked_rtmnyr[!is.na(fi_score_nocancer) &
                         !is.na(worked_rtmnyr)], na.rm = TRUE)
    } else NA_real_,
    last_observed_cycle_year = if (any(!is.na(fi_score_nocancer))) {
      max(cycle_year[!is.na(fi_score_nocancer)], na.rm = TRUE)
    } else NA_real_,
    own_cancer_month = first_nonmissing(own_cancer_month),
    .groups = "drop"
  ) %>%
  left_join(participant_death, by = "id") %>%
  mutate(
    death_relative_index_age = coalesce(
      (death_month - index_date) / 12,
      age_at_death - index_age
    ),
    own_cancer_relative_index_age = (own_cancer_month - index_date) / 12,
    last_observed_relative_index_age = last_observed_age - index_age,
    later_own_cancer = as.character(Group) == "Control" &
      !is.na(own_cancer_month) & own_cancer_month > index_date,
    death_before_administrative_end = !is.na(death_month) &
      death_month <= administrative_end_month,
    own_cancer_before_administrative_end = later_own_cancer &
      own_cancer_month <= administrative_end_month,
    death_is_first_event = death_before_administrative_end &
      (!own_cancer_before_administrative_end |
       death_month <= own_cancer_month),
    own_cancer_is_first_event = own_cancer_before_administrative_end &
      (!death_before_administrative_end |
       own_cancer_month < death_month),
    censoring_category = case_when(
      death_is_first_event ~ "Death before administrative end",
      own_cancer_is_first_event ~ "Later own-cancer censoring",
      is.na(last_observed_relative_index_age) ~ "No observed FI",
      last_observed_cycle_year >= administrative_end_year ~
        "Administrative censoring at 2020",
      TRUE ~ "Loss to follow-up/nonresponse before 2020"
    ),
    censoring_category = factor(
      censoring_category,
      levels = censoring_category_levels
    ),
    censoring_relative_index_age = case_when(
      death_is_first_event ~ death_relative_index_age,
      own_cancer_is_first_event ~ own_cancer_relative_index_age,
      TRUE ~ last_observed_relative_index_age
    )
  )

if (anyNA(assignment_history$censoring_category)) {
  stop("At least one assignment lacks a censoring category.")
}
if (any(
  assignment_history$censoring_category == "Administrative censoring at 2020" &
    assignment_history$last_observed_cycle_year != administrative_end_year
)) {
  stop("Administrative censoring was assigned without an observed 2020 FI.")
}

censoring_category_summary <- assignment_history %>%
  count(Analysis, Cohort, Group, censoring_category, name = "n_assignments") %>%
  group_by(Analysis, Cohort, Group) %>%
  mutate(
    n_total_assignments = sum(n_assignments),
    percent_assignments = 100 * n_assignments / n_total_assignments,
    administrative_end_year = administrative_end_year,
    administrative_end_month = administrative_end_month
  ) %>%
  ungroup() %>%
  arrange(Analysis, Cohort, Group, censoring_category)

category_reconciliation <- censoring_category_summary %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    n_categories_total = sum(n_assignments),
    n_expected = first(n_total_assignments),
    .groups = "drop"
  )
if (any(category_reconciliation$n_categories_total !=
        category_reconciliation$n_expected)) {
  stop("Censoring categories do not reconcile to the assignment denominator.")
}

relative_age_grid <- tibble(relative_index_age = seq(-20, 20, by = 2))

censoring_category_relative <- tidyr::crossing(
  assignment_history %>%
    select(
      Analysis, Cohort, Group, assignment_id, censoring_category,
      censoring_relative_index_age
    ),
  relative_age_grid
) %>%
  group_by(Analysis, Cohort, Group, censoring_category, relative_index_age) %>%
  summarize(
    n_category_assignments = n(),
    n_category_endpoints_by_relative_age = sum(
      !is.na(censoring_relative_index_age) &
        censoring_relative_index_age <= relative_index_age
    ),
    .groups = "drop"
  ) %>%
  left_join(
    censoring_category_summary %>%
      distinct(Analysis, Cohort, Group, n_total_assignments),
    by = c("Analysis", "Cohort", "Group")
  ) %>%
  mutate(
    percent_all_assignments_by_relative_age =
      100 * n_category_endpoints_by_relative_age / n_total_assignments
  ) %>%
  arrange(Analysis, Cohort, Group, censoring_category, relative_index_age)

death_dropout_relative <- tidyr::crossing(
  assignment_history %>%
    select(
      Analysis, Cohort, Group, assignment_id, censoring_category,
      censoring_relative_index_age
    ),
  relative_age_grid
) %>%
  group_by(Analysis, Cohort, Group, relative_index_age) %>%
  summarize(
    n_assignments = n(),
    n_death_ages_observed = sum(
      censoring_category == "Death before administrative end" &
        !is.na(censoring_relative_index_age)
    ),
    n_died_by_relative_age = sum(
      censoring_category == "Death before administrative end" &
        !is.na(censoring_relative_index_age) &
        censoring_relative_index_age <= relative_index_age
    ),
    n_dropout_ages_observed = sum(
      censoring_category == "Loss to follow-up/nonresponse before 2020" &
        !is.na(censoring_relative_index_age)
    ),
    n_dropout_by_relative_age = sum(
      censoring_category == "Loss to follow-up/nonresponse before 2020" &
        !is.na(censoring_relative_index_age) &
        censoring_relative_index_age <= relative_index_age
    ),
    percent_died_by_relative_age =
      100 * n_died_by_relative_age / n_assignments,
    percent_dropout_by_relative_age =
      100 * n_dropout_by_relative_age / n_assignments,
    .groups = "drop"
  )

attrition_event_summary <- censoring_category_summary %>%
  select(
    Analysis, Cohort, Group, censoring_category,
    n_assignments, percent_assignments
  ) %>%
  pivot_wider(
    names_from = censoring_category,
    values_from = c(n_assignments, percent_assignments),
    values_fill = 0,
    names_glue = "{.value}__{censoring_category}"
  )

write_csv_base(
  death_dropout_relative,
  file.path(results_dir, "3.3_death_dropout_relative_to_index_age.csv")
)
write_csv_base(
  attrition_event_summary,
  file.path(results_dir, "3.3_attrition_event_summary.csv")
)
write_csv_base(
  censoring_category_summary,
  file.path(results_dir, "3.3_censoring_category_summary.csv")
)
write_csv_base(
  censoring_category_relative,
  file.path(results_dir, "3.3_censoring_category_by_relative_index_age.csv")
)

censoring_category_summary_display <- censoring_category_summary %>%
  transmute(
    Analysis, Cohort, Group, `Endpoint category` = censoring_category,
    `Assignments, n (%)` = paste0(
      n_assignments, " (", round(percent_assignments, 1), "%)"
    )
  )

knitr::kable(censoring_category_summary_display, digits = 2)
```

```{r censoring-category-distribution-plot, echo=FALSE, fig.width=13, fig.height=10}
censoring_category_distribution_plot <- ggplot(
  censoring_category_summary,
  aes(x = Group, y = percent_assignments, fill = censoring_category)
) +
  geom_col(width = 0.72) +
  geom_text(
    aes(label = if_else(percent_assignments >= 4,
                        paste0(round(percent_assignments, 1), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 2.7,
    color = "white",
    fontface = "bold"
  ) +
  facet_wrap(~ Cohort, ncol = 2) +
  scale_fill_manual(values = censoring_category_cols, drop = TRUE) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title = "Mutually exclusive follow-up endpoint categories",
    subtitle = "Administrative censoring at 2020 is separated from loss to follow-up/nonresponse",
    x = NULL,
    y = "Assignments (%)",
    fill = "Endpoint category"
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 9),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
censoring_category_distribution_plot
```

```{r censoring-category-relative-time-plot, echo=FALSE, fig.width=14, fig.height=14}
censoring_category_relative_plot <- censoring_category_relative %>%
  filter(censoring_category != "No observed FI") %>%
  ggplot(
    aes(
      x = relative_index_age,
      y = percent_all_assignments_by_relative_age,
      color = Group
    )
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
  facet_grid(censoring_category ~ Cohort, scales = "free_y") +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  scale_color_manual(values = group_cols) +
  labs(
    title = "Timing of follow-up endpoints relative to assigned index age",
    subtitle = "Curves are cumulative percentages of all matched assignments within each cohort and group",
    x = "Years relative to assigned index age",
    y = "Assignments with endpoint by time (%)",
    color = NULL
  ) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7.5)
  )
censoring_category_relative_plot
```
