---
title: "Observed Frailty Trajectories in Risk-Set Matched Cancer Cohorts"
author: "Nemo Zhou"
date: "2026-09-21"
output:
  html_document:
    self_contained: true
    theme: readable
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  4.0.2_GLME_descriptive_trajectories.Rmd
Author:  Nemo Zhou
Date started: 2026-09-21
Date last updated: 2026-09-21

Purpose: Read-only visual report of empirical, binned frailty trajectories
generated by 4.0.2_GLME_descriptive_trajectories.R. No modeling, smoothing,
inference, matching, or persistent PNG files are created here.
Output: Results/cancer/visuals/4.0.2_GLME_descriptive_trajectories.html
-->

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
results_dir <- file.path(project_dir, "Results", "cancer", "data")
required <- c(
  "4.0.2_descriptive_observed_means.csv",
  "4.0.2_descriptive_group_difference.csv",
  "4.0.2_descriptive_support.csv",
  "4.0.2_descriptive_m1_eligibility.csv",
  "4.0.2_descriptive_run_metadata.rds"
)
missing <- required[!file.exists(file.path(results_dir, required))]
if (length(missing)) stop("Run 4.0.2_GLME_descriptive_trajectories.R first. Missing: ", paste(missing, collapse = ", "))

observed <- read.csv(file.path(results_dir, required[[1]]), stringsAsFactors = FALSE)
difference <- read.csv(file.path(results_dir, required[[2]]), stringsAsFactors = FALSE)
support <- read.csv(file.path(results_dir, required[[3]]), stringsAsFactors = FALSE)
eligibility <- read.csv(file.path(results_dir, required[[4]]), stringsAsFactors = FALSE)
metadata <- readRDS(file.path(results_dir, required[[5]]))
cohort_order <- c("Low/Moderate Burden Cohort", "High Burden Cohort", "Smoking-Related Cancer Cohort", "Obesity-Related Cancer Cohort", "All Cancer Cohort")
for (name in c("observed", "difference", "support", "eligibility")) {
  x <- get(name)
  x$Cohort <- factor(x$Cohort, levels = cohort_order)
  assign(name, x)
}
group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
  fig.path = file.path(tempdir(), "4.0.2_descriptive_figures", ""), dev = "png", fig.retina = 1)
```

## Purpose and interpretation

These figures summarize observed cancer-excluded frailty index (FI) values in the exact M1 analytic population, without spline fitting, random effects, covariate adjustment, confidence intervals, or hypothesis tests. The x-axis is years relative to the case diagnosis or assigned matched-control index time.

Controls can be reused across matching sets. Every valid matched assignment is retained, so the observed means are assignment-row weighted, just as the formal GLME retains those rows. Points are connected solely to help follow the empirical bin sequence; the lines are not smooth fits. These figures describe the observed data and do not provide causal or adjusted estimates.

```{r eligibility-table}
knitr::kable(eligibility, caption = "M1-eligible descriptive population by cohort")
```

## A. Observed FI by group

```{r observed-figure, fig.width=12, fig.height=13, out.width='100%'}
p_observed <- observed %>% filter(display_supported) %>%
  ggplot(aes(relative_time_mid, mean_fi, color = Group, group = Group)) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.35, color = "grey35") +
  facet_wrap(~ Cohort, ncol = 1, scales = "free_y") +
  scale_color_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(x = "Years relative to diagnosis or assigned index", y = "Observed mean no-cancer FI", color = NULL) +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom", panel.grid.minor = element_blank())
print(p_observed)
```

## B. Observed cancer-minus-control FI difference

```{r difference-figure, fig.width=12, fig.height=13, out.width='100%'}
p_difference <- difference %>% filter(display_supported) %>%
  ggplot(aes(relative_time_mid, observed_difference, group = 1)) +
  geom_hline(yintercept = 0, linewidth = 0.35, color = "grey35") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.35, color = "grey35") +
  geom_line(linewidth = 0.65, color = "#54278f") +
  geom_point(size = 1.5, color = "#54278f") +
  facet_wrap(~ Cohort, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(x = "Years relative to diagnosis or assigned index", y = "Observed FI difference: cancer minus control") +
  theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
print(p_difference)
```

## C. Observation support

```{r support-figure, fig.width=15, fig.height=15, out.width='100%'}
support_long <- support %>% filter(display_supported) %>%
  select(Cohort, rel_time_bin, relative_time_mid, n_case_participants, n_control_participants,
         n_case_assignments, n_control_assignments) %>%
  pivot_longer(-c(Cohort, rel_time_bin, relative_time_mid), names_to = "measure", values_to = "n") %>%
  mutate(
    Group = if_else(grepl("case", measure), "Cancer Case", "Control"),
    unit = if_else(grepl("participants", measure), "Unique participants", "Matched assignments")
  )
p_support <- ggplot(support_long, aes(relative_time_mid, n, color = unit, linetype = unit, group = unit)) +
  geom_line(linewidth = 0.55) + geom_point(size = 1.1) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.35, color = "grey35") +
  facet_grid(Cohort ~ Group, scales = "free_y") +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(x = "Years relative to diagnosis or assigned index", y = "Count", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom", panel.grid.minor = element_blank())
print(p_support)
```

Only bins meeting the predefined two-arm support thresholds are shown: at least `r metadata$support_thresholds[["case_participants"]]` cancer-case participants and `r metadata$support_thresholds[["control_participants"]]` control participants. A missing or unsupported internal bin breaks the display rather than allowing the plotted line to cross it.
