---
title: "M1 Spline df=3 versus df=4: Fitted Trajectories"
author: "Nemo Zhou"
date: "2026-09-23"
output:
  html_document:
    self_contained: true
    theme: readable
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  4.0.3_GLME_m1_df_trajectory_comparison.Rmd
Author:  Nemo Zhou
Date started:      2026-09-23
Date last updated: 2026-09-23

Purpose:
  Read-only visual report of the saved M1 df=3 versus df=4 ML comparison
  trajectories across the five active 4.x cohorts. Uses the paired prediction
  CSV, fit-status CSV, AIC table and run metadata produced by
  4.0.3_GLME_m1_df_aic_comparison.R. It does not fit models or regenerate
  predictions. Confidence bands are conventional model-based, pointwise 95%
  intervals from the ML fit covariance; they are not CR2 intervals or the basis
  for formal inference. The paired AIC comparison remains within cohort.

Output:
  Results/cancer/visuals/4.0.3_GLME_m1_df_trajectory_comparison.html
-->

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
results_dir <- file.path(project_dir, "Results", "cancer", "data")
runner_path <- file.path(project_dir, "Code", "2_data_analysis",
                         "4.0.3_GLME_m1_df_aic_comparison.R")
engine_path <- file.path(project_dir, "Code", "2_data_analysis",
                         "4.0_GLME_spline_functions.R")
required_files <- c(
  "4.0.3_m1_spline_df_aic_comparison.csv",
  "4.0.3_m1_spline_df_fit_status.csv",
  "4.0.3_m1_spline_df_predicted_trajectories.csv",
  "4.0.3_m1_spline_df_run_metadata.rds"
)
missing_files <- required_files[!file.exists(file.path(results_dir, required_files))]
if (length(missing_files)) {
  stop(
    "Run Code/2_data_analysis/4.0.3_GLME_m1_df_aic_comparison.R first; missing: ",
    paste(missing_files, collapse = ", "), call. = FALSE
  )
}

comparison <- read.csv(file.path(results_dir, required_files[[1]]),
                       stringsAsFactors = FALSE, check.names = FALSE)
fit_status <- read.csv(file.path(results_dir, required_files[[2]]),
                       stringsAsFactors = FALSE, check.names = FALSE)
predictions <- read.csv(file.path(results_dir, required_files[[3]]),
                        stringsAsFactors = FALSE, check.names = FALSE)
metadata <- readRDS(file.path(results_dir, required_files[[4]]))

if (!identical(unname(tools::md5sum(runner_path)), metadata$runner_md5) ||
    !identical(unname(tools::md5sum(engine_path)), metadata$shared_engine_md5)) {
  stop(
    "The saved outputs were made by different runner or engine code. Rerun ",
    "the 4.0.3 comparison before rendering this report.", call. = FALSE
  )
}
for (input_name in names(metadata$inputs)) {
  input_record <- metadata$inputs[[input_name]]
  provenance_pairs <- list(
    c(input_record$input_path, input_record$input_md5),
    c(input_record$gate_path, input_record$gate_md5),
    c(input_record$matching_run_path, input_record$matching_run_md5)
  )
  for (provenance_pair in provenance_pairs) {
    provenance_path <- provenance_pair[[1]]
    expected_md5 <- provenance_pair[[2]]
    if (is.na(provenance_path) || !file.exists(provenance_path) ||
        !identical(unname(tools::md5sum(provenance_path)), expected_md5)) {
      stop("Saved matched-data or Gate G4 provenance changed after the ",
           "comparison for ", input_name, ". Rerun 4.0.3 before rendering.",
           call. = FALSE)
    }
  }
}

cohort_order <- c(
  "Low/Moderate Burden Cohort",
  "High Burden Cohort",
  "Smoking-Related Cancer Cohort",
  "Obesity-Related Cancer Cohort",
  "All Cancer Cohort"
)
model_order <- c("M1_df3", "M1_df4")
model_labels <- c("M1_df3" = "M1 spline, df = 3 (ML)",
                  "M1_df4" = "M1 spline, df = 4 (ML)")
group_order <- c("Control", "Cancer Case")
group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")

if (!nrow(comparison) || !all(c("AIC_df3", "AIC_df4",
                                  "delta_AIC_df4_minus_df3",
                                  "row_keys_identical", "status",
                                  "prediction_status") %in% names(comparison))) {
  stop("The saved comparison table is missing required paired AIC fields.",
       call. = FALSE)
}
if (!nrow(fit_status) || !all(c("model_id", "Cohort", "status",
                                  "prediction_status", "ml_verified",
                                  "fit_row_keys_match") %in% names(fit_status))) {
  stop("The saved fit-status table is missing required validation fields.",
       call. = FALSE)
}
if (!nrow(predictions)) {
  stop("No paired fitted trajectories were saved. Inspect the 4.0.3 fit-status output.",
       call. = FALSE)
}

successful_pairs <- comparison %>%
  filter(status == "fit", prediction_status == "saved",
         row_keys_identical %in% TRUE) %>%
  pull(Cohort) %>%
  unique()
if (!length(successful_pairs)) {
  stop("No cohort has a complete, validated df=3/df=4 prediction pair.",
       call. = FALSE)
}
if (any(!fit_status$ml_verified[fit_status$Cohort %in% successful_pairs] %in% TRUE) ||
    any(!fit_status$fit_row_keys_match[fit_status$Cohort %in% successful_pairs] %in% TRUE)) {
  stop("A plotted cohort has an unverified ML fit or fitted row-key mismatch.",
       call. = FALSE)
}

comparison <- comparison %>%
  mutate(Cohort = factor(Cohort, levels = cohort_order)) %>%
  arrange(Cohort)
predictions <- predictions %>%
  filter(Cohort %in% successful_pairs, model_id %in% model_order) %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_order),
    Model = factor(model_id, levels = model_order,
                   labels = unname(model_labels[model_order])),
    Group = factor(Group, levels = group_order)
  ) %>%
  arrange(Cohort, Model, Group, Age_Centered)

expected_groups <- predictions %>%
  distinct(Cohort, model_id, Group) %>%
  count(Cohort, model_id, name = "n_groups")
model_coverage <- predictions %>%
  group_by(Cohort) %>%
  summarize(
    has_df3 = any(model_id == "M1_df3"),
    has_df4 = any(model_id == "M1_df4"),
    .groups = "drop"
  )
if (any(expected_groups$n_groups != length(group_order)) ||
    any(!model_coverage$has_df3 | !model_coverage$has_df4)) {
  stop("Saved predictions do not contain both groups and both spline degrees for every plotted cohort.",
       call. = FALSE)
}

format_aic <- function(x) formatC(x, format = "f", digits = 2, big.mark = ",")
comparison_display <- comparison %>%
  mutate(
    Cohort = as.character(Cohort),
    `AIC (df=3)` = format_aic(AIC_df3),
    `AIC (df=4)` = format_aic(AIC_df4),
    `Delta AIC (df=4 - df=3)` = formatC(
      delta_AIC_df4_minus_df3, format = "f", digits = 2, big.mark = ","
    ),
    `Observations` = formatC(n_obs, format = "d", big.mark = ","),
    `Participants` = formatC(n_participants, format = "d", big.mark = ",")
  ) %>%
  select(Cohort, `AIC (df=3)`, `AIC (df=4)`,
         `Delta AIC (df=4 - df=3)`, Observations, Participants, status,
         prediction_status)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.path = file.path(tempdir(), "4.0.3_m1_df_trajectory_figures", ""),
  dev = "png", fig.retina = 1
)
options(knitr.kable.NA = "—")
```

## Purpose and interpretation

This report displays the paired M1 natural-spline trajectories from the df=3
versus df=4 **maximum-likelihood (ML) comparison**. Each pair uses the same
M1-complete-case rows, covariates, correlated participant random intercept and
attained-age random slope. Only the relative-time spline degrees of freedom
change. Existing df=3 production models and their REML/CR2 inference are not
replaced by these comparison fits.

The curves show adjusted fixed-effect predictions at the shared M1 reference
profile for each cohort. Shaded regions are pointwise 95% intervals computed
from the conventional model-based covariance matrix of each ML fit. They are
included to display curve uncertainty, are **not CR2 intervals**, and are not
used for the AIC comparison or formal inference. Predictions are restricted
to the contiguous diagnosis-relative interval whose two-year bins meet the
shared support thresholds of at least 50 distinct cancer-case participants
and 250 distinct control participants. The dashed vertical line marks diagnosis
or the matched-control assigned index time.

## A. Within-cohort AIC comparison

Negative `AIC(df=4) - AIC(df=3)` favors df=4. AIC values are interpreted only
within their cohort; do not rank AIC across different cohorts.

```{r aic-table}
knitr::kable(
  comparison_display,
  caption = "Paired M1 ML AIC and sample sizes by cohort",
  align = "lrrrrrll"
) %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c("striped", "condensed"),
                font_size = 9)
```

## B. Fitted M1 trajectories by spline degree

Each row is a cancer cohort; the two columns compare df=3 and df=4 on the same
cohort-specific analytic rows. Case and control curves share the same color
scheme as the 4.6 summary report.

```{r trajectory-plots, fig.width=16, fig.height=14, out.width='100%'}
p_trajectory <- ggplot(
  predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4,
             color = "grey35") +
  facet_grid(Cohort ~ Model) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "M1 fitted trajectories by natural-spline degrees of freedom",
    subtitle = paste(
      "Paired ML fits; shaded regions are conventional model-based",
      "pointwise 95% intervals (not CR2)"
    ),
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted cancer-excluded frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.text.y = element_text(angle = 0)
  )
print(p_trajectory)
```

## Fit and prediction status

The figure includes only cohorts with both models accepted, their common fitted
row keys verified, and paired predictions saved. The status file records the
convergence rung, optimizer, ML check, support interval, and any prediction
failure for each candidate fit.

```{r fit-status}
fit_status %>%
  filter(Cohort %in% successful_pairs) %>%
  mutate(
    Cohort = as.character(Cohort),
    model_id = factor(model_id, levels = model_order),
    prediction_window = ifelse(
      is.finite(prediction_window_min) & is.finite(prediction_window_max),
      paste0(prediction_window_min, " to ", prediction_window_max, " years"),
      NA_character_
    )
  ) %>%
  select(Cohort, model_id, status, method, ml_verified, fit_row_keys_match,
         n_obs_fit, n_participants, convergence_rung, optimizer,
         prediction_status, prediction_window, prediction_message) %>%
  knitr::kable(caption = "Model and saved-trajectory status") %>%
  kable_styling(full_width = TRUE,
                bootstrap_options = c("striped", "condensed"),
                font_size = 8)
```

## Model specification

For both candidates, the outcome is `fi_score_nocancer`; fixed effects are
`Group × natural spline(Age_Centered)` plus index age, race, marital status,
and living arrangement. The random effects are a correlated participant
intercept and attained-age slope, `(1 + (age_at_cycle - 60) / 4 | id)`. The
estimator is ML (`REML = FALSE`). The comparison runner allows numerical
optimizer retries but does not simplify the random-effects structure.
