---
title: "Frailty Trajectories by Cancer Burden: Stochastic Drawing & Multiple Imputation"
author: "Analyst"
date: "`r Sys.Date()`"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 4.1_aggressive cancer_drawing_MI.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Combines stochastic drawing with multiple imputation for analyses of frailty trajectories by aggressive cancer burden.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary)
library(purrr)       # For map/lapply iterations
library(mice)        # For Rubin's Rules pooling
library(broom.mixed) # Allows mice to pool lmer models
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
data_path <- "/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
fi_long_merged <- readRDS(data_path)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables (Age Scale)

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) 
```

## 4. Extract Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Longitudinal Dataset

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    
    is_high_burden = if_else(
      coalesce(cancer_lung, 0) == 1 | coalesce(cancer_pancreas, 0) == 1 | 
      coalesce(cancer_stomach, 0) == 1 | coalesce(cancer_liver, 0) == 1 | 
      coalesce(cancer_brain, 0) == 1 | coalesce(cancer_lal, 0) == 1 |
      coalesce(cancer_esoph, 0) == 1 | coalesce(cancer_myeloma, 0) == 1,
      1L, 0L
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  ungroup()
```

## 6. Check Follow-Up Structure

```{r followup-checks}
zero_fu_check <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(age_at_cycle, na.rm = TRUE),
    max_fu_yrs = max(age_at_cycle, na.rm = TRUE) - min(age_at_cycle, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu_yrs == 0)

cat("Total observations in analysis_long_age =", nrow(analysis_long_age), "\n")
cat("Patients with zero follow-up years =", nrow(zero_fu_check), "\n")
```

## 7. Setup for Stochastic Drawing

```{r setup-drawing}
patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = round(first(base_age), 4),
    min_age = round(first(min_age), 4),
    max_age = round(first(max_age), 4),
    is_case = first(Ever_Cancer),
    is_high = first(is_high_burden),
    true_age_at_cancer = round(first(true_age_at_cancer), 4),
    .groups = "drop"
  ) %>%
  mutate(age_bin = floor(base_age / 5) * 5)

# Split into target populations
cases_high <- patient_summary_age %>% filter(is_case == 1, is_high == 1, true_age_at_cancer > min_age)
cases_low <- patient_summary_age %>% filter(is_case == 1, is_high == 0, true_age_at_cancer > min_age)
controls_pool <- patient_summary_age %>% filter(is_case == 0)

# Functionalized Drawing Strategy
perform_drawing <- function(cases_df, controls_df) {
  imputed_ages <- rep(NA_real_, nrow(controls_df))
  
  for (i in seq_len(nrow(controls_df))) {
    current_control <- controls_df[i, ]
    
    eligible_cases <- cases_df %>%
      filter(
        age_bin == current_control$age_bin,
        true_age_at_cancer >= current_control$min_age,
        true_age_at_cancer <= current_control$max_age
      )
    
    if (nrow(eligible_cases) > 0) {
      imputed_ages[i] <- sample(eligible_cases$true_age_at_cancer, size = 1)
    }
  }
  
  controls_assigned <- controls_df %>%
    mutate(assigned_event_age = imputed_ages) %>%
    filter(!is.na(assigned_event_age)) # Drop controls who couldn't find a donor
    
  cases_assigned <- cases_df %>%
    select(id, assigned_event_age = true_age_at_cancer)
  
  bind_rows(cases_assigned, controls_assigned) %>%
    select(id, assigned_event_age) %>%
    distinct(id, .keep_all = TRUE)
}

prep_model_data <- function(assigned_df, cohort_name, case_label) {
  analysis_long_age %>%
    inner_join(assigned_df, by = "id") %>%
    mutate(
      Cohort = cohort_name,
      Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", case_label)),
      Age_Centered = age_at_cycle - assigned_event_age,
      Post = if_else(Age_Centered >= 0, 1L, 0L),
      Age_Post = if_else(Post == 1L, Age_Centered, 0),
      index_age_z = as.numeric(scale(assigned_event_age))
    )
}
```

## 8. Generate m=5 Imputed Datasets

```{r generate-datasets}
set.seed(2026)
m_imputations <- 5

data_high_list <- list()
data_low_list <- list()

for (m in 1:m_imputations) {
  # 1. Draw independent ages for this iteration
  assigned_high_m <- perform_drawing(cases_high, controls_pool)
  assigned_low_m <- perform_drawing(cases_low, controls_pool)
  
  # 2. Prep the datasets and store them
  data_high_list[[m]] <- prep_model_data(assigned_high_m, "High Burden Cohort", "High Burden Case")
  data_low_list[[m]] <- prep_model_data(assigned_low_m, "Low/Moderate Burden Cohort", "Low/Mod Burden Case")
}

cat("Successfully generated", m_imputations, "datasets for each cohort.\n")
```

## 9. Exploratory Analysis & Sanity Checks (Using Draw m=1)

*Note: EDA is run on the first imputed dataset to avoid inflating n-counts.*

```{r descriptive-data}
# Isolate the first draw for EDA
data_combined_eda <- bind_rows(data_high_list[[1]], data_low_list[[1]]) %>%
  mutate(Analysis_Group = paste(Group, "(", Cohort, ")"))

baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- data_combined_eda %>% 
  group_by(id, Cohort) %>%
  summarize(
    Group = first(Group),
    Analysis_Group = first(Analysis_Group),
    Ever_Cancer = first(Ever_Cancer),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    assigned_event_age = first(assigned_event_age),
    num_obs = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = first(min_age),
    max_age = first(max_age),
    max_fu_yrs = first(max_followup_yrs), 
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id")
```

### Table 1: Baseline Characteristics (Draw 1)

```{r table1}
table1_age <- followup_by_patient_age %>%
  select(Cohort, Group, base_age, assigned_event_age, fi_baseline, base_race, base_marital, base_living, base_pckgr) %>%
  gtsummary::tbl_strata(
    strata = Cohort,
    ~ gtsummary::tbl_summary(
        .x,
        by = Group,
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{n} ({p}%)"),
        label = list(base_age ~ "Age at Study Entry", assigned_event_age ~ "Age at Index", fi_baseline ~ "Baseline Frailty"),
        missing_text = "(Missing)"
      ) %>% gtsummary::add_p()
  )
table1_age
```

### Density Plot of Assigned Index Ages (Draw 1)

```{r descriptive-plots, fig.width=10, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Group, color = Group)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~ Cohort) +
  scale_fill_manual(values = c("Control" = "#3182bd", "High Burden Case" = "#de2d26", "Low/Mod Burden Case" = "#fd8d3c")) +
  scale_color_manual(values = c("Control" = "#3182bd", "High Burden Case" = "#de2d26", "Low/Mod Burden Case" = "#fd8d3c")) +
  theme_minimal() +
  labs(title = "Distribution of Age at Index Event (Representative Draw m=1)", x = "Assigned Index Age (years)", y = "Density") +
  theme(legend.position = "bottom", legend.title=element_blank())
print(p_index_density_age)
```

## 10. Run Independent LME Models & Pool via Rubin's Rules

```{r lme-models-pooled}
# Define the core formula
lme_formula <- fi_score_nocancer ~ Group * Age_Centered + Group * Age_Post + 
                                   index_age_z + base_race + base_marital + base_pckgr +
                                   (1 + Age_Centered | id)

# Function to safely run models over the lists
run_models <- function(data_list) {
  lapply(data_list, function(df) {
    lmer(lme_formula, data = df, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
  })
}

# Execute on all m=5 datasets
models_high_list <- run_models(data_high_list)
models_low_list <- run_models(data_low_list)

# Pool the results using mice
pooled_high <- pool(as.mira(models_high_list))
pooled_low <- pool(as.mira(models_low_list))

cat("\n### Pooled Results: High Burden Cohort ###\n")
summary(pooled_high, conf.int = TRUE)

cat("\n### Pooled Results: Low/Moderate Burden Cohort ###\n")
summary(pooled_low, conf.int = TRUE)
```

## 11. Generate Averaged Stratified Predictions

```{r generate-predictions}
create_pooled_grid <- function(model_list, data_subset_list) {
  # Base grid assumes all m iterations share identical factor levels
  base_grid <- expand.grid(
    Age_Centered = seq(-15, 15, by = 0.1),
    Group = factor(levels(data_subset_list[[1]]$Group), levels = levels(data_subset_list[[1]]$Group))
  ) %>%
    mutate(
      Cohort = data_subset_list[[1]]$Cohort[1],
      Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
      index_age_z = 0, # Hold at sample mean
      base_race = factor(levels(data_subset_list[[1]]$base_race)[1], levels = levels(data_subset_list[[1]]$base_race)),
      base_marital = factor(levels(data_subset_list[[1]]$base_marital)[1], levels = levels(data_subset_list[[1]]$base_marital)),
      base_pckgr = factor(levels(data_subset_list[[1]]$base_pckgr)[1], levels = levels(data_subset_list[[1]]$base_pckgr))
    )
  
  # Predict across all m models
  predictions <- lapply(model_list, function(mod) {
    predict(mod, newdata = base_grid, re.form = NA)
  })
  
  # Average the predictions across the m arrays
  pred_matrix <- do.call(cbind, predictions)
  base_grid$pred <- rowMeans(pred_matrix)
  
  return(base_grid)
}

plot_grid_high_pooled <- create_pooled_grid(models_high_list, data_high_list)
plot_grid_low_pooled <- create_pooled_grid(models_low_list, data_low_list)

# Combine for faceted plotting
plot_grid_combined <- bind_rows(plot_grid_high_pooled, plot_grid_low_pooled) %>%
  mutate(
    Line_Type = if_else(grepl("Control", Group), "Drawn Control (Pooled)", as.character(Group))
  )
```

## 12. Plot the Pooled Trajectories

```{r plot-trajectories, fig.width=10, fig.height=5}
p_trajectory_faceted <- ggplot(plot_grid_combined, aes(x = Age_Centered, y = pred, color = Line_Type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  facet_wrap(~ Cohort) + 
  scale_color_manual(values = c(
    "Drawn Control (Pooled)" = "#3182bd", 
    "High Burden Case" = "#de2d26",
    "Low/Mod Burden Case" = "#fd8d3c"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Frailty Trajectories by Target Cohort",
    subtitle = paste("Pooled estimates from", m_imputations, "stochastic random draws"),
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-15, 15, by = 5), expand = expansion(mult = c(0, 0)))

print(p_trajectory_faceted)
```

\`\`\`
