---
# Project: Frailty Trajectories Before and After Incident Cancer in the Health
#          Professionals Follow-up Study
# Script:  4.5_GLME_overall_report.Rmd
# Author:  Nemo Zhou
# Date started:      2026-07-18
# Date last updated: 2026-07-20 (removed quadratic model; renumbered spline models)
# Purpose: Read the saved overall-cancer matching and GLME CSV/RDS artifacts,
#          display primary estimates and diagnostics, and render a self-contained
#          HTML report only under Results/cancer/visuals. This report never fits
#          a model, rebuilds matching, or persists PNG files.
title: "Overall Incident Cancer Risk-Set Matching and GLME"
author: "Nemo Zhou"
date: "2026-07-18"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

```{r setup, include=FALSE}
project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
results_dir <- file.path(project_dir, "Results", "cancer", "data")
matching_dir <- file.path(results_dir, "matching_diagnostics")
sensitivity_dir <- file.path(results_dir, "4.5_sensitivities")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = file.path(tempdir(), "4.5_overall_report_figures", ""),
  dev = "png",
  dpi = 120
)

read_csv_if_present <- function(name, directory = results_dir) {
  path <- file.path(directory, name)
  if (!file.exists(path)) return(NULL)
  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

show_table <- function(x, caption, digits = 4) {
  if (is.null(x) || nrow(x) == 0) {
    cat("*Artifact unavailable or empty at render time.*\n")
  } else {
    print(knitr::kable(x, caption = caption, digits = digits))
  }
}

primary_files <- list(
  trajectories = read_csv_if_present("4.5_spline_predicted_trajectories.csv"),
  difference = read_csv_if_present("4.5_spline_group_difference.csv"),
  derivative = read_csv_if_present("4.5_spline_derivative.csv"),
  theta = read_csv_if_present("4.5_spline_theta.csv"),
  status = read_csv_if_present("4.5_spline_model_status.csv"),
  omnibus = read_csv_if_present("4.5_spline_omnibus_tests.csv"),
  support = read_csv_if_present("4.5_support_by_time_bin.csv"),
  filter_log = read_csv_if_present("4.5_model_filter_log.csv"),
  convergence = read_csv_if_present("4.5_convergence_attempts.csv"),
  variance = read_csv_if_present("4.5_spline_variance_components.csv"),
  residual = read_csv_if_present("4.5_residual_diagnostics.csv"),
  random_effect = read_csv_if_present("4.5_random_effect_diagnostics.csv"),
  bounded = read_csv_if_present("4.5_spline_bounded_prediction_checks.csv"),
  influence = read_csv_if_present("4.5_spline_influence_refits.csv")
)
sensitivity_files <- list(
  status = read_csv_if_present("4.5_sensitivity_status.csv", sensitivity_dir),
  theta = read_csv_if_present("4.5_sensitivity_theta.csv", sensitivity_dir),
  convergence = read_csv_if_present("4.5_sensitivity_convergence_attempts.csv", sensitivity_dir),
  support = read_csv_if_present("4.5_sensitivity_support_by_time_bin.csv", sensitivity_dir),
  s12_sets = read_csv_if_present("4.5_S12_matched_set_support.csv", sensitivity_dir)
)
```

## Run provenance and readiness

This report is a view of saved artifacts. It does not run matching or fit a model. The overall GLME is interpretable only after Gate G4 passes for the corrected overall matched cohort. Previously generated overall IPCW, FDA, and joint-model products are stale after that cohort is rebuilt and are not used here.

```{r artifact-inventory}
artifact_paths <- c(
  list.files(results_dir, pattern = "^4\\.5_", full.names = TRUE),
  if (dir.exists(sensitivity_dir)) list.files(sensitivity_dir, full.names = TRUE) else character(0)
)
artifact_inventory <- if (length(artifact_paths) == 0) {
  data.frame(artifact = character(0), modified = character(0), bytes = numeric(0))
} else {
  info <- file.info(artifact_paths)
  data.frame(
    artifact = basename(artifact_paths),
    modified = format(info$mtime, tz = "America/New_York"),
    bytes = info$size,
    stringsAsFactors = FALSE
  )
}
show_table(artifact_inventory, "Saved overall matching/GLME artifacts available to this report", digits = 0)
```

```{r matching-integrity}
integrity_paths <- if (dir.exists(matching_dir)) {
  list.files(matching_dir, pattern = "overall.*(set_integrity|caliper_integrity|balance|inactive_cases|scaling).*\\.(csv|rds)$", full.names = TRUE, ignore.case = TRUE)
} else character(0)
integrity_inventory <- data.frame(
  artifact = basename(integrity_paths),
  modified = if (length(integrity_paths)) format(file.info(integrity_paths)$mtime, tz = "America/New_York") else character(0),
  stringsAsFactors = FALSE
)
show_table(integrity_inventory, "Overall matching integrity and scaling artifacts", digits = 0)
```

## Model status and primary estimands

```{r status-table}
show_table(primary_files$status, "Spline model convergence and covariance status")
```

```{r theta-table}
show_table(primary_files$theta, "Post-minus-pre cancer-control slope contrast and pre-index slope")
```

```{r omnibus-table}
show_table(primary_files$omnibus, "Time-structure interaction omnibus tests")
```

The numerical and closed-form versions of `theta` must agree within `1e-6`. A model status row using CR0 or model-based covariance is retained but must be interpreted with its reported covariance label. In particular, an acceptably fitted M3 remains informative when robust covariance is unavailable.

## Adjusted trajectories

```{r trajectory-plot, fig.width=8, fig.height=5.5}
tr <- primary_files$trajectories
if (is.null(tr) || !all(c("Age_Centered", "Group", "pred") %in% names(tr))) {
  cat("*Adjusted trajectory artifact is unavailable or does not contain the expected columns.*\n")
} else {
  cohort_col <- if ("Cohort" %in% names(tr)) "Cohort" else NULL
  model_col <- if ("model_id" %in% names(tr)) "model_id" else if ("Model" %in% names(tr)) "Model" else NULL
  if (!is.null(model_col) && any(tr[[model_col]] == "M1_primary_spline", na.rm = TRUE)) {
    tr <- tr[tr[[model_col]] == "M1_primary_spline", ]
  }
  p <- ggplot2::ggplot(tr, ggplot2::aes(x = Age_Centered, y = pred, color = Group, fill = Group)) +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, color = "grey45") +
    ggplot2::geom_line(linewidth = 0.8)
  if (all(c("lwr", "upr") %in% names(tr))) {
    p <- p + ggplot2::geom_ribbon(ggplot2::aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA)
  }
  p +
    ggplot2::coord_cartesian(xlim = c(-20, 20)) +
    ggplot2::labs(x = "Years relative to cancer diagnosis / assigned index", y = "Adjusted frailty index", color = NULL, fill = NULL) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
}
```

## Cancer-minus-control difference and derivative

```{r difference-plot, fig.width=8, fig.height=5}
df <- primary_files$difference
time_name <- if (!is.null(df) && "Age_Centered" %in% names(df)) "Age_Centered" else if (!is.null(df) && "time" %in% names(df)) "time" else NULL
estimate_name <- if (!is.null(df) && "estimate" %in% names(df)) "estimate" else if (!is.null(df) && "diff" %in% names(df)) "diff" else NULL
if (is.null(time_name) || is.null(estimate_name)) {
  cat("*Cancer-minus-control difference artifact is unavailable or does not contain the expected columns.*\n")
} else {
  p <- ggplot2::ggplot(df, ggplot2::aes(x = .data[[time_name]], y = .data[[estimate_name]])) +
    ggplot2::geom_hline(yintercept = 0, color = "grey55") +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, color = "grey45") +
    ggplot2::geom_line(linewidth = 0.8, color = "#8b1a1a")
  if (all(c("lwr", "upr") %in% names(df))) {
    p <- p + ggplot2::geom_ribbon(ggplot2::aes(ymin = lwr, ymax = upr), fill = "#8b1a1a", alpha = 0.16)
  }
  p + ggplot2::coord_cartesian(xlim = c(-20, 20)) +
    ggplot2::labs(x = "Years relative to index", y = "Cancer case minus control FI") +
    ggplot2::theme_minimal(base_size = 12)
}
```

```{r derivative-plot, fig.width=8, fig.height=5}
dv <- primary_files$derivative
time_name <- if (!is.null(dv) && "Age_Centered" %in% names(dv)) "Age_Centered" else if (!is.null(dv) && "time" %in% names(dv)) "time" else NULL
estimate_name <- if (!is.null(dv) && "estimate" %in% names(dv)) "estimate" else if (!is.null(dv) && "derivative" %in% names(dv)) "derivative" else NULL
if (is.null(time_name) || is.null(estimate_name)) {
  cat("*Derivative artifact is unavailable or does not contain the expected columns.*\n")
} else {
  p <- ggplot2::ggplot(dv, ggplot2::aes(x = .data[[time_name]], y = .data[[estimate_name]])) +
    ggplot2::geom_hline(yintercept = 0, color = "grey55") +
    ggplot2::geom_vline(xintercept = 0, linetype = 2, color = "grey45") +
    ggplot2::geom_line(linewidth = 0.8, color = "#1f5a91")
  if (all(c("lwr", "upr") %in% names(dv))) {
    p <- p + ggplot2::geom_ribbon(ggplot2::aes(ymin = lwr, ymax = upr), fill = "#1f5a91", alpha = 0.16)
  }
  p + ggplot2::coord_cartesian(xlim = c(-20, 20)) +
    ggplot2::labs(x = "Years relative to index", y = "Derivative of cancer-control difference") +
    ggplot2::theme_minimal(base_size = 12)
}
```

## Diagnostics and sensitivities

```{r diagnostic-tables}
show_table(primary_files$filter_log, "Complete-case filtering by arm and model")
show_table(primary_files$convergence, "Strict convergence-ladder attempts")
show_table(primary_files$variance, "Variance components")
show_table(primary_files$residual, "Residual diagnostics")
show_table(primary_files$random_effect, "Random-effect diagnostics")
show_table(primary_files$bounded, "Bounded-outcome prediction checks")
show_table(primary_files$influence, "Deterministic matched-set-decile influence refits")
```

```{r sensitivity-results}
show_table(sensitivity_files$status, "S1-S12 model status")
show_table(sensitivity_files$theta, "S1-S12 theta estimates")
show_table(sensitivity_files$convergence, "S1-S12 strict convergence attempts")
show_table(sensitivity_files$s12_sets, "S12 retained matched-set integrity")
```

```{r sensitivity-inventory}
saved_sensitivity_paths <- if (dir.exists(sensitivity_dir)) list.files(sensitivity_dir, full.names = TRUE) else character(0)
sensitivity_inventory <- data.frame(
  artifact = basename(saved_sensitivity_paths),
  modified = if (length(saved_sensitivity_paths)) format(file.info(saved_sensitivity_paths)$mtime, tz = "America/New_York") else character(0),
  stringsAsFactors = FALSE
)
show_table(sensitivity_inventory, "S1–S12 saved sensitivity artifacts", digits = 0)
```

## Interpretation constraints

- A flat near-zero pre-index difference derivative supports comparable pre-diagnosis trajectories; a widening pre-index derivative limits a simple acceleration interpretation.
- `theta` is conditional on survival and observation under the fitted mixed-model assumptions.
- Predictions and contrasts are shown only over supported ±20 years even though all eligible relative-time observations enter model fitting.
- Sparse-tail behavior is not interpreted beyond adequate case and control support.
- IPCW, FDA, and joint-model findings must be regenerated from the corrected overall matched cohort before cross-method comparison.
