---
title: "Frailty Trajectories: Smoking-Related vs. Non-Smoking Cancers"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 4.5_Smoking Cancers.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Compares frailty trajectories for smoking-related versus non-smoking-related cancers using longitudinal frailty data and mixed-effects models.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary)
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
# Update path if necessary
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables (Age Scale)

*Note: Converting all indicator variables to numeric to prevent coalesce() errors.*

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12,
    
    # Ensure all cancer flags are numeric
    cancer_lung       = as.numeric(as.character(cancer_lung)),
    cancer_bladder    = as.numeric(as.character(cancer_bladder)),
    cancer_kidney     = as.numeric(as.character(cancer_kidney)),
    cancer_oralc      = as.numeric(as.character(cancer_oralc)),
    cancer_pharyn     = as.numeric(as.character(cancer_pharyn)),
    cancer_esoph      = as.numeric(as.character(cancer_esoph)),
    cancer_cancerdth  = as.numeric(as.character(cancer_cancerdth))
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) 
```

## 4. Extract Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Dataset (Smoking-Related Split)

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    
    # Identify Smoking-Associated Cancers
    is_smoking_cancer = if_else(
      coalesce(cancer_lung, 0) == 1 | coalesce(cancer_bladder, 0) == 1 |
      coalesce(cancer_kidney, 0) == 1 | coalesce(cancer_oralc, 0) == 1 |
      coalesce(cancer_pharyn, 0) == 1 | coalesce(cancer_esoph, 0) == 1,
      1L, 0L
    ),
    
    # Identify Non-Smoking Cancers
    is_nonsmoking_cancer = if_else(Ever_Cancer == 1 & is_smoking_cancer == 0, 1L, 0L)
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  ungroup()
```

## 6. Check Follow-Up Structure

```{r followup-checks}
cat("Total observations =", nrow(analysis_long_age), "\n")
```

## 7. Setup for Stochastic Drawing

```{r setup-drawing}
patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = round(first(base_age), 4),
    min_age = round(first(min_age), 4),
    max_age = round(first(max_age), 4),
    is_case = first(Ever_Cancer),
    is_smoking = first(is_smoking_cancer),
    true_age_at_cancer = round(first(true_age_at_cancer), 4),
    .groups = "drop"
  ) %>%
  mutate(age_bin = floor(base_age / 5) * 5)

# Split into target populations
cases_smoking <- patient_summary_age %>% filter(is_case == 1, is_smoking == 1, true_age_at_cancer > min_age)
cases_nonsmoking <- patient_summary_age %>% filter(is_case == 1, is_smoking == 0, true_age_at_cancer > min_age)
controls_pool <- patient_summary_age %>% filter(is_case == 0)

# Functionalized Drawing Strategy
perform_drawing <- function(cases_df, controls_df) {
  imputed_ages <- rep(NA_real_, nrow(controls_df))
  for (i in seq_len(nrow(controls_df))) {
    current_control <- controls_df[i, ]
    eligible_cases <- cases_df %>%
      filter(
        age_bin == current_control$age_bin,
        true_age_at_cancer >= current_control$min_age,
        true_age_at_cancer <= current_control$max_age
      )
    if (nrow(eligible_cases) > 0) {
      imputed_ages[i] <- sample(eligible_cases$true_age_at_cancer, size = 1)
    }
  }
  controls_assigned <- controls_df %>%
    mutate(assigned_event_age = imputed_ages) %>%
    filter(!is.na(assigned_event_age)) 
    
  cases_assigned <- cases_df %>%
    select(id, assigned_event_age = true_age_at_cancer)
  
  bind_rows(cases_assigned, controls_assigned) %>%
    select(id, assigned_event_age) %>%
    distinct(id, .keep_all = TRUE)
}

prep_model_data <- function(assigned_df, cohort_name, case_label) {
  analysis_long_age %>%
    inner_join(assigned_df, by = "id") %>%
    mutate(
      Cohort = cohort_name,
      Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", case_label)),
      Age_Centered = age_at_cycle - assigned_event_age,
      Post = if_else(Age_Centered >= 0, 1L, 0L),
      Age_Post = if_else(Post == 1L, Age_Centered, 0),
      index_age_z = as.numeric(scale(assigned_event_age))
    )
}
```

## 8. Generate Analysis Datasets (Single Draw)

```{r generate-datasets}
set.seed(2026) 

# Perform the drawing once for each cohort
assigned_smoke <- perform_drawing(cases_smoking, controls_pool)
assigned_nonsmoke <- perform_drawing(cases_nonsmoking, controls_pool)

# Prepare model datasets
data_smoke <- prep_model_data(assigned_smoke, "Smoking-Related Cancers", "Smoking-Related Case")
data_nonsmoke <- prep_model_data(assigned_nonsmoke, "Non-Smoking Cancers", "Non-Smoking Case")

# Combine for exploratory data analysis
data_combined <- bind_rows(data_smoke, data_nonsmoke) %>%
  mutate(Analysis_Group = paste(Group, "(", Cohort, ")"))

cat("Smoking-Related Cohort - Total Subjects:", n_distinct(data_smoke$id), "\n")
cat("Non-Smoking Cohort - Total Subjects:", n_distinct(data_nonsmoke$id), "\n")
```

## 9. Exploratory Analysis & Sanity Checks

```{r descriptive-data}
baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- data_combined %>% 
  group_by(id, Cohort) %>%
  summarize(
    Group = first(Group),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    assigned_event_age = first(assigned_event_age),
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id")
```

### Density Plot of Assigned Index Ages

```{r descriptive-plots, fig.width=10, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Group, color = Group)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~ Cohort) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Smoking-Related Case" = "#de2d26", "Non-Smoking Case" = "#fd8d3c")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Smoking-Related Case" = "#de2d26", "Non-Smoking Case" = "#fd8d3c")) +
  theme_minimal() +
  labs(title = "Distribution of Age at Index Event", x = "Assigned Index Age (years)", y = "Density") +
  theme(legend.position = "bottom", legend.title=element_blank())
print(p_index_density_age)
```
```{r}
# Bin the baseline ages for conditional visualization
plot_data_binned <- followup_by_patient_age %>%
  mutate(
    age_bin_factor = cut(base_age, 
                         breaks = c(40, 50, 60, 70, 80, 100), 
                         labels = c("40-49", "50-59", "60-69", "70-79", "80+"),
                         right = FALSE)
  ) %>%
  group_by(Cohort, age_bin_factor, Group) %>%
  filter(n() > 10) %>% # Filter out sparse bins for plotting clarity
  ungroup()

p_index_density_stratified <- ggplot(plot_data_binned, aes(x = assigned_event_age, fill = Group, color = Group)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_grid(Cohort ~ age_bin_factor, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "#3182bd", "Smoking-Related Case" = "#de2d26", "Non-Smoking Case" = "#fd8d3c")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Smoking-Related Case" = "#de2d26", "Non-Smoking Case" = "#fd8d3c")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Distribution of Index Age Conditioned on Baseline Age", 
    subtitle = "Within each baseline age stratum, the control index ages perfectly mirror the case index ages",
    x = "Assigned Index Age (years)", 
    y = "Density"
  ) +
  theme(
    legend.position = "bottom", 
    legend.title=element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f4f4f4", color = NA),
    strip.text = element_text(face = "bold")
  )

print(p_index_density_stratified)
```
```{r boxplot-index-stratified, fig.width=12, fig.height=8}
# 1. Create the binned dataset (skip if already created for the density plot)
plot_data_binned <- followup_by_patient_age %>%
  mutate(
    age_bin_factor = cut(base_age, 
                         breaks = c(40, 50, 60, 70, 80, 100), 
                         labels = c("40-49", "50-59", "60-69", "70-79", "80+"),
                         right = FALSE)
  ) %>%
  group_by(Cohort, age_bin_factor, Group) %>%
  filter(n() > 10) %>% # Filter out sparse bins for plotting clarity
  ungroup()

# 2. Generate the stratified box plot with median labels
p_boxplot_stratified <- ggplot(plot_data_binned, aes(x = Group, y = assigned_event_age, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  
  # --- NEW: Automatically calculate and plot the median values ---
  stat_summary(
    fun = median, 
    geom = "label", # Using a label instead of text gives it a clean white background
    aes(label = round(after_stat(y), 1)), 
    vjust = -0.3,   # Push the label slightly above the median line
    size = 3.5,     # Font size
    fill = "white", # White background so the line doesn't strike through the number
    label.size = NA, # Remove the border around the label
    alpha = 0.8     # Slight transparency
  ) +
  
  # Create a matrix: Cohort on the rows, Age Bin on the columns
  facet_grid(Cohort ~ age_bin_factor) + 
  
  # Colors from your established palette
  scale_fill_manual(values = c(
    "Control" = "#3182bd", 
    "Smoking-Related Case" = "#de2d26", 
    "Non-Smoking Case" = "#fd8d3c"
  )) +
  
  theme_minimal(base_size = 12) +
  labs(
    title = "Box Plot of Index Age Conditioned on Baseline Age",
    subtitle = "Comparing the median and spread of assigned index ages within each demographic stratum",
    x = NULL, 
    y = "Assigned Index Age (Years)"
  ) +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(face = "bold", angle = 45, hjust = 1), 
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(), 
    strip.background = element_rect(fill = "#f4f4f4", color = NA),
    strip.text = element_text(face = "bold")
  )

print(p_boxplot_stratified)
```

## 10. Run Independent LME Models

```{r lme-models}
lme_formula <- fi_score_nocancer ~ Group * Age_Centered + Group * Age_Post + 
                                   index_age_z + base_race + base_marital + base_pckgr +
                                   (1 + Age_Centered | id)

model_smoke <- lmer(lme_formula, data = data_smoke, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
cat("\n### Model Results: Smoking-Related Cancers ###\n")
summary(model_smoke)

model_nonsmoke <- lmer(lme_formula, data = data_nonsmoke, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
cat("\n### Model Results: Non-Smoking Cancers ###\n")
summary(model_nonsmoke)
```

## 11. Generate Stratified Predictions

```{r generate-predictions}
create_pred_grid <- function(model, data_subset) {
  base_grid <- expand.grid(
    Age_Centered = seq(-15, 15, by = 0.1),
    Group = factor(levels(data_subset$Group), levels = levels(data_subset$Group))
  ) %>%
    mutate(
      Cohort = data_subset$Cohort[1],
      Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
      index_age_z = 0, 
      base_race = factor(levels(data_subset$base_race)[1], levels = levels(data_subset$base_race)),
      base_marital = factor(levels(data_subset$base_marital)[1], levels = levels(data_subset$base_marital)),
      base_pckgr = factor(levels(data_subset$base_pckgr)[1], levels = levels(data_subset$base_pckgr))
    )
  
  base_grid$pred <- predict(model, newdata = base_grid, re.form = NA)
  return(base_grid)
}

plot_grid_smoke <- create_pred_grid(model_smoke, data_smoke)
plot_grid_nonsmoke <- create_pred_grid(model_nonsmoke, data_nonsmoke)

plot_grid_combined <- bind_rows(plot_grid_smoke, plot_grid_nonsmoke) %>%
  mutate(Line_Type = if_else(grepl("Control", Group), "Drawn Control", as.character(Group)))
```

## 12. Plot the Trajectories

```{r plot-trajectories, fig.width=10, fig.height=5}
p_trajectory_faceted <- ggplot(plot_grid_combined, aes(x = Age_Centered, y = pred, color = Line_Type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  facet_wrap(~ Cohort) + 
  scale_color_manual(values = c(
    "Drawn Control" = "#3182bd", 
    "Smoking-Related Case" = "#de2d26",
    "Non-Smoking Case" = "#fd8d3c"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Frailty Trajectories: Smoking-Related vs. Non-Smoking Cancers",
    subtitle = "Assessing pre-clinical frailty differences based on tumor etiology",
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-15, 15, by = 5), expand = expansion(mult = c(0, 0)))

print(p_trajectory_faceted)
```
