---
title: "Risk-Set Matched Cancer Cohort Summary and GLME Results"
author: "Nemo Zhou"
date: "2026-07-21"
output:
  html_document:
    self_contained: true
    theme: readable
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  4.6_GLME_summary_report.Rmd
Author:  Nemo Zhou
Date started:      2026-07-21
Date last updated: 2026-07-27 (temporarily hid Joint Wald tests from rendered HTML while retaining the source blocks)

Purpose:
  Read-only summary report for the active non-survival risk-set matched cancer
  cohorts. The report displays:
    1. An assignment-level Table One of model and descriptive covariates by
       cancer-case and cancer-control status, in paired cohort columns.
    2. Saved event-study and natural-spline joint Wald tests retained in the
       source but temporarily hidden from rendered HTML.
    3. Trajectory plots with one row per cancer cohort and four model columns:
       M0, M1, M2, and M3.
    4. A separate M2-only trajectory plot with one cancer cohort per row.

  The report reads saved RDS/CSV artifacts only. It does not rebuild matching,
  fit GLME models, run survival-stratified analyses, or save persistent PNG
  files. Reused controls are retained as separate matched assignments, while
  repeated FI observations are collapsed to one row per assignment for Table
  One so the table is not inflated by the longitudinal measurement count.

Inputs:
  Data/riskset_matched_analysis_long.rds
  Data/riskset_matched_smoking_long.rds
  Data/riskset_matched_obesity_long.rds
  Data/riskset_matched_overall_long.rds
  Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds
  Results/cancer/data/4.1_*, 4.2_*, 4.3_*, and 4.5_* saved artifacts

Output:
  Results/cancer/visuals/4.6_GLME_summary_report.html
-->

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyr)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
data_dir <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "Results", "cancer", "data")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = file.path(tempdir(), "4.6_GLME_summary_report_figures", ""),
  dev = "png",
  fig.retina = 1
)
options(knitr.kable.NA = "—")

cohort_inputs <- tibble::tribble(
  ~prefix, ~matched_file,
  "4.1", "riskset_matched_analysis_long.rds",
  "4.2", "riskset_matched_smoking_long.rds",
  "4.3", "riskset_matched_obesity_long.rds",
  "4.5", "riskset_matched_overall_long.rds"
)

cohort_order <- c(
  "Low/Moderate Burden Cohort",
  "High Burden Cohort",
  "Smoking-Related Cancer Cohort",
  "Obesity-Related Cancer Cohort",
  "All Cancer Cohort"
)

cohort_labels <- c(
  "Low/Moderate Burden Cohort" = "Low/moderate burden cancer status",
  "High Burden Cohort" = "High burden cancer status",
  "Smoking-Related Cancer Cohort" = "Smoking-related cancer status",
  "Obesity-Related Cancer Cohort" = "Obesity-related cancer status",
  "All Cancer Cohort" = "All cancer status"
)

group_order <- c("Cancer Case", "Control")
group_labels <- c("Cancer Case" = "Cancer case", "Control" = "Control")

required_matched_columns <- c(
  "id", "match_set", "role", "Cohort", "Group", "index_age",
  "Age_Centered", "fi_score_nocancer", "base_race", "base_marital",
  "base_living", "base_pckgr", "base_calor", "base_sat",
  "base_diet_chol", "base_alco", "post_own_cancer"
)

as_numeric_safely <- function(x) {
  suppressWarnings(as.numeric(as.character(x)))
}

first_nonmissing <- function(x) {
  keep <- !is.na(x)
  if (!any(keep)) return(x[NA_integer_][1])
  x[which(keep)[1]]
}

last_nonmissing <- function(x) {
  keep <- !is.na(x)
  if (!any(keep)) return(x[NA_integer_][1])
  x[tail(which(keep), 1)]
}

mean_sd <- function(x, digits = 2) {
  x <- as_numeric_safely(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_character_)
  paste0(formatC(mean(x), format = "f", digits = digits),
         " (", formatC(stats::sd(x), format = "f", digits = digits), ")")
}

format_p <- function(x) {
  if (is.na(x)) return(NA_character_)
  if (x < 0.001) return("&lt;0.001")
  formatC(x, format = "f", digits = 3)
}

read_matched <- function(prefix, file_name) {
  path <- file.path(data_dir, file_name)
  if (!file.exists(path)) stop("Matched dataset not found: ", path, call. = FALSE)
  dat <- readRDS(path)
  missing_columns <- setdiff(required_matched_columns, names(dat))
  if (length(missing_columns)) {
    stop("Matched dataset is missing required columns: ",
         paste(missing_columns, collapse = ", "), call. = FALSE)
  }
  dat %>%
    mutate(
      Analysis_Prefix = prefix,
      id = as.character(id),
      match_set = as.character(match_set),
      role = as.character(role),
      Cohort = as.character(Cohort),
      Group = factor(as.character(Group), levels = group_order),
      index_age = as_numeric_safely(index_age),
      Age_Centered = as_numeric_safely(Age_Centered),
      fi_score_nocancer = as_numeric_safely(fi_score_nocancer),
      post_own_cancer = as.logical(post_own_cancer)
    ) %>%
    filter(!post_own_cancer)
}

matched_long <- bind_rows(lapply(seq_len(nrow(cohort_inputs)), function(i) {
  read_matched(cohort_inputs$prefix[[i]], cohort_inputs$matched_file[[i]])
}))

raw_path <- file.path(data_dir, "FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")
baseline_extra <- NULL
if (file.exists(raw_path)) {
  raw <- readRDS(raw_path)
  raw_names <- names(raw)
  baseline_extra <- raw %>%
    mutate(id = as.character(id), cycle = as.character(cycle)) %>%
    filter(cycle == "86") %>%
    distinct(id, .keep_all = TRUE) %>%
    transmute(
      id,
      base_bmi = if ("bmi" %in% raw_names) as_numeric_safely(bmi) else NA_real_,
      base_activity = if ("act" %in% raw_names) as_numeric_safely(act) else NA_real_
    )
}

assignment_summary <- matched_long %>%
  arrange(Analysis_Prefix, Cohort, match_set, id, role, Age_Centered) %>%
  group_by(Analysis_Prefix, Cohort, match_set, id, role, Group) %>%
  summarize(
    index_age = first_nonmissing(index_age),
    fi_first_observed = first_nonmissing(fi_score_nocancer),
    fi_last_pre_index = last_nonmissing(fi_score_nocancer[Age_Centered < 0]),
    n_fi_observations = sum(!is.na(fi_score_nocancer)),
    n_pre_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered < 0),
    n_post_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered >= 0),
    base_race = first_nonmissing(base_race),
    base_marital = first_nonmissing(base_marital),
    base_living = first_nonmissing(base_living),
    base_pckgr = first_nonmissing(base_pckgr),
    base_calor = first_nonmissing(base_calor),
    base_sat = first_nonmissing(base_sat),
    base_diet_chol = first_nonmissing(base_diet_chol),
    base_alco = first_nonmissing(base_alco),
    .groups = "drop"
  )

if (!is.null(baseline_extra)) {
  assignment_summary <- assignment_summary %>% left_join(baseline_extra, by = "id")
}

assignment_summary <- assignment_summary %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_order),
    Group = factor(Group, levels = group_order)
  )

continuous_vars <- c(
  index_age = "Age at diagnosis/index, years",
  fi_first_observed = "Frailty index, first observed analytic cycle",
  fi_last_pre_index = "Frailty index, last pre-index observation",
  base_calor = "Calories",
  base_sat = "Saturated fat",
  base_diet_chol = "Dietary cholesterol",
  base_alco = "Alcohol"
)

optional_continuous_vars <- c(
  base_bmi = "BMI",
  base_activity = "Physical activity"
)
if (any(names(optional_continuous_vars) %in% names(assignment_summary))) {
  continuous_vars <- c(
    continuous_vars,
    optional_continuous_vars[names(optional_continuous_vars) %in% names(assignment_summary)]
  )
}

categorical_vars <- c(
  base_race = "Race",
  base_marital = "Marital status",
  base_living = "Living arrangement",
  base_pckgr = "Pack-year group"
)

continuous_vars <- continuous_vars[vapply(
  names(continuous_vars),
  function(v) any(!is.na(assignment_summary[[v]])),
  logical(1)
)]
categorical_vars <- categorical_vars[vapply(
  names(categorical_vars),
  function(v) any(!is.na(assignment_summary[[v]])),
  logical(1)
)]

summarize_continuous <- function(dat, var, label) {
  dat %>%
    group_by(Group) %>%
    summarize(Statistic = mean_sd(.data[[var]]), .groups = "drop") %>%
    mutate(Variable = label, Level = NA_character_, Level_order = 0L)
}

summarize_categorical <- function(dat, var, label) {
  pack_year_levels <- c(
    "<10 pack-years", "10-24 pack-years", "25-44 pack-years",
    "45-64 pack-years", "65+ pack-years", "Missing", "Never"
  )

  level_summary <- dat %>%
    filter(!is.na(.data[[var]])) %>%
    count(Group, Level = as.character(.data[[var]]), name = "n") %>%
    mutate(
      Level_order = if (var == "base_pckgr") {
        match(Level, pack_year_levels)
      } else {
        match(Level, sort(unique(Level)))
      }
    ) %>%
    group_by(Group) %>%
    mutate(Statistic = paste0(formatC(100 * n / sum(n), format = "f", digits = 1), "%")) %>%
    ungroup() %>%
    transmute(
      Group, Statistic, Variable = label,
      Level = htmltools::htmlEscape(Level, attribute = FALSE), Level_order
    )

  variable_row <- tibble::tibble(
    Group = factor(group_order, levels = group_order),
    Statistic = "",
    Variable = label,
    Level = NA_character_,
    Level_order = 0L
  )

  bind_rows(variable_row, level_summary)
}

table1_long <- bind_rows(lapply(seq_len(nrow(cohort_inputs)), function(i) {
  prefix_data <- assignment_summary %>% filter(Analysis_Prefix == cohort_inputs$prefix[[i]])
  bind_rows(lapply(unique(as.character(prefix_data$Cohort)), function(cohort_name) {
    cohort_data <- prefix_data %>% filter(as.character(Cohort) == cohort_name)
    bind_rows(
      lapply(names(continuous_vars), function(v) summarize_continuous(cohort_data, v, continuous_vars[[v]])),
      lapply(names(categorical_vars), function(v) summarize_categorical(cohort_data, v, categorical_vars[[v]]))
    ) %>% mutate(Cohort = cohort_name)
  }))
}))

table1_row_order <- c(unname(continuous_vars), unname(categorical_vars))
table1_long <- table1_long %>%
  mutate(
    Variable = factor(Variable, levels = table1_row_order),
    Group = factor(Group, levels = group_order),
    Cohort = factor(Cohort, levels = cohort_order)
  ) %>%
  mutate(
    variable_order = match(as.character(Variable), table1_row_order)
  )

table1_wide <- table1_long %>%
  mutate(
    Characteristic = if_else(is.na(Level), as.character(Variable), paste0("&nbsp;&nbsp;", as.character(Level))),
    Characteristic_order = variable_order * 1000 + Level_order
  ) %>%
  select(Cohort, Group, Characteristic, Characteristic_order, Statistic) %>%
  tidyr::pivot_wider(
    id_cols = c(Characteristic, Characteristic_order),
    names_from = c(Cohort, Group), values_from = Statistic, names_sep = "__"
  ) %>%
  arrange(Characteristic_order) %>%
  select(-Characteristic_order)

table1_columns <- unlist(lapply(cohort_order, function(cohort) paste(cohort, group_order, sep = "__")), use.names = FALSE)
table1_wide <- table1_wide %>% select(Characteristic, any_of(table1_columns))

table1_counts <- assignment_summary %>%
  count(Cohort, Group, name = "n") %>%
  tidyr::complete(Cohort = factor(cohort_order, levels = cohort_order), Group = factor(group_order, levels = group_order), fill = list(n = 0))

read_csv_artifact <- function(prefix, suffix) {
  path <- file.path(results_dir, paste0(prefix, suffix))
  if (!file.exists(path)) return(NULL)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

joint_tests <- bind_rows(lapply(cohort_inputs$prefix, function(prefix) {
  x <- read_csv_artifact(prefix, "_eventstudy_joint_tests.csv")
  if (is.null(x) || !nrow(x)) return(NULL)
  x$Analysis_Prefix <- prefix
  x
}))

spline_tests <- bind_rows(lapply(cohort_inputs$prefix, function(prefix) {
  x <- read_csv_artifact(prefix, "_spline_omnibus_tests.csv")
  if (is.null(x) || !nrow(x)) return(NULL)
  x$Analysis_Prefix <- prefix
  x
}))

model_order <- c(
  M0_raw = "M0 raw",
  M1_primary_spline = "M1 primary spline",
  M2_full_spline = "M2 full spline",
  M3_primary_matching_spline = "M3 matching-set spline"
)

trajectory_predictions <- bind_rows(lapply(cohort_inputs$prefix, function(prefix) {
  x <- read_csv_artifact(prefix, "_spline_predicted_trajectories.csv")
  if (is.null(x) || !nrow(x)) return(NULL)
  required <- c("Cohort", "model_id", "Group", "Age_Centered", "pred", "lwr", "upr")
  missing <- setdiff(required, names(x))
  if (length(missing)) stop("Trajectory artifact for ", prefix, " is missing: ", paste(missing, collapse = ", "), call. = FALSE)
  x
})) %>%
  filter(model_id %in% names(model_order)) %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_order),
    Model = factor(unname(model_order[model_id]), levels = unname(model_order)),
    Group = factor(Group, levels = c("Control", "Cancer Case"))
  )

if (!nrow(trajectory_predictions)) stop("No saved 4.1/4.2/4.3/4.5 trajectory predictions were found.", call. = FALSE)

group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")

format_wald_table <- function(x) {
  if (is.null(x) || !nrow(x)) return(NULL)
  if (!"vcov_type" %in% names(x)) x$vcov_type <- NA_character_
  if (!"infer_method" %in% names(x)) x$infer_method <- NA_character_
  x %>%
    mutate(
      Cohort = as.character(Cohort),
      Fstat = formatC(Fstat, format = "f", digits = 3),
      p_value = vapply(p_value, format_p, character(1)),
      df_num = formatC(df_num, format = "f", digits = 0),
      df_denom = ifelse(is.infinite(df_denom), "Inf", formatC(df_denom, format = "f", digits = 0))
    ) %>%
    select(Cohort, Test, Fstat, df_num, df_denom, p_value, vcov_type, infer_method)
}
```

## Table One: matched case and control assignments

The table is summarized at the matched-assignment level. A control participant
may therefore contribute to more than one control assignment because the
risk-set design permits matching with replacement. Repeated FI observations are
not counted as separate Table One participants. Continuous variables are shown
as mean (SD), and categorical variables as n (%).

```{r table-one, results='asis'}
header_top <- c("Characteristic" = 1)
for (cohort in cohort_order) header_top <- c(header_top, setNames(2, cohort_labels[[cohort]]))
subheader <- c("Characteristic")
for (cohort in cohort_order) {
  for (group in group_order) {
    n_value <- table1_counts$n[table1_counts$Cohort == cohort & table1_counts$Group == group]
    subheader <- c(subheader, paste0(group_labels[[group]], "<br>(n = ", n_value, ")"))
  }
}
table1_wide %>%
  kable(format = "html", escape = FALSE, col.names = subheader,
        caption = "Table One. Baseline characteristics of active risk-set matched cancer cases and controls",
        align = c("l", rep("c", ncol(table1_wide) - 1))) %>%
  add_header_above(header_top, escape = FALSE) %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(width = "100%", height = "700px")
```

The current M1/M3 model covariates are index age, race, marital status, and
living arrangement; M2 additionally includes pack-year group, calories,
saturated fat, dietary cholesterol, and alcohol. BMI and physical activity are
displayed when available from the source data as descriptive covariates and are
not silently treated as current 4.x model adjustments.

<!--
The Joint Wald tests are temporarily hidden from rendered HTML. The original
source and chunk labels are retained so this section can be restored later by
removing this HTML comment and the `eval=FALSE, include=FALSE` chunk options.

## Joint Wald tests

These are saved results from the active GLME artifacts. The report preserves the
reported covariance and inference method; it does not relabel model-based
results as CR2.

### Event-study joint Wald tests

```{r event-study-wald, results='asis', eval=FALSE, include=FALSE}
event_wald_display <- format_wald_table(joint_tests)
if (is.null(event_wald_display)) {
  cat("*Saved event-study joint Wald tests are unavailable.*\n")
} else {
  event_wald_display %>%
    kable(format = "html", escape = FALSE, caption = "Saved event-study joint Wald tests") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9)
}
```

### Natural-spline trajectory-shape tests

```{r spline-wald, results='asis', eval=FALSE, include=FALSE}
spline_wald_display <- format_wald_table(spline_tests)
if (is.null(spline_wald_display)) {
  cat("*Saved natural-spline joint Wald tests are unavailable.*\n")
} else {
  spline_wald_display %>%
    kable(format = "html", escape = FALSE, caption = "Saved natural-spline trajectory-shape joint Wald tests") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9)
}
```
-->

## Trajectory plots

Each row is one active cancer cohort, and each column is one saved model
specification. Shaded areas are the saved pointwise 95% confidence intervals;
the vertical dashed line marks the cancer diagnosis or assigned index time.

```{r trajectory-plots, fig.width=16, fig.height=14, out.width='100%'}
p_trajectory <- ggplot(
  trajectory_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_grid(Cohort ~ Model) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "Trajectory plots across model specifications",
    subtitle = "One row per cancer cohort; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.text.y = element_text(angle = 0)
  )
print(p_trajectory)
```

### M2 trajectory plots

The following plot displays only the M2 full-covariate spline model, with one
cancer cohort per row.

```{r m2-trajectory-plots, fig.width=14, fig.height=10, out.width='100%'}
m2_predictions <- trajectory_predictions %>%
  filter(model_id == "M2_full_spline")

p_m2_trajectory <- ggplot(
  m2_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_wrap(~ Cohort, ncol = 2) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "M2 trajectory plots",
    subtitle = "Full-covariate natural-spline model; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text.y = element_text(angle = 0, face = "bold")
  )
print(p_m2_trajectory)
```
