---
title: "Risk-Set Matched Cancer Cohort Summary and GLME Results"
author: "Nemo Zhou"
date: "2026-07-21"
output:
  html_document:
    self_contained: true
    theme: readable
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  4.6_GLME_summary_report.Rmd
Author:  Nemo Zhou
Date started:      2026-07-21
Date last updated: 2026-09-29 (added a separate 4.7 Table One and M2 trajectories)

Purpose:
  Read-only summary report for the active non-survival risk-set matched cancer
  cohorts. The report displays:
    1. An assignment-level Table One for the grouped cohorts, with paired
       cancer-case and cancer-control columns.
    2. A separate Table One with the same structure for the six 4.7
       individual cancer-type cohorts.
    3. Saved natural-spline joint Wald tests retained in the
       source but temporarily hidden from rendered HTML.
    4. Trajectory plots with one row per grouped cancer cohort and four model
       columns: M0, M1, M2, and M3.
    5. Saved M0--M3 trajectory plots for the six 4.7 individual cancer types.
    6. Separate M2-only trajectory plots for grouped and individual cohorts.

  The report reads saved RDS/CSV artifacts only. It does not rebuild matching,
  fit GLME models, run survival-stratified analyses, or save persistent PNG
  files. Reused controls are retained as separate matched assignments, while
  repeated FI observations are collapsed to one row per assignment for each
  Table One so neither table is inflated by the longitudinal measurement count.

Inputs:
  Data/riskset_matched_analysis_long.rds
  Data/riskset_matched_smoking_long.rds
  Data/riskset_matched_obesity_long.rds
  Data/riskset_matched_overall_long.rds
  Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds
  Results/cancer/data/4.1_*, 4.2_*, 4.3_*, 4.5_*, and 4.7_* saved artifacts
  Data/riskset_matched_site_<site>_long.rds and complete matching ledgers

Output:
  Results/cancer/visuals/4.6_GLME_summary_report.html
-->

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyr)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
data_dir <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "Results", "cancer", "data")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = file.path(tempdir(), "4.6_GLME_summary_report_figures", ""),
  dev = "png",
  fig.retina = 1
)
options(knitr.kable.NA = "—")

cohort_inputs <- tibble::tribble(
  ~prefix, ~matched_file,
  "4.1", "riskset_matched_analysis_long.rds",
  "4.2", "riskset_matched_smoking_long.rds",
  "4.3", "riskset_matched_obesity_long.rds",
  "4.5", "riskset_matched_overall_long.rds"
)

cohort_order <- c(
  "Low/Moderate Burden Cohort",
  "High Burden Cohort",
  "Smoking-Related Cancer Cohort",
  "Obesity-Related Cancer Cohort",
  "All Cancer Cohort"
)

cohort_labels <- c(
  "Low/Moderate Burden Cohort" = "Low/moderate burden cancer status",
  "High Burden Cohort" = "High burden cancer status",
  "Smoking-Related Cancer Cohort" = "Smoking-related cancer status",
  "Obesity-Related Cancer Cohort" = "Obesity-related cancer status",
  "All Cancer Cohort" = "All cancer status"
)

group_order <- c("Cancer Case", "Control")
group_labels <- c("Cancer Case" = "Cancer case", "Control" = "Control")

required_matched_columns <- c(
  "id", "match_set", "role", "Cohort", "Group", "index_age",
  "Age_Centered", "fi_score_nocancer", "base_race", "base_marital",
  "base_living", "base_pckgr", "base_calor", "base_sat",
  "base_diet_chol", "base_alco", "post_own_cancer"
)

as_numeric_safely <- function(x) {
  suppressWarnings(as.numeric(as.character(x)))
}

first_nonmissing <- function(x) {
  keep <- !is.na(x)
  if (!any(keep)) return(x[NA_integer_][1])
  x[which(keep)[1]]
}

last_nonmissing <- function(x) {
  keep <- !is.na(x)
  if (!any(keep)) return(x[NA_integer_][1])
  x[tail(which(keep), 1)]
}

mean_sd <- function(x, digits = 2) {
  x <- as_numeric_safely(x)
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_character_)
  paste0(formatC(mean(x), format = "f", digits = digits),
         " (", formatC(stats::sd(x), format = "f", digits = digits), ")")
}

format_p <- function(x) {
  if (is.na(x)) return(NA_character_)
  if (x < 0.001) return("&lt;0.001")
  formatC(x, format = "f", digits = 3)
}

read_matched <- function(prefix, file_name) {
  path <- file.path(data_dir, file_name)
  if (!file.exists(path)) stop("Matched dataset not found: ", path, call. = FALSE)
  source("/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Code/2_data_analysis/2.0_matching_provenance.R")
  provenance <- validate_matching_provenance(path)
  # Do not combine rebuilt cohorts with pre-rebuild fitted curves.
  metadata_path <- file.path(results_dir, paste0(prefix, "_spline_metadata.rds"))
  if (!file.exists(metadata_path)) stop("Rerun the corresponding 4.x GLME before rendering: ", prefix)
  fit_metadata <- readRDS(metadata_path)
  if (!length(fit_metadata) || !all(vapply(fit_metadata, function(x)
      identical(x$input_md5, provenance$input_md5) &&
      identical(x$assignment_md5, provenance$assignment_md5), logical(1))))
    stop("Saved ", prefix, " results use an older matched cohort. Rerun its GLME before rendering.")
  dat <- readRDS(path)
  missing_columns <- setdiff(required_matched_columns, names(dat))
  if (length(missing_columns)) {
    stop("Matched dataset is missing required columns: ",
         paste(missing_columns, collapse = ", "), call. = FALSE)
  }
  dat %>%
    mutate(
      Analysis_Prefix = prefix,
      id = as.character(id),
      match_set = as.character(match_set),
      role = as.character(role),
      Cohort = as.character(Cohort),
      Group = factor(as.character(Group), levels = group_order),
      index_age = as_numeric_safely(index_age),
      Age_Centered = as_numeric_safely(Age_Centered),
      fi_score_nocancer = as_numeric_safely(fi_score_nocancer),
      post_own_cancer = as.logical(post_own_cancer)
    ) %>%
    filter(!post_own_cancer)
}

matched_long <- bind_rows(lapply(seq_len(nrow(cohort_inputs)), function(i) {
  read_matched(cohort_inputs$prefix[[i]], cohort_inputs$matched_file[[i]])
}))

raw_path <- file.path(data_dir, "FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")
baseline_extra <- NULL
if (file.exists(raw_path)) {
  raw <- readRDS(raw_path)
  raw_names <- names(raw)
  baseline_extra <- raw %>%
    mutate(id = as.character(id), cycle = as.character(cycle)) %>%
    filter(cycle == "86") %>%
    distinct(id, .keep_all = TRUE) %>%
    transmute(
      id,
      base_bmi = if ("bmi" %in% raw_names) as_numeric_safely(bmi) else NA_real_,
      base_activity = if ("act" %in% raw_names) as_numeric_safely(act) else NA_real_
    )
}

matched_assignments <- bind_rows(lapply(seq_len(nrow(cohort_inputs)), function(i) {
  read_matching_assignments(file.path(data_dir, cohort_inputs$matched_file[i])) %>%
    mutate(Analysis_Prefix = cohort_inputs$prefix[i])
}))

assignment_summary <- matched_long %>%
  arrange(Analysis_Prefix, Cohort, match_set, id, role, Age_Centered) %>%
  group_by(Analysis_Prefix, Cohort, match_set, id, role, Group) %>%
  summarize(
    index_age = first_nonmissing(index_age),
    fi_first_observed = first_nonmissing(fi_score_nocancer),
    fi_last_pre_index = last_nonmissing(fi_score_nocancer[Age_Centered < 0]),
    n_fi_observations = sum(!is.na(fi_score_nocancer)),
    n_pre_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered < 0),
    n_post_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered >= 0),
    base_race = first_nonmissing(base_race),
    base_marital = first_nonmissing(base_marital),
    base_living = first_nonmissing(base_living),
    base_pckgr = first_nonmissing(base_pckgr),
    base_calor = first_nonmissing(base_calor),
    base_sat = first_nonmissing(base_sat),
    base_diet_chol = first_nonmissing(base_diet_chol),
    base_alco = first_nonmissing(base_alco),
    .groups = "drop"
  )

assignment_summary <- complete_assignment_summary(matched_assignments, assignment_summary,
  keys = c("Analysis_Prefix", "Cohort", "match_set", "id", "role", "Group"),
  count_columns = c("n_fi_observations", "n_pre_index_fi", "n_post_index_fi"))

if (!is.null(baseline_extra)) {
  assignment_summary <- assignment_summary %>% left_join(baseline_extra, by = "id")
}

assignment_summary <- assignment_summary %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_order),
    Group = factor(Group, levels = group_order)
  )

continuous_vars <- c(
  index_age = "Age at diagnosis/index, years",
  fi_first_observed = "Frailty index, first observed analytic cycle",
  fi_last_pre_index = "Frailty index, last pre-index observation",
  base_calor = "Calories",
  base_sat = "Saturated fat",
  base_diet_chol = "Dietary cholesterol",
  base_alco = "Alcohol"
)

optional_continuous_vars <- c(
  base_bmi = "BMI",
  base_activity = "Physical activity"
)
if (any(names(optional_continuous_vars) %in% names(assignment_summary))) {
  continuous_vars <- c(
    continuous_vars,
    optional_continuous_vars[names(optional_continuous_vars) %in% names(assignment_summary)]
  )
}

categorical_vars <- c(
  base_race = "Race",
  base_marital = "Marital status",
  base_living = "Living arrangement",
  base_pckgr = "Pack-year group"
)

continuous_vars <- continuous_vars[vapply(
  names(continuous_vars),
  function(v) any(!is.na(assignment_summary[[v]])),
  logical(1)
)]
categorical_vars <- categorical_vars[vapply(
  names(categorical_vars),
  function(v) any(!is.na(assignment_summary[[v]])),
  logical(1)
)]

summarize_continuous <- function(dat, var, label) {
  dat %>%
    group_by(Group) %>%
    summarize(Statistic = mean_sd(.data[[var]]), .groups = "drop") %>%
    mutate(Variable = label, Level = NA_character_, Level_order = 0L)
}

summarize_categorical <- function(dat, var, label) {
  pack_year_levels <- c(
    "<10 pack-years", "10-24 pack-years", "25-44 pack-years",
    "45-64 pack-years", "65+ pack-years", "Missing", "Never"
  )

  level_summary <- dat %>%
    filter(!is.na(.data[[var]])) %>%
    count(Group, Level = as.character(.data[[var]]), name = "n") %>%
    mutate(
      Level_order = if (var == "base_pckgr") {
        match(Level, pack_year_levels)
      } else {
        match(Level, sort(unique(Level)))
      }
    ) %>%
    group_by(Group) %>%
    mutate(Statistic = paste0(formatC(100 * n / sum(n), format = "f", digits = 1), "%")) %>%
    ungroup() %>%
    transmute(
      Group, Statistic, Variable = label,
      Level = htmltools::htmlEscape(Level, attribute = FALSE), Level_order
    )

  variable_row <- tibble::tibble(
    Group = factor(group_order, levels = group_order),
    Statistic = "",
    Variable = label,
    Level = NA_character_,
    Level_order = 0L
  )

  bind_rows(variable_row, level_summary)
}

table1_long <- bind_rows(lapply(seq_len(nrow(cohort_inputs)), function(i) {
  prefix_data <- assignment_summary %>% filter(Analysis_Prefix == cohort_inputs$prefix[[i]])
  bind_rows(lapply(unique(as.character(prefix_data$Cohort)), function(cohort_name) {
    cohort_data <- prefix_data %>% filter(as.character(Cohort) == cohort_name)
    bind_rows(
      lapply(names(continuous_vars), function(v) summarize_continuous(cohort_data, v, continuous_vars[[v]])),
      lapply(names(categorical_vars), function(v) summarize_categorical(cohort_data, v, categorical_vars[[v]]))
    ) %>% mutate(Cohort = cohort_name)
  }))
}))

table1_row_order <- c(unname(continuous_vars), unname(categorical_vars))
table1_long <- table1_long %>%
  mutate(
    Variable = factor(Variable, levels = table1_row_order),
    Group = factor(Group, levels = group_order),
    Cohort = factor(Cohort, levels = cohort_order)
  ) %>%
  mutate(
    variable_order = match(as.character(Variable), table1_row_order)
  )

table1_wide <- table1_long %>%
  mutate(
    Characteristic = if_else(is.na(Level), as.character(Variable), paste0("&nbsp;&nbsp;", as.character(Level))),
    Characteristic_order = variable_order * 1000 + Level_order
  ) %>%
  select(Cohort, Group, Characteristic, Characteristic_order, Statistic) %>%
  tidyr::pivot_wider(
    id_cols = c(Characteristic, Characteristic_order),
    names_from = c(Cohort, Group), values_from = Statistic, names_sep = "__"
  ) %>%
  arrange(Characteristic_order) %>%
  select(-Characteristic_order)

table1_columns <- unlist(lapply(cohort_order, function(cohort) paste(cohort, group_order, sep = "__")), use.names = FALSE)
table1_wide <- table1_wide %>% select(Characteristic, any_of(table1_columns))

table1_counts <- assignment_summary %>%
  count(Cohort, Group, name = "n") %>%
  tidyr::complete(Cohort = factor(cohort_order, levels = cohort_order), Group = factor(group_order, levels = group_order), fill = list(n = 0))

read_csv_artifact <- function(prefix, suffix) {
  path <- file.path(results_dir, paste0(prefix, suffix))
  if (!file.exists(path)) return(NULL)
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

spline_tests <- bind_rows(lapply(cohort_inputs$prefix, function(prefix) {
  x <- read_csv_artifact(prefix, "_spline_omnibus_tests.csv")
  if (is.null(x) || !nrow(x)) return(NULL)
  x$Analysis_Prefix <- prefix
  x
}))

model_order <- c(
  M0_raw = "M0 raw",
  M1_primary_spline = "M1 primary spline",
  M2_full_spline = "M2 full spline",
  M3_primary_matching_spline = "M3 matching-set spline"
)

individual_type_inputs <- tibble::tribble(
  ~site, ~cohort_label, ~prefix,
  "prostate", "Prostate Cancer Cohort", "4.7_prostate",
  "lung", "Lung/Trachea/Bronchus Cancer Cohort", "4.7_lung",
  "colorectal", "Colorectal/Anal Cancer Cohort", "4.7_colorectal",
  "bladder", "Bladder Cancer Cohort", "4.7_bladder",
  "kidney", "Kidney/Other Urinary Cancer Cohort", "4.7_kidney",
  "pancreas", "Pancreatic Cancer Cohort", "4.7_pancreas"
)
individual_type_labels <- c(
  "Prostate Cancer Cohort" = "Prostate",
  "Lung/Trachea/Bronchus Cancer Cohort" = "Lung / trachea / bronchus",
  "Colorectal/Anal Cancer Cohort" = "Colorectal / anal",
  "Bladder Cancer Cohort" = "Bladder",
  "Kidney/Other Urinary Cancer Cohort" = "Kidney / other urinary",
  "Pancreatic Cancer Cohort" = "Pancreas"
)

individual_type_run_path <- file.path(results_dir, "4.7_individual_type_run_status.csv")
if (!file.exists(individual_type_run_path)) {
  stop("Saved 4.7 run status was not found: ", individual_type_run_path, call. = FALSE)
}
individual_type_run_status <- read.csv(
  individual_type_run_path, stringsAsFactors = FALSE, check.names = FALSE
)
required_run_columns <- c(
  "site", "status", "input_md5", "matched_md5", "builder_md5",
  "matching_engine_md5", "glme_engine_md5", "m1_status"
)
missing_run_columns <- setdiff(required_run_columns, names(individual_type_run_status))
if (length(missing_run_columns)) {
  stop("4.7 run-status file is missing: ", paste(missing_run_columns, collapse = ", "), call. = FALSE)
}
if (anyDuplicated(individual_type_run_status$site)) {
  stop("4.7 run-status file has duplicate site rows.", call. = FALSE)
}

individual_type_source_paths <- c(
  input = raw_path,
  builder = file.path(project_dir, "Code", "2_data_analysis", "2.6_create_riskset_individual_types.R"),
  matching_engine = file.path(project_dir, "Code", "2_data_analysis", "2.0_riskset_matching_functions.R"),
  glme_engine = file.path(project_dir, "Code", "2_data_analysis", "4.0_GLME_spline_functions.R")
)
if (any(!file.exists(individual_type_source_paths))) {
  stop("A current 4.7 source or input file is missing: ",
       paste(names(individual_type_source_paths)[!file.exists(individual_type_source_paths)], collapse = ", "),
       call. = FALSE)
}
individual_type_source_hashes <- vapply(
  individual_type_source_paths, function(path) unname(tools::md5sum(path)), character(1)
)

read_individual_type_artifacts <- function(site, cohort_label, prefix) {
  run_row <- individual_type_run_status[
    individual_type_run_status$site == site, , drop = FALSE
  ]
  if (nrow(run_row) != 1L || !identical(as.character(run_row$status[[1]]), "fit") ||
      !as.character(run_row$m1_status[[1]]) %in% c("fit", "fit_boundary_matching_variance_zero")) {
    stop("4.7 site ", site, " is not recorded as a successful current fit; inspect ",
         individual_type_run_path, call. = FALSE)
  }
  hash_columns <- c("input_md5", "builder_md5", "matching_engine_md5", "glme_engine_md5")
  if (!all(vapply(seq_along(hash_columns), function(i) {
    identical(as.character(run_row[[hash_columns[[i]]]][[1]]), individual_type_source_hashes[[i]])
  }, logical(1)))) {
    stop("4.7 site ", site, " was run with different source/input files. Rerun 4.7 before rendering.",
         call. = FALSE)
  }

  matched_path <- file.path(data_dir, paste0("riskset_matched_site_", site, "_long.rds"))
  provenance <- validate_matching_provenance(matched_path)
  if (!identical(as.character(run_row$matched_md5[[1]]), provenance$input_md5)) {
    stop("4.7 site ", site, " results do not match the current matched input. Rerun 4.7 before rendering.",
         call. = FALSE)
  }
  metadata_path <- file.path(results_dir, paste0(prefix, "_spline_metadata.rds"))
  if (!file.exists(metadata_path)) stop("4.7 metadata is missing for site ", site, call. = FALSE)
  fit_metadata <- readRDS(metadata_path)
  if (!length(fit_metadata) || !all(vapply(fit_metadata, function(x)
      identical(x$input_md5, provenance$input_md5) &&
      identical(x$assignment_md5, provenance$assignment_md5), logical(1)))) {
    stop("4.7 site ", site, " saved results use older matched data. Rerun 4.7 before rendering.",
         call. = FALSE)
  }

  matched_data <- readRDS(matched_path)
  missing_matched_columns <- setdiff(required_matched_columns, names(matched_data))
  if (length(missing_matched_columns)) {
    stop("4.7 matched data for ", site, " is missing: ",
         paste(missing_matched_columns, collapse = ", "), call. = FALSE)
  }
  matched_data <- matched_data %>%
    mutate(
      Analysis_Prefix = site,
      id = as.character(id),
      match_set = as.character(match_set),
      role = as.character(role),
      Cohort = as.character(Cohort),
      Group = factor(as.character(Group), levels = group_order),
      index_age = as_numeric_safely(index_age),
      Age_Centered = as_numeric_safely(Age_Centered),
      fi_score_nocancer = as_numeric_safely(fi_score_nocancer),
      post_own_cancer = as.logical(post_own_cancer)
    ) %>%
    filter(!post_own_cancer)
  matched_assignments <- read_matching_assignments(matched_path) %>%
    mutate(
      Analysis_Prefix = site,
      id = as.character(id),
      match_set = as.character(match_set),
      role = as.character(role),
      Cohort = as.character(Cohort),
      Group = factor(as.character(Group), levels = group_order)
    )

  model_status <- read_csv_artifact(prefix, "_spline_model_status.csv")
  predictions <- read_csv_artifact(prefix, "_spline_predicted_trajectories.csv")
  if (is.null(model_status) || is.null(predictions)) {
    stop("One or more saved 4.7 model artifacts are missing for site ", site, call. = FALSE)
  }
  required_status <- c("Cohort", "model_id", "status", "n_obs", "n_id", "convergence")
  required_predictions <- c("Cohort", "model_id", "Group", "Age_Centered", "pred", "lwr", "upr")
  required_artifacts <- list(
    model_status = required_status,
    predictions = required_predictions
  )
  for (artifact_name in names(required_artifacts)) {
    artifact <- get(artifact_name, inherits = FALSE)
    missing <- setdiff(required_artifacts[[artifact_name]], names(artifact))
    if (length(missing)) {
      stop("4.7 ", artifact_name, " for ", site, " is missing: ",
           paste(missing, collapse = ", "), call. = FALSE)
    }
  }
  for (artifact in list(model_status, predictions)) {
    if (!all(as.character(artifact$Cohort) == cohort_label)) {
      stop("4.7 cohort label does not match the report configuration for site ", site, call. = FALSE)
    }
  }
  if (anyDuplicated(model_status$model_id)) {
    stop("4.7 site ", site, " has duplicate model rows.", call. = FALSE)
  }
  expected_models <- names(model_order)
  if (!setequal(model_status$model_id, expected_models)) {
    stop("4.7 site ", site, " does not contain the complete M0--M3 result set.", call. = FALSE)
  }
  m1_status <- model_status[model_status$model_id == "M1_primary_spline", , drop = FALSE]
  if (nrow(m1_status) != 1L ||
      !as.character(m1_status$status[[1]]) %in% c("fit", "fit_boundary_matching_variance_zero") ||
      !isTRUE(as.logical(m1_status$convergence[[1]]))) {
    stop("4.7 M1 failed its required convergence check for site ", site, call. = FALSE)
  }
  list(
    matched_data = matched_data,
    matched_assignments = matched_assignments,
    model_status = model_status,
    predictions = predictions
  )
}

individual_type_artifacts <- lapply(seq_len(nrow(individual_type_inputs)), function(i) {
  read_individual_type_artifacts(
    individual_type_inputs$site[[i]], individual_type_inputs$cohort_label[[i]],
    individual_type_inputs$prefix[[i]]
  )
})

individual_type_matched_long <- bind_rows(lapply(individual_type_artifacts, `[[`, "matched_data"))
individual_type_matched_assignments <- bind_rows(
  lapply(individual_type_artifacts, `[[`, "matched_assignments")
)

individual_type_assignment_summary <- individual_type_matched_long %>%
  arrange(Analysis_Prefix, Cohort, match_set, id, role, Age_Centered) %>%
  group_by(Analysis_Prefix, Cohort, match_set, id, role, Group) %>%
  summarize(
    index_age = first_nonmissing(index_age),
    fi_first_observed = first_nonmissing(fi_score_nocancer),
    fi_last_pre_index = last_nonmissing(fi_score_nocancer[Age_Centered < 0]),
    n_fi_observations = sum(!is.na(fi_score_nocancer)),
    n_pre_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered < 0),
    n_post_index_fi = sum(!is.na(fi_score_nocancer) & Age_Centered >= 0),
    base_race = first_nonmissing(base_race),
    base_marital = first_nonmissing(base_marital),
    base_living = first_nonmissing(base_living),
    base_pckgr = first_nonmissing(base_pckgr),
    base_calor = first_nonmissing(base_calor),
    base_sat = first_nonmissing(base_sat),
    base_diet_chol = first_nonmissing(base_diet_chol),
    base_alco = first_nonmissing(base_alco),
    .groups = "drop"
  )

individual_type_assignment_summary <- complete_assignment_summary(
  individual_type_matched_assignments,
  individual_type_assignment_summary,
  keys = c("Analysis_Prefix", "Cohort", "match_set", "id", "role", "Group"),
  count_columns = c("n_fi_observations", "n_pre_index_fi", "n_post_index_fi")
)
if (!is.null(baseline_extra)) {
  individual_type_assignment_summary <- individual_type_assignment_summary %>%
    left_join(baseline_extra, by = "id")
}
individual_type_assignment_summary <- individual_type_assignment_summary %>%
  mutate(
    Cohort = factor(Cohort, levels = individual_type_inputs$cohort_label),
    Group = factor(Group, levels = group_order)
  )

individual_type_table1_long <- bind_rows(lapply(seq_len(nrow(individual_type_inputs)), function(i) {
  site_data <- individual_type_assignment_summary %>%
    filter(Analysis_Prefix == individual_type_inputs$site[[i]])
  bind_rows(lapply(unique(as.character(site_data$Cohort)), function(cohort_name) {
    cohort_data <- site_data %>% filter(as.character(Cohort) == cohort_name)
    bind_rows(
      lapply(names(continuous_vars), function(v) summarize_continuous(cohort_data, v, continuous_vars[[v]])),
      lapply(names(categorical_vars), function(v) summarize_categorical(cohort_data, v, categorical_vars[[v]]))
    ) %>% mutate(Cohort = cohort_name)
  }))
})) %>%
  mutate(
    Variable = factor(Variable, levels = table1_row_order),
    Group = factor(Group, levels = group_order),
    Cohort = factor(Cohort, levels = individual_type_inputs$cohort_label),
    variable_order = match(as.character(Variable), table1_row_order)
  )

individual_type_table1_wide <- individual_type_table1_long %>%
  mutate(
    Characteristic = if_else(is.na(Level), as.character(Variable), paste0("&nbsp;&nbsp;", as.character(Level))),
    Characteristic_order = variable_order * 1000 + Level_order
  ) %>%
  select(Cohort, Group, Characteristic, Characteristic_order, Statistic) %>%
  tidyr::pivot_wider(
    id_cols = c(Characteristic, Characteristic_order),
    names_from = c(Cohort, Group), values_from = Statistic, names_sep = "__"
  ) %>%
  arrange(Characteristic_order) %>%
  select(-Characteristic_order)
individual_type_table1_columns <- unlist(lapply(
  individual_type_inputs$cohort_label,
  function(cohort) paste(cohort, group_order, sep = "__")
), use.names = FALSE)
individual_type_table1_wide <- individual_type_table1_wide %>%
  select(Characteristic, any_of(individual_type_table1_columns))
individual_type_table1_counts <- individual_type_assignment_summary %>%
  count(Cohort, Group, name = "n") %>%
  tidyr::complete(
    Cohort = factor(individual_type_inputs$cohort_label,
                    levels = individual_type_inputs$cohort_label),
    Group = factor(group_order, levels = group_order),
    fill = list(n = 0)
  )

individual_type_predictions <- bind_rows(lapply(individual_type_artifacts, `[[`, "predictions")) %>%
  filter(model_id %in% names(model_order)) %>%
  mutate(
    Cohort = factor(Cohort, levels = individual_type_inputs$cohort_label),
    Model = factor(unname(model_order[model_id]), levels = unname(model_order)),
    Group = factor(as.character(Group), levels = c("Control", "Cancer Case"))
  )
if (!nrow(individual_type_predictions)) {
  stop("No saved 4.7 individual-type trajectory predictions were found.", call. = FALSE)
}

trajectory_predictions <- bind_rows(lapply(cohort_inputs$prefix, function(prefix) {
  x <- read_csv_artifact(prefix, "_spline_predicted_trajectories.csv")
  if (is.null(x) || !nrow(x)) return(NULL)
  required <- c("Cohort", "model_id", "Group", "Age_Centered", "pred", "lwr", "upr")
  missing <- setdiff(required, names(x))
  if (length(missing)) stop("Trajectory artifact for ", prefix, " is missing: ", paste(missing, collapse = ", "), call. = FALSE)
  x
})) %>%
  filter(model_id %in% names(model_order)) %>%
  mutate(
    Cohort = factor(Cohort, levels = cohort_order),
    Model = factor(unname(model_order[model_id]), levels = unname(model_order)),
    Group = factor(Group, levels = c("Control", "Cancer Case"))
  )

if (!nrow(trajectory_predictions)) stop("No saved 4.1/4.2/4.3/4.5 trajectory predictions were found.", call. = FALSE)

group_cols <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")

format_wald_table <- function(x) {
  if (is.null(x) || !nrow(x)) return(NULL)
  if (!"vcov_type" %in% names(x)) x$vcov_type <- NA_character_
  if (!"infer_method" %in% names(x)) x$infer_method <- NA_character_
  x %>%
    mutate(
      Cohort = as.character(Cohort),
      Fstat = formatC(Fstat, format = "f", digits = 3),
      p_value = vapply(p_value, format_p, character(1)),
      df_num = formatC(df_num, format = "f", digits = 0),
      df_denom = ifelse(is.infinite(df_denom), "Inf", formatC(df_denom, format = "f", digits = 0))
    ) %>%
    select(Cohort, Test, Fstat, df_num, df_denom, p_value, vcov_type, infer_method)
}
```

## Table One: matched case and control assignments

The table includes all assignments in the complete matched ledger, including participants with no FI observations. FI statistics use available observations after own-cancer censoring; fitted models further apply their covariate-completeness requirements. A control participant
may therefore contribute to more than one control assignment because the
risk-set design permits matching with replacement. Repeated FI observations are
not counted as separate Table One participants. Continuous variables are shown
as mean (SD), and categorical variables as n (%).

```{r table-one, results='asis'}
header_top <- c("Characteristic" = 1)
for (cohort in cohort_order) header_top <- c(header_top, setNames(2, cohort_labels[[cohort]]))
subheader <- c("Characteristic")
for (cohort in cohort_order) {
  for (group in group_order) {
    n_value <- table1_counts$n[table1_counts$Cohort == cohort & table1_counts$Group == group]
    subheader <- c(subheader, paste0(group_labels[[group]], "<br>(n = ", n_value, ")"))
  }
}
table1_wide %>%
  kable(format = "html", escape = FALSE, col.names = subheader,
        caption = "Table One. Baseline characteristics of active risk-set matched cancer cases and controls",
        align = c("l", rep("c", ncol(table1_wide) - 1))) %>%
  add_header_above(header_top, escape = FALSE) %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(width = "100%", height = "700px")
```

The current M1/M3 model covariates are index age, race, marital status, and
living arrangement; M2 additionally includes pack-year group, calories,
saturated fat, dietary cholesterol, and alcohol. BMI and physical activity are
displayed when available from the source data as descriptive covariates and are
not silently treated as current 4.x model adjustments.

## Table One (4.7): matched individual cancer-type assignments

This separate table uses the same assignment-level summaries and paired
cancer-case/control columns as the grouped-cohort Table One. It is kept separate
to make the six site-specific cohorts easier to read. Reused controls remain
separate assignments, and assignments without FI observations remain in the
denominators.

```{r individual-type-table-one, results='asis'}
individual_type_header_top <- c("Characteristic" = 1)
for (cohort in individual_type_inputs$cohort_label) {
  individual_type_header_top <- c(
    individual_type_header_top,
    setNames(2, individual_type_labels[[cohort]])
  )
}
individual_type_subheader <- c("Characteristic")
for (cohort in individual_type_inputs$cohort_label) {
  for (group in group_order) {
    n_value <- individual_type_table1_counts$n[
      individual_type_table1_counts$Cohort == cohort &
        individual_type_table1_counts$Group == group
    ]
    individual_type_subheader <- c(
      individual_type_subheader,
      paste0(group_labels[[group]], "<br>(n = ", n_value, ")")
    )
  }
}
individual_type_table1_wide %>%
  kable(
    format = "html", escape = FALSE, col.names = individual_type_subheader,
    caption = "Table One (4.7). Baseline characteristics of matched individual cancer-type cases and controls",
    align = c("l", rep("c", ncol(individual_type_table1_wide) - 1))
  ) %>%
  add_header_above(individual_type_header_top, escape = FALSE) %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(width = "100%", height = "700px")
```

<!--
The Joint Wald tests are temporarily hidden from rendered HTML. The original
source and chunk labels are retained so this section can be restored later by
removing this HTML comment and the `eval=FALSE, include=FALSE` chunk options.

## Joint Wald tests

These are saved results from the active GLME artifacts. The report preserves the
reported covariance and inference method; it does not relabel model-based
results as CR2.

### Natural-spline trajectory-shape tests

```{r spline-wald, results='asis', eval=FALSE, include=FALSE}
spline_wald_display <- format_wald_table(spline_tests)
if (is.null(spline_wald_display)) {
  cat("*Saved natural-spline joint Wald tests are unavailable.*\n")
} else {
  spline_wald_display %>%
    kable(format = "html", escape = FALSE, caption = "Saved natural-spline trajectory-shape joint Wald tests") %>%
    kable_styling(full_width = TRUE, bootstrap_options = c("striped", "condensed"), font_size = 9)
}
```
-->

## Trajectory plots

Each row is one active cancer cohort, and each column is one saved model
specification. Shaded areas are the saved pointwise 95% confidence intervals;
the vertical dashed line marks the cancer diagnosis or assigned index time.

```{r trajectory-plots, fig.width=16, fig.height=14, out.width='100%'}
p_trajectory <- ggplot(
  trajectory_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_grid(Cohort ~ Model) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "Trajectory plots across model specifications",
    subtitle = "One row per cancer cohort; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.text.y = element_text(angle = 0)
  )
print(p_trajectory)
```

## Individual cancer-type trajectories (4.7)

The six site-specific cohorts use the same M0--M3 model definitions as the
grouped analyses, with a separate risk-set match for each cancer type. Shaded
areas show the saved pointwise 95% confidence intervals. Curves stop at each
site's saved prediction limits.

### Individual-type trajectories by model

Each row is one 4.7 site cohort, and each column is one saved model. Shaded
areas are the saved pointwise 95% confidence intervals. Lines stop at the
site/model prediction limits shown in the table above.

```{r individual-type-trajectories, fig.width=16, fig.height=16, out.width='100%'}
p_individual_type_trajectory <- ggplot(
  individual_type_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_grid(Cohort ~ Model) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "Individual cancer-type trajectories across model specifications",
    subtitle = "Six site-specific cohorts; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    strip.text.y = element_text(angle = 0)
  )
print(p_individual_type_trajectory)
```

### M2 trajectory plots

These plots display only the M2 full-covariate spline model. The first plot has
one grouped cancer cohort per panel.

```{r m2-trajectory-plots, fig.width=14, fig.height=10, out.width='100%'}
m2_predictions <- trajectory_predictions %>%
  filter(model_id == "M2_full_spline")

p_m2_trajectory <- ggplot(
  m2_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_wrap(~ Cohort, ncol = 2) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "M2 trajectory plots",
    subtitle = "Full-covariate natural-spline model; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text.y = element_text(angle = 0, face = "bold")
  )
print(p_m2_trajectory)
```

#### Individual cancer-type M2 trajectories (4.7)

The following plot shows the saved M2 trajectories for the six site-specific
cohorts. Curves stop at each site's saved prediction limits.

```{r individual-type-m2-trajectories, fig.width=14, fig.height=10, out.width='100%'}
individual_type_m2_predictions <- individual_type_predictions %>%
  filter(model_id == "M2_full_spline")

p_individual_type_m2_trajectory <- ggplot(
  individual_type_m2_predictions,
  aes(x = Age_Centered, y = pred, group = Group, color = Group, fill = Group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.16, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey35") +
  facet_wrap(~ Cohort, ncol = 2) +
  scale_color_manual(values = group_cols) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4)) +
  labs(
    title = "Individual cancer-type M2 trajectories",
    subtitle = "Full-covariate natural-spline model; shaded areas are saved pointwise 95% confidence intervals",
    x = "Years relative to cancer diagnosis / assigned index",
    y = "Predicted frailty index",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
print(p_individual_type_m2_trajectory)
```
