---
title: "Frailty Trajectories Before and After Cancer (Age Scale)"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 4_age_drawing.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Creates age-scale stochastic-drawing frailty trajectory analyses before and after cancer diagnosis.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary) 
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables

Assuming original date variables are in months since a reference epoch. Dividing by 12 converts to continuous years (Age).

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    years_since_baseline = (worked_rtmnyr - first(worked_rtmnyr)) / 12
  ) %>%
  ungroup()
```

## 4. Extract 1986 Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build the Analytic Longitudinal Dataset (Age Scale)

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    baseline_time = first(worked_rtmnyr),
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    
    # Prevalent: Cancer diagnosed before their first valid FI age in the target cycles
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    Cancer_Status_TimeVarying = if_else(
      Ever_Cancer == 1L & age_at_cycle >= true_age_at_cancer, 
      1L, 
      0L
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  ungroup()

# Z-score base age for LME adjustment
base_age_scaling <- analysis_long_age %>%
  distinct(id, base_age) %>%
  mutate(base_age_z = as.numeric(scale(base_age))) %>%
  select(id, base_age_z)

analysis_long_age <- analysis_long_age %>%
  left_join(base_age_scaling, by = "id") %>%
  mutate(
    terminal_decline = if_else(
      !is.na(years_to_death) & years_to_death >= -1 & years_to_death <= 0, 
      1L, 
      0L
    ),
    Group_Label = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

## 6. Check Follow-Up Structure

```{r followup-checks}
zero_fu_check <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(age_at_cycle, na.rm = TRUE),
    max_fu_yrs = max(age_at_cycle, na.rm = TRUE) - min(age_at_cycle, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu_yrs == 0)

cat("Total observations in analysis_long_age =", nrow(analysis_long_age), "\n")
cat("Patients with zero follow-up years =", nrow(zero_fu_check), "\n")

if (nrow(zero_fu_check) > 0) {
  print(zero_fu_check)
} else {
  cat("No zero follow-up patients remain.\n")
}
```

## 7. Stochastic Imputation of Index Age for Controls

Controls receive a randomly drawn event age from a case within their same 5-year baseline age bin. Crucially, the drawn age must fall within the control's observed follow-up window `[min_age, max_age]`.

```{r impute-index-age}
set.seed(2026)

patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = first(base_age),
    min_age = first(min_age),
    max_age = first(max_age),
    is_case = first(Ever_Cancer),
    true_age_at_cancer = first(true_age_at_cancer),
    .groups = "drop"
  ) %>%
  mutate(
    age_bin = floor(base_age / 5) * 5 
  )

cases_summary <- patient_summary_age %>%
  filter(is_case == 1, !is.na(true_age_at_cancer), true_age_at_cancer > min_age)

controls_summary <- patient_summary_age %>%
  filter(is_case == 0)

imputed_ages <- rep(NA_real_, nrow(controls_summary))

for (i in seq_len(nrow(controls_summary))) {
  current_control <- controls_summary[i, ]
  
  eligible_cases <- cases_summary %>%
    filter(
      age_bin == current_control$age_bin,
      true_age_at_cancer >= current_control$min_age,
      true_age_at_cancer <= current_control$max_age
    )
  
  if (nrow(eligible_cases) > 0) {
    imputed_ages[i] <- sample(eligible_cases$true_age_at_cancer, size = 1)
  } else {
    imputed_ages[i] <- NA_real_
  }
}

controls_summary$assigned_event_age <- imputed_ages
cases_summary$assigned_event_age <- cases_summary$true_age_at_cancer

assigned_ages_df <- bind_rows(cases_summary, controls_summary) %>%
  filter(!is.na(assigned_event_age)) %>%
  select(id, assigned_event_age) %>%
  distinct(id, .keep_all = TRUE)

cat("Controls dropped due to no eligible donor in their age window =", sum(is.na(imputed_ages)), "\n")
```

## 8. Generate Time Scale Variables for LME

Merging the assigned dates back and creating the pre/post spline variables centered at the index age.

```{r generate-splines}
model_data_age <- analysis_long_age %>%
  inner_join(assigned_ages_df, by = "id") %>%
  mutate(
    Group = Ever_Cancer,
    
    # 1. Center age around the event age
    Age_Centered = age_at_cycle - assigned_event_age,
    
    # 2. Create the knot indicator and post-event continuous time
    Post = if_else(Age_Centered >= 0, 1L, 0L),
    Age_Post = if_else(Post == 1L, Age_Centered, 0)
  )

cat("Final model_data_age row count =", nrow(model_data_age), "\n")
```

## 9. Descriptive Analysis Before LME

```{r descriptive-data}
baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- model_data_age %>% 
  group_by(id) %>%
  summarize(
    Group_Label = first(Group_Label),
    Ever_Cancer = first(Ever_Cancer),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    true_age_at_cancer = first(true_age_at_cancer),
    assigned_event_age = first(assigned_event_age),
    num_obs = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = first(min_age),
    max_age = first(max_age),
    max_fu_yrs = first(max_followup_yrs), # <-- Fix here!
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id") %>%
  mutate(
    Index_Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

### Table 1: Baseline Characteristics

```{r table1}
table1_age <- followup_by_patient_age %>%
  select(
    Index_Group,
    base_age,
    assigned_event_age,
    fi_baseline,
    base_race,
    base_marital,
    base_living,
    base_pckgr
  ) %>%
  gtsummary::tbl_summary(
    by = Index_Group,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      base_age ~ "Age at Study Entry (years)",
      assigned_event_age ~ "Age at Cancer / Assigned Index Age",
      fi_baseline ~ "Baseline Frailty Index"
    ),
    missing_text = "(Missing)"
  ) %>%
  gtsummary::add_p() %>%
  gtsummary::add_overall() %>%
  gtsummary::bold_labels()

table1_age
```

### Assigned Index Age Summary

```{r index-age-summary}
index_age_summary <- followup_by_patient_age %>%
  group_by(Index_Group) %>%
  summarize(
    N = n(),
    Mean_Index_Age = round(mean(assigned_event_age, na.rm = TRUE), 2),
    SD_Index_Age = round(sd(assigned_event_age, na.rm = TRUE), 2),
    Median_Index_Age = round(median(assigned_event_age, na.rm = TRUE), 2),
    Min_Index_Age = round(min(assigned_event_age, na.rm = TRUE), 2),
    Max_Index_Age = round(max(assigned_event_age, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(as.data.frame(index_age_summary))

# Test for difference in event age distributions
wilcox.test(assigned_event_age ~ Index_Group, data = followup_by_patient_age)
```

### Constraint Sanity Checks

```{r sanity-checks}
control_age_check <- followup_by_patient_age %>%
  filter(Index_Group == "Control") %>%
  summarize(
    n_controls = n(),
    n_bad_age_early = sum(assigned_event_age < min_age, na.rm = TRUE),
    n_bad_age_late = sum(assigned_event_age > max_age, na.rm = TRUE)
  )

print(as.data.frame(control_age_check))
```

### Descriptive Plots

```{r descriptive-plots, fig.width=8, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Index_Group, color = Index_Group)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Distribution of Age at Index Event",
    x = "Assigned Index Age (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  )

print(p_index_density_age)

p_index_box_age <- ggplot(followup_by_patient_age, aes(x = Index_Group, y = assigned_event_age, fill = Index_Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.25) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Age at Index Event by Group",
    x = NULL,
    y = "Assigned Index Age (years)"
  ) +
  theme(legend.position = "none")

print(p_index_box_age)
```

```{r}
## 10. Run the Piecewise LME Model

# We use the bobyqa optimizer because LMEs with both random intercepts 
# and random slopes often fail to converge with the default optimizer.
model_piecewise <- lmer(
  fi_score_nocancer ~ Group * Age_Centered + 
                      Group * Age_Post + 
                      base_age_z + base_race + base_marital + base_pckgr +
                      (1 + Age_Centered | id), 
  data = model_data_age,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

# Print the summary
summary(model_piecewise)

# Extract fixed effects for clean reporting
model_results <- as.data.frame(summary(model_piecewise)$coefficients)
model_results$Term <- rownames(model_results)
rownames(model_results) <- NULL
print(model_results)
```

```{r}
## 11. Generate Predictions using predict()

plot_grid <- expand.grid(
  Age_Centered = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    # 1. Recreate the spline
    Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
    
    # 2. Hold continuous covariates at their mean (0 for Z-scores)
    base_age_z = 0,
    
    # 3. Hold categorical covariates at their reference level (level 1)
    # Note: Make sure to use 'model_data_age' here to match your current dataset!
    base_race = factor(levels(model_data_age$base_race)[1], levels = levels(model_data_age$base_race)),
    base_marital = factor(levels(model_data_age$base_marital)[1], levels = levels(model_data_age$base_marital)),
    base_pckgr = factor(levels(model_data_age$base_pckgr)[1], levels = levels(model_data_age$base_pckgr)),
    
    # 4. Format labels
    Group_Label = factor(Group, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )

# Now predict() will find every variable it needs!
plot_grid$pred <- predict(model_piecewise, newdata = plot_grid, re.form = NA)

## 12. Plot the Trajectories

p_trajectory_simple <- ggplot(plot_grid, aes(x = Age_Centered, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  labs(
    title = "Model-Implied Piecewise Linear Frailty Trajectories",
    subtitle = "Centered at Age of Cancer Diagnosis (or Matched Index Age)",
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-20, 20, by = 4))

print(p_trajectory_simple)
```

```{r}

```
