---
title: "Frailty Trajectories by Cancer Burden: Age Scale & Risk-Set Matching"
author: "Analyst"
date: "`r Sys.Date()`"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 5.1_age_match_aggressive cancer.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Runs age-scale risk-set matching for aggressive cancer analyses, estimating frailty trajectories for matched cancer and control groups.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary) 
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
# Update path if necessary
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables (Age Scale)
Ensure that your raw cancer site variables (cancer_lung, cancer_pancreas, etc.) are included in the dataset here so they can be used in Step 5.

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) 
```

## 4. Extract Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Longitudinal Dataset (Stratified by Cancer Burden)

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    
    # Prevalent filter
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    
    # Define High Burden Systemic/Aggressive Cancers
    is_high_burden = if_else(
      coalesce(cancer_lung, 0) == 1 | coalesce(cancer_pancreas, 0) == 1 | 
      coalesce(cancer_stomach, 0) == 1 | coalesce(cancer_liver, 0) == 1 | 
      coalesce(cancer_brain, 0) == 1 | coalesce(cancer_lal, 0) == 1 |
      coalesce(cancer_esoph, 0) == 1 | coalesce(cancer_myeloma, 0) == 1,
      1L, 0L
    ),
    
    # Construct the 3-Level Grouping Variable
    Cancer_Burden = case_when(
      Ever_Cancer == 0 ~ "Control",
      Ever_Cancer == 1 & is_high_burden == 1 ~ "High Burden",
      Ever_Cancer == 1 & is_high_burden == 0 ~ "Low/Moderate Burden"
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  ungroup() %>%
  mutate(
    # Set Control as the reference level
    Cancer_Burden = factor(Cancer_Burden, levels = c("Control", "Low/Moderate Burden", "High Burden"))
  )
```

## 6. Check Follow-Up Structure

```{r followup-checks}
zero_fu_check <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(age_at_cycle, na.rm = TRUE),
    max_fu_yrs = max(age_at_cycle, na.rm = TRUE) - min(age_at_cycle, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu_yrs == 0)

cat("Total observations in analysis_long_age =", nrow(analysis_long_age), "\n")
cat("Patients with zero follow-up years =", nrow(zero_fu_check), "\n")
```

## 7. Incidence Density (Risk-Set) Matching (1:2 Ratio)
Assigns controls an index age based on active risk-set timing, rounding to avoid precision errors.

```{r impute-index-age}
set.seed(2026)

patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = round(first(base_age), 4),
    min_age = round(first(min_age), 4),
    max_age = round(first(max_age), 4),
    is_case = first(Ever_Cancer),
    true_age_at_cancer = round(first(true_age_at_cancer), 4),
    .groups = "drop"
  ) %>%
  mutate(
    age_bin = floor(base_age / 5) * 5 
  )

cases_summary <- patient_summary_age %>%
  filter(is_case == 1, !is.na(true_age_at_cancer), true_age_at_cancer > min_age)

controls_summary <- patient_summary_age %>%
  filter(is_case == 0)

# Target matching ratio
MATCH_RATIO <- 2 

available_controls <- controls_summary$id
assigned_times_list <- list()

for (i in seq_len(nrow(cases_summary))) {
  current_case <- cases_summary[i, ]
  case_age <- current_case$true_age_at_cancer
  
  eligible_controls <- controls_summary %>%
    filter(
      id %in% available_controls,
      age_bin == current_case$age_bin,
      min_age <= case_age,    
      max_age >= case_age     
    )
  
  if (nrow(eligible_controls) > 0) {
    n_to_sample <- min(MATCH_RATIO, nrow(eligible_controls))
    sampled_controls <- sample(eligible_controls$id, size = n_to_sample, replace = FALSE)
    
    assigned_times_list[[i]] <- data.frame(
      id = sampled_controls,
      assigned_event_age = case_age,
      matched_to_case = current_case$id
    )
    
    available_controls <- setdiff(available_controls, sampled_controls)
  }
}

controls_assigned <- bind_rows(assigned_times_list)
cases_assigned <- cases_summary %>% 
  select(id, assigned_event_age = true_age_at_cancer)

assigned_ages_df <- bind_rows(cases_assigned, controls_assigned) %>%
  select(id, assigned_event_age) %>%
  distinct(id, .keep_all = TRUE)

cat("Total Cases:", nrow(cases_summary), "\n")
cat("Total Controls Successfully Matched:", nrow(controls_assigned), "\n")
```

## 8. Generate Time Scale & Spline Variables

```{r generate-splines}
model_data_age <- analysis_long_age %>%
  inner_join(assigned_ages_df, by = "id") %>%
  mutate(
    # Center age around the event age
    Age_Centered = age_at_cycle - assigned_event_age,
    
    # Create the post-event knot
    Post = if_else(Age_Centered >= 0, 1L, 0L),
    Age_Post = if_else(Post == 1L, Age_Centered, 0),
    
    # Create Z-score of Index Age to adjust intercept in the LME model
    index_age_z = as.numeric(scale(assigned_event_age))
  )

cat("Final model_data_age row count =", nrow(model_data_age), "\n")
```

## 9. Descriptive Analysis Before LME

```{r descriptive-data}
baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- model_data_age %>% 
  group_by(id) %>%
  summarize(
    Cancer_Burden = first(Cancer_Burden),
    Ever_Cancer = first(Ever_Cancer),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    true_age_at_cancer = first(true_age_at_cancer),
    assigned_event_age = first(assigned_event_age),
    num_obs = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = first(min_age),
    max_age = first(max_age),
    max_fu_yrs = first(max_followup_yrs), 
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id") %>%
  mutate(
    Index_Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

### Table 1: Baseline Characteristics

```{r table1}
table1_age <- followup_by_patient_age %>%
  select(
    Cancer_Burden, # Using the 3-level burden factor
    base_age,
    assigned_event_age,
    fi_baseline,
    base_race,
    base_marital,
    base_living,
    base_pckgr
  ) %>%
  gtsummary::tbl_summary(
    by = Cancer_Burden,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      base_age ~ "Age at Study Entry (years)",
      assigned_event_age ~ "Age at Cancer / Assigned Index Age",
      fi_baseline ~ "Baseline Frailty Index"
    ),
    missing_text = "(Missing)"
  ) %>%
  gtsummary::add_p() %>%
  gtsummary::add_overall() %>%
  gtsummary::bold_labels()

table1_age
```

### Density Plot of Assigned Index Ages

```{r descriptive-plots, fig.width=8, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Cancer_Burden, color = Cancer_Burden)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  scale_fill_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  scale_color_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  theme_minimal() +
  labs(
    title = "Distribution of Age at Index Event by Cancer Burden",
    x = "Assigned Index Age (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_index_density_age)
```

## 10. Run the Stratified Piecewise LME Model

```{r lme-model}
model_piecewise_stratified <- lmer(
  fi_score_nocancer ~ Cancer_Burden * Age_Centered + 
                      Cancer_Burden * Age_Post + 
                      index_age_z + 
                      base_race + base_marital + base_pckgr +
                      (1 + Age_Centered | id), 
  data = model_data_age,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(model_piecewise_stratified)

# Extract fixed effects
model_results <- as.data.frame(summary(model_piecewise_stratified)$coefficients)
model_results$Term <- rownames(model_results)
rownames(model_results) <- NULL
print(model_results)
```

## 11. Generate Stratified Predictions

```{r generate-predictions}
plot_grid_strat <- expand.grid(
  Age_Centered = seq(-15, 15, by = 0.1),
  Cancer_Burden = factor(
    c("Control", "Low/Moderate Burden", "High Burden"), 
    levels = c("Control", "Low/Moderate Burden", "High Burden")
  )
) %>%
  mutate(
    Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
    index_age_z = 0, # Hold index age at the sample mean
    
    # Hold covariates at reference levels
    base_race = factor(levels(model_data_age$base_race)[1], levels = levels(model_data_age$base_race)),
    base_marital = factor(levels(model_data_age$base_marital)[1], levels = levels(model_data_age$base_marital)),
    base_pckgr = factor(levels(model_data_age$base_pckgr)[1], levels = levels(model_data_age$base_pckgr))
  )

# Predict marginal effects
plot_grid_strat$pred <- predict(model_piecewise_stratified, newdata = plot_grid_strat, re.form = NA)
```

## 12. Plot the Stratified Trajectories

```{r plot-trajectories, fig.width=8, fig.height=6}
p_trajectory_strat <- ggplot(plot_grid_strat, aes(x = Age_Centered, y = pred, color = Cancer_Burden)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  
  # 3-color palette to distinguish the groups
  scale_color_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Frailty Trajectories by Cancer Burden",
    subtitle = "Centered at Age of Cancer Diagnosis (or Matched Index Age)",
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-15, 15, by = 2), expand = expansion(mult = c(0, 0)))

print(p_trajectory_strat)
```
