---
title: "Frailty Trajectories: Prostate Cancer Split (Risk-Set Matching)"
author: "Analyst"
date: "`r Sys.Date()`"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 5.2_match_prostate cancer.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Runs risk-set matching analyses for prostate cancer subgroups, especially aggressive versus indolent prostate cancer trajectory comparisons.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary) 
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
# Update path if necessary
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables (Age Scale)

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12,
    
    # Force prostate flags to numeric to prevent coalesce errors
    cancer_newag      = as.numeric(as.character(cancer_newag)),
    cancer_prosnoa1   = as.numeric(as.character(cancer_prosnoa1)),
    cancer_status     = as.numeric(as.character(cancer_status))
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) 
```

## 4. Extract Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Dataset (Pure Prostate Cohort)

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    
    # Define Prostate Subgroups
    is_prostate_agg = if_else(coalesce(cancer_newag, 0) == 1, 1L, 0L),
    is_prostate_indolent = if_else(
      (coalesce(cancer_prosnoa1, 0) == 1 | coalesce(cancer_status, 0) == 1) & is_prostate_agg == 0, 
      1L, 0L
    ),
    is_any_prostate = if_else(is_prostate_agg == 1 | is_prostate_indolent == 1, 1L, 0L),
    
    # Identify non-prostate cancers to exclude them
    is_other_cancer = if_else(Ever_Cancer == 1 & is_any_prostate == 0, 1L, 0L)
  ) %>%
  # Filter out prevalent cases, zero follow-up, AND non-prostate cancers
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  filter(is_other_cancer == 0) %>% 
  ungroup()
```

## 6. Check Follow-Up Structure

```{r followup-checks}
cat("Total observations (Controls + Prostate Only) =", nrow(analysis_long_age), "\n")
```

## 7. Separate Risk-Set Matching (Capped Ratio)

```{r impute-index-age}
set.seed(2026)

patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = round(first(base_age), 4),
    min_age = round(first(min_age), 4),
    max_age = round(first(max_age), 4),
    is_case = first(Ever_Cancer),
    is_agg = first(is_prostate_agg),
    true_age_at_cancer = round(first(true_age_at_cancer), 4),
    .groups = "drop"
  ) %>%
  mutate(age_bin = floor(base_age / 5) * 5)

# Split into target populations
cases_agg <- patient_summary_age %>% filter(is_case == 1, is_agg == 1, true_age_at_cancer > min_age)
cases_ind <- patient_summary_age %>% filter(is_case == 1, is_agg == 0, true_age_at_cancer > min_age)
controls_pool <- patient_summary_age %>% filter(is_case == 0)

# Matching Function (Capped at 1:4 to prevent age clumping artifacts)
perform_matching <- function(cases_df, controls_df, max_ratio = 4) {
  available_controls <- controls_df$id
  assigned_times_list <- list()
  
  for (i in seq_len(nrow(cases_df))) {
    current_case <- cases_df[i, ]
    case_age <- current_case$true_age_at_cancer
    
    eligible_controls <- controls_df %>%
      filter(
        id %in% available_controls,
        age_bin == current_case$age_bin,
        min_age <= case_age,    
        max_age >= case_age      
      )
    
    if (nrow(eligible_controls) > 0) {
      n_to_sample <- min(max_ratio, nrow(eligible_controls))
      sampled_controls <- sample(eligible_controls$id, size = n_to_sample, replace = FALSE)
      
      assigned_times_list[[i]] <- data.frame(
        id = sampled_controls,
        assigned_event_age = case_age,
        matched_to_case = current_case$id
      )
      
      available_controls <- setdiff(available_controls, sampled_controls)
    }
  }
  
  controls_assigned <- bind_rows(assigned_times_list)
  cases_assigned <- cases_df %>% select(id, assigned_event_age = true_age_at_cancer)
  
  bind_rows(cases_assigned, controls_assigned) %>%
    distinct(id, .keep_all = TRUE)
}

# Run independently for both cohorts
assigned_agg_df <- perform_matching(cases_agg, controls_pool, max_ratio = 4)
assigned_ind_df <- perform_matching(cases_ind, controls_pool, max_ratio = 4)

cat("Aggressive Cohort - Cases:", nrow(cases_agg), "| Total Matched Dataset:", nrow(assigned_agg_df), "\n")
cat("Indolent Cohort - Cases:", nrow(cases_ind), "| Total Matched Dataset:", nrow(assigned_ind_df), "\n")
```

## 8. Generate Independent Analysis Datasets

```{r generate-splines}
prep_model_data <- function(assigned_df, cohort_name, case_label) {
  analysis_long_age %>%
    inner_join(assigned_df, by = "id") %>%
    mutate(
      Cohort = cohort_name,
      Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Matched Control", case_label)),
      Age_Centered = age_at_cycle - assigned_event_age,
      Post = if_else(Age_Centered >= 0, 1L, 0L),
      Age_Post = if_else(Post == 1L, Age_Centered, 0),
      index_age_z = as.numeric(scale(assigned_event_age))
    )
}

data_agg <- prep_model_data(assigned_agg_df, "Aggressive Prostate Cohort", "Aggressive Case")
data_ind <- prep_model_data(assigned_ind_df, "Indolent Prostate Cohort", "Indolent Case")

# Combine for exploratory analysis
data_combined <- bind_rows(data_agg, data_ind) %>%
  mutate(Analysis_Group = paste(Group, "(", Cohort, ")"))
```

## 9. Exploratory Analysis & Sanity Checks

```{r descriptive-data}
baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- data_combined %>% 
  group_by(id, Cohort) %>%
  summarize(
    Group = first(Group),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_pckgr = first(base_pckgr),
    assigned_event_age = first(assigned_event_age),
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id")
```

### Table 1: Baseline Characteristics

```{r table1}
table1_age <- followup_by_patient_age %>%
  select(Cohort, Group, base_age, assigned_event_age, fi_baseline, base_race, base_marital, base_pckgr) %>%
  gtsummary::tbl_strata(
    strata = Cohort,
    ~ gtsummary::tbl_summary(
        .x,
        by = Group,
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{n} ({p}%)"),
        label = list(base_age ~ "Age at Study Entry", assigned_event_age ~ "Age at Index", fi_baseline ~ "Baseline Frailty"),
        missing_text = "(Missing)"
      ) %>% gtsummary::add_p()
  )

table1_age
```

### Density Plot of Assigned Index Ages

```{r descriptive-plots, fig.width=10, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Group, color = Group)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~ Cohort) +
  scale_fill_manual(values = c("Matched Control" = "#3182bd", "Aggressive Case" = "#de2d26", "Indolent Case" = "#fd8d3c")) +
  scale_color_manual(values = c("Matched Control" = "#3182bd", "Aggressive Case" = "#de2d26", "Indolent Case" = "#fd8d3c")) +
  theme_minimal() +
  labs(title = "Distribution of Age at Index Event", subtitle = "Capped 1:4 Matching ensures controls mirror case distributions", x = "Assigned Index Age (years)", y = "Density") +
  theme(legend.position = "bottom", legend.title=element_blank())

print(p_index_density_age)
```

## 10. Run the Independent LME Models

```{r lme-model-high}
lme_formula <- fi_score_nocancer ~ Group * Age_Centered + Group * Age_Post + 
                                   index_age_z + base_race + base_marital + base_pckgr +
                                   (1 + Age_Centered | id)

model_agg <- lmer(lme_formula, data = data_agg, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
cat("\n### Model Results: Aggressive Prostate Cohort ###\n")
summary(model_agg)
```

```{r lme-model-low}
model_ind <- lmer(lme_formula, data = data_ind, control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
cat("\n### Model Results: Indolent Prostate Cohort ###\n")
summary(model_ind)
```

## 11. Generate Stratified Predictions

```{r generate-predictions}
create_pred_grid <- function(model, data_subset) {
  grid <- expand.grid(
    Age_Centered = seq(-15, 15, by = 0.1),
    Group = factor(levels(data_subset$Group), levels = levels(data_subset$Group))
  ) %>%
    mutate(
      Cohort = data_subset$Cohort[1],
      Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
      index_age_z = 0, # Hold at sample mean
      base_race = factor(levels(data_subset$base_race)[1], levels = levels(data_subset$base_race)),
      base_marital = factor(levels(data_subset$base_marital)[1], levels = levels(data_subset$base_marital)),
      base_pckgr = factor(levels(data_subset$base_pckgr)[1], levels = levels(data_subset$base_pckgr))
    )
  
  grid$pred <- predict(model, newdata = grid, re.form = NA)
  return(grid)
}

plot_grid_agg <- create_pred_grid(model_agg, data_agg)
plot_grid_ind <- create_pred_grid(model_ind, data_ind)

plot_grid_combined <- bind_rows(plot_grid_agg, plot_grid_ind) %>%
  mutate(Line_Type = if_else(grepl("Control", Group), "Matched Control", as.character(Group)))
```

## 12. Plot the Separate Trajectories

```{r plot-trajectories, fig.width=10, fig.height=5}
p_trajectory_faceted <- ggplot(plot_grid_combined, aes(x = Age_Centered, y = pred, color = Line_Type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  facet_wrap(~ Cohort) + 
  scale_color_manual(values = c(
    "Matched Control" = "#3182bd", 
    "Aggressive Case" = "#de2d26",
    "Indolent Case" = "#fd8d3c"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Frailty Trajectories: Aggressive vs. Indolent Prostate Cancer",
    subtitle = "Matched cohort design restricting to prostate cases and healthy controls",
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-15, 15, by = 5), expand = expansion(mult = c(0, 0)))

print(p_trajectory_faceted)
```

\`\`\`
