---
title: "IPCW-GEE Frailty Trajectories"
author: "Nemo Zhou"
date: "2026-07-28"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

<!--
Project: Frailty Trajectories Before and After Incident Cancer in the
         Health Professionals Follow-up Study
Script:  5.3_ipcw_gee_trajectory_visualization.Rmd
Author:  Nemo Zhou
Date started:      2026-07-28
Date last updated: 2026-07-28

Purpose:
  Read and visualize saved outputs from the current 5.0-5.2 IPCW-GEE
  sensitivity workflow without rebuilding matching, constructing weights, or
  fitting any model. The report displays separate unweighted-GEE and
  IPCW-weighted-GEE plots, with cancer-case and control trajectories overlaid
  within each cohort, and an IPCW-weighted Table One.

  The High Burden Cohort is displayed only through +12 years because later
  support is not retained for that cohort. Death is terminal and is not
  treated as IPCW censoring by the 5.x engine.

Inputs:
  Results/cancer/data/5.1_overall_m2_gee_predicted_trajectories.csv
  Results/cancer/data/5.2_low_moderate_burden_m2_gee_predicted_trajectories.csv
  Results/cancer/data/5.2_high_burden_m2_gee_predicted_trajectories.csv
  Results/cancer/data/5.2_smoking_m2_gee_predicted_trajectories.csv
  Results/cancer/data/5.2_obesity_m2_gee_predicted_trajectories.csv
  Data/*_long_ipcw_m2_gee.rds

Output:
  Results/cancer/visuals/5.3_ipcw_gee_trajectory_visualization.html

No persistent PNG files are saved. This report does not run matching, GLME,
GEE, or IPCW weight construction.
-->

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(readr)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
results_dir <- file.path(project_dir, "Results", "cancer", "data")
data_dir <- file.path(project_dir, "Data")

group_levels <- c("Control", "Cancer Case")
model_levels <- c("Unweighted GEE", "IPCW-weighted GEE")
group_colors <- c("Control" = "#3182bd", "Cancer Case" = "#de2d26")

read_required_csv <- function(path) {
  if (!file.exists(path)) {
    stop("Required 5.x GEE trajectory artifact is missing: ", path,
         ". Run 5.1 and 5.2 before rendering this report.", call. = FALSE)
  }
  readr::read_csv(path, show_col_types = FALSE)
}

read_prediction <- function(path, analysis_label, cohort_filter = NULL) {
  out <- read_required_csv(path)
  if (!is.null(cohort_filter)) out <- out %>% filter(Cohort == cohort_filter)
  out %>%
    transmute(
      Analysis = analysis_label,
      Cohort,
      Group = factor(Group, levels = group_levels),
      Model = factor(Model, levels = model_levels),
      Age_Centered,
      pred,
      se,
      lwr,
      upr,
      vcov_type
    )
}

read_gee_panel <- function(path, analysis_label, cohort_filter = NULL) {
  if (!file.exists(path)) {
    stop("Required 5.x IPCW-GEE ledger is missing: ", path,
         ". Run 5.1 or 5.2 before rendering this report.", call. = FALSE)
  }
  out <- readRDS(path)
  if (!is.null(cohort_filter)) out <- out %>% filter(Cohort == cohort_filter)
  out %>%
    filter(
      retained_gee_row,
      Age_Centered >= 0,
      Age_Centered <= 16,
      Group %in% group_levels,
      is.finite(fi_score_nocancer),
      is.finite(sw_ipcw),
      sw_ipcw > 0
    ) %>%
    transmute(
      Analysis = analysis_label,
      Cohort,
      Group = factor(Group, levels = group_levels),
      trajectory_id,
      sw_ipcw,
      index_age,
      base_race,
      base_marital,
      base_living,
      base_pckgr,
      base_calor,
      base_sat,
      base_diet_chol,
      base_alco
    )
}

weighted_mean <- function(x, w) {
  keep <- is.finite(x) & is.finite(w) & w > 0
  if (!any(keep)) return(NA_real_)
  sum(x[keep] * w[keep]) / sum(w[keep])
}

weighted_sd <- function(x, w) {
  keep <- is.finite(x) & is.finite(w) & w > 0
  if (!any(keep)) return(NA_real_)
  m <- weighted_mean(x[keep], w[keep])
  sqrt(sum(w[keep] * (x[keep] - m)^2) / sum(w[keep]))
}

format_mean_sd <- function(m, s) {
  if (!is.finite(m)) return("Missing")
  sprintf("%.2f (%.2f)", m, s)
}

format_percent <- function(p) {
  if (!is.finite(p)) return("Missing")
  sprintf("%.1f%%", 100 * p)
}

smd <- function(mean_case, sd_case, mean_control, sd_control) {
  denominator <- sqrt((sd_case^2 + sd_control^2) / 2)
  if (!is.finite(denominator) || denominator == 0) return(NA_real_)
  (mean_case - mean_control) / denominator
}

build_table_one <- function(data) {
  continuous <- c(
    index_age = "Index age, years",
    base_calor = "Calories, kcal/day",
    base_sat = "Saturated fat, g/day",
    base_diet_chol = "Dietary cholesterol, mg/day",
    base_alco = "Alcohol, g/day"
  )
  categorical <- c(
    base_race = "Race",
    base_marital = "Marital status",
    base_living = "Living arrangement",
    base_pckgr = "Pack-year group"
  )

  pieces <- list()
  k <- 1L
  for (d in split(data, interaction(data$Analysis, data$Cohort, drop = TRUE))) {
    analysis_label <- unique(d$Analysis)
    cohort_label <- unique(d$Cohort)
    control <- d[d$Group == "Control", , drop = FALSE]
    case <- d[d$Group == "Cancer Case", , drop = FALSE]

    for (v in names(continuous)) {
      cu <- c(mean = mean(control[[v]], na.rm = TRUE), sd = sd(control[[v]], na.rm = TRUE))
      xu <- c(mean = mean(case[[v]], na.rm = TRUE), sd = sd(case[[v]], na.rm = TRUE))
      cw <- c(mean = weighted_mean(control[[v]], control$sw_ipcw), sd = weighted_sd(control[[v]], control$sw_ipcw))
      xw <- c(mean = weighted_mean(case[[v]], case$sw_ipcw), sd = weighted_sd(case[[v]], case$sw_ipcw))
      pieces[[k]] <- data.frame(
        Analysis = analysis_label, Cohort = cohort_label,
        Variable = unname(continuous[[v]]), Level_or_statistic = "Mean (SD)",
        Control_unweighted = format_mean_sd(cu[["mean"]], cu[["sd"]]),
        Cancer_case_unweighted = format_mean_sd(xu[["mean"]], xu[["sd"]]),
        Control_IPCW_weighted = format_mean_sd(cw[["mean"]], cw[["sd"]]),
        Cancer_case_IPCW_weighted = format_mean_sd(xw[["mean"]], xw[["sd"]]),
        SMD_unweighted = smd(xu[["mean"]], xu[["sd"]], cu[["mean"]], cu[["sd"]]),
        SMD_IPCW_weighted = smd(xw[["mean"]], xw[["sd"]], cw[["mean"]], cw[["sd"]])
      )
      k <- k + 1L
    }

    for (v in names(categorical)) {
      control_x <- as.character(control[[v]])
      case_x <- as.character(case[[v]])
      control_x[is.na(control_x)] <- "Missing"
      case_x[is.na(case_x)] <- "Missing"
      for (level in sort(unique(c(control_x, case_x)))) {
        cp <- mean(control_x == level)
        xp <- mean(case_x == level)
        cw <- weighted_mean(as.numeric(control_x == level), control$sw_ipcw)
        xw <- weighted_mean(as.numeric(case_x == level), case$sw_ipcw)
        pieces[[k]] <- data.frame(
          Analysis = analysis_label, Cohort = cohort_label,
          Variable = unname(categorical[[v]]), Level_or_statistic = level,
          Control_unweighted = format_percent(cp),
          Cancer_case_unweighted = format_percent(xp),
          Control_IPCW_weighted = format_percent(cw),
          Cancer_case_IPCW_weighted = format_percent(xw),
          SMD_unweighted = smd(xp, sqrt(xp * (1 - xp)), cp, sqrt(cp * (1 - cp))),
          SMD_IPCW_weighted = smd(xw, sqrt(xw * (1 - xw)), cw, sqrt(cw * (1 - cw)))
        )
        k <- k + 1L
      }
    }
  }
  bind_rows(pieces) %>% mutate(across(starts_with("SMD"), ~ round(.x, 3)))
}
```

# Analysis scope

This report uses the current 5.x IPCW-GEE outputs. The observation probabilities were estimated in the complete full-cohort response ledger and transported to the matched assignments before the weighted GEE was fitted. The two trajectory plots are separate: one is the unweighted GEE and one is the IPCW-weighted GEE. Within each plot, cancer cases and controls use the same model and are overlaid directly. No cancer-minus-control difference plot or weighted-minus-unweighted change plot is produced.

# IPCW-weighted Table One

The Table One describes observed retained post-index person-time from 0 through +16 years. The weighted columns use the saved stabilized `sw_ipcw`; this is an observation-process diagnostic, not a treatment-balance table. Death is not converted into a censoring event.

```{r table-one}
gee_panel <- bind_rows(
  read_gee_panel(
    file.path(data_dir, "riskset_matched_overall_long_ipcw_m2_gee.rds"),
    "Overall Cancer"
  ),
  read_gee_panel(
    file.path(data_dir, "riskset_matched_low_moderate_burden_long_ipcw_m2_gee.rds"),
    "Cancer Burden", "Low/Moderate Burden Cohort"
  ),
  read_gee_panel(
    file.path(data_dir, "riskset_matched_high_burden_long_ipcw_m2_gee.rds"),
    "Cancer Burden", "High Burden Cohort"
  ),
  read_gee_panel(
    file.path(data_dir, "riskset_matched_smoking_long_ipcw_m2_gee.rds"),
    "Smoking-Related Cancer"
  ),
  read_gee_panel(
    file.path(data_dir, "riskset_matched_obesity_long_ipcw_m2_gee.rds"),
    "Obesity-Related Cancer"
  )
)

table_one_counts <- gee_panel %>%
  group_by(Analysis, Cohort, Group) %>%
  summarize(
    observed_person_visits = n(),
    trajectory_assignments = n_distinct(trajectory_id),
    sum_IPCW_weights = sum(sw_ipcw),
    .groups = "drop"
  )

knitr::kable(
  table_one_counts,
  digits = 1,
  caption = "Observed retained person-time contributing to the IPCW-GEE Table One"
)

knitr::kable(
  build_table_one(gee_panel),
  digits = 3,
  caption = "IPCW-GEE Table One: unweighted and IPCW-weighted observed post-index person-time"
)
```

# Load GEE trajectories

```{r trajectories}
gee_trajectories <- bind_rows(
  read_prediction(
    file.path(results_dir, "5.1_overall_m2_gee_predicted_trajectories.csv"),
    "Overall Cancer"
  ),
  read_prediction(
    file.path(results_dir, "5.2_low_moderate_burden_m2_gee_predicted_trajectories.csv"),
    "Cancer Burden", "Low/Moderate Burden Cohort"
  ),
  read_prediction(
    file.path(results_dir, "5.2_high_burden_m2_gee_predicted_trajectories.csv"),
    "Cancer Burden", "High Burden Cohort"
  ),
  read_prediction(
    file.path(results_dir, "5.2_smoking_m2_gee_predicted_trajectories.csv"),
    "Smoking-Related Cancer"
  ),
  read_prediction(
    file.path(results_dir, "5.2_obesity_m2_gee_predicted_trajectories.csv"),
    "Obesity-Related Cancer"
  )
) %>%
  filter(
    Age_Centered >= -16,
    Age_Centered <= 16,
    !(Cohort == "High Burden Cohort" & Age_Centered > 12)
  )

trajectory_coverage <- gee_trajectories %>%
  group_by(Analysis, Cohort, Group, Model) %>%
  summarize(
    n_prediction_rows = n(),
    min_time = min(Age_Centered),
    max_time = max(Age_Centered),
    vcov_type = paste(unique(vcov_type), collapse = "; "),
    .groups = "drop"
  )

knitr::kable(
  trajectory_coverage,
  digits = 2,
  caption = "Saved 5.x GEE trajectory-grid coverage"
)
```

# Separate case/control trajectory plots

```{r trajectory-plots, fig.width=15, fig.height=10}
make_gee_plot <- function(data, selected_model) {
  ggplot(
    data %>% filter(Model == selected_model),
    aes(x = Age_Centered, y = pred, color = Group, fill = Group)
  ) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.10, color = NA) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey35") +
    facet_wrap(~ Cohort, scales = "free_y") +
    scale_color_manual(values = group_colors) +
    scale_fill_manual(values = group_colors) +
    scale_x_continuous(breaks = seq(-16, 16, by = 4)) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(size = 8)
    ) +
    labs(
      title = "Cancer-case and control frailty trajectories",
      subtitle = paste0(
        "Current 5.x model: ", selected_model,
        ". Ribbons are GEE empirical-sandwich 95% confidence intervals."
      ),
      x = "Years relative to assigned index",
      y = "Predicted frailty index",
      color = NULL,
      fill = NULL
    )
}

unweighted_gee_plot <- make_gee_plot(gee_trajectories, model_levels[[1]])
ipcw_weighted_gee_plot <- make_gee_plot(gee_trajectories, model_levels[[2]])

unweighted_gee_plot
ipcw_weighted_gee_plot
```

# Interpretation notes

* Each plot compares cancer cases and controls within the same GEE model.
* The 5.x trajectories use the M2 reference profile from the shared engine: factor covariates are held at the most common observed level and continuous covariates at their observed means, with `index_age_z = 0`.
* Confidence intervals use the GEE empirical sandwich covariance.
* This report reads saved outputs only and does not run matching, IPCW, GEE, or GLME models.
