---
title: "Overall cancer joint-model mortality sensitivity"
author: "Nemo Zhou"
output:
  html_document:
    toc: true
    theme: readable
params:
  run_dir: null
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS/Results/cancer/visuals") })
---

```{r setup, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the
#          Health Professionals Follow-up Study
# Author: Nemo Zhou
# Date started: 2026-09-20
# Date last updated: 2026-09-29 (M2 specification and explicit date-policy audit)
# Purpose: Render a selected 8.1 run's preparation, survival, model comparison,
# and chain diagnostics without fitting any model. Requires an explicit run_dir
# (or JM_RUN_DIR); never selects a stale run implicitly. HTML belongs exclusively
# in Results/cancer/visuals. Inline plots use embedded temporary graphics;
# no persistent PNG or separate figure folder is retained.
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, dev="svglite", fig.width=10, fig.height=5)
run_dir <- params$run_dir
if (is.null(run_dir) || !nzchar(run_dir)) run_dir <- Sys.getenv("JM_RUN_DIR", "")
if (!nzchar(run_dir) || !file.exists(file.path(run_dir,"run_status.rds"))) stop("Set params$run_dir or JM_RUN_DIR to an explicit 8.1 run directory.")
run_dir <- normalizePath(run_dir)
project_dir <- "/Users/nemo/Library/CloudStorage/OneDrive-HarvardUniversity/Research/Frailty HPFS"
visuals_dir <- normalizePath(file.path(project_dir,"Results","cancer","visuals"),mustWork=FALSE)
knitr::opts_knit$set(root.dir=project_dir)
read_csv <- function(name) {
  p <- file.path(run_dir,name)
  if (file.exists(p)) utils::read.csv(p,check.names=FALSE) else NULL
}
read_rds <- function(name) {
  p <- file.path(run_dir,name)
  if (file.exists(p)) readRDS(p) else NULL
}
show_table <- function(x) {
  if (is.null(x) || !nrow(x)) cat("Not available for this run.\n\n") else print(knitr::kable(x,digits=4))
}
status <- read_rds("run_status.rds")
config <- read_rds("configuration.rds")
alignment <- if(!is.null(config$alignment_model_id)) config$alignment_model_id else "LEGACY / M2 alignment not established"
synthetic <- !is.null(config$input_hashes) && any(grepl("synthetic",config$input_hashes,ignore.case=TRUE))
report_m2 <- identical(alignment,"M2_full_spline")
library(ggplot2)
```

## Run status and interpretation

```{r status, results='asis'}
cat("**Profile:** ",status$profile,"  \n**Status:** ",status$status,"  \n",sep="")
cat("**Run directory:** `",run_dir,"`\n\n",sep="")
cat("**Alignment:** ",alignment,"\n\n",sep="")
if(synthetic) cat("**SYNTHETIC SOFTWARE QA: these values are test fixtures, not HPFS estimates.**\n\n")
if(identical(status$status,"blocked_date_policy")) cat("**FITTING BLOCKED — unresolved month-level date policy. No component or joint model was fitted. Source cases and dates were preserved.**\n\n")
if(!report_m2) cat("**Legacy run:** this report does not establish M2 alignment. The M2 specification below describes the revised workflow, not this historical fit.\n\n")
if (!is.null(status$error)) cat("**Failure:** ",status$error,"\n\n",sep="")
if(status$profile=="prepare") cat("**Preparation only: no models were fitted.**\n\n")
if(status$profile=="pilot") cat("**NON-INFERENTIAL PILOT.** All cases plus a seeded 25% sample of retained controls; component and joint comparisons use this same sample.\n\n")
if(status$profile=="full") cat("Numerical diagnostic success requires visual review of the trace plots below before inference. A returned model alone does not establish convergence.\n\n")
```

## Revised M2 specification

The longitudinal and mortality components include index age, baseline race,
marital status, living arrangement, categorical pack-years, calories, saturated
fat, dietary cholesterol, and alcohol. The longitudinal fixed effects use Group
by df-3 natural spline of years since index. Canonical M2 spline and dietary
scaling is applied once. Random effects use participant intercepts and attained-age
slopes, `(index_age + Age_Centered - 60)/4`, evaluated dynamically at hazard times.

Current-value and current-value-plus-slope associations are prespecified.
The joint baseline hazard uses quadratic B-splines with nine segments, identity
time, and second-order smoothing. The Cox component initializes the joint model;
it does not imply an unspecified Cox baseline in the final joint likelihood.

This is a secondary analysis of one selected matched assignment per participant.
Case assignments take priority; otherwise the earliest control assignment is
retained. Selection occurs before FI/covariate exclusions. This changes the
original matched population and does not preserve its balance automatically.

Index is time zero. Pre-index FI contributes history; mortality follow-up starts
at index conditional on being alive then, and ends at death, own cancer for
controls, or the fixed 2020 administrative endpoint. Cancer-date ties censor
controls. Questionnaire nonresponse is not treated as end of mortality
ascertainment. Censoring at own cancer may be informative.

Reported curves are fixed-effect FI trajectories at the common reference
covariates, with random effects set to zero. They are not dynamically
survivor-standardized averages. Joint-versus-longitudinal differences are
model-based sensitivity differences, not identified mortality-bias corrections.
The analysis neither defines FI after death nor reverses selection from survival
to index. Residual matched-set dependence and the selected cohort remain limitations.

## Date-policy decision and audit

The agreed policy is to keep fitting blocked while resolving date precision.
Diagnosis and death fields are recorded in months since 1900. Readable dates
assign day 15; this is a convention rather than an observed day. A same-month
record is not evidence of missing dates or death on the diagnosis day.
Exact-day availability on the HPFS server has not been verified. No case is
excluded and no date is shifted to make survival duration positive.

Counts below cover selected assignments **before FI or M2 covariate exclusions**.
Issue counts can overlap; the overlap table makes the two principal categories
explicit. Absence of a death date is not an error when the participant is present
in the death-source panel; conflicting recorded death dates are a separate issue.

```{r date_audit, results='asis'}
show_table(read_csv("date_audit_summary.csv"))
show_table(read_csv("date_issue_overlap.csv"))
show_table(read_csv("M2_covariate_availability.csv"))
```

## Preparation and cohort changes

```{r filters, results='asis'}
show_table(read_csv("filter_log.csv"))
show_table(read_csv("source_hash_mismatch.csv"))
show_table(read_csv("survival_summary.csv"))
show_table(read_csv("clock_basis_contract.csv"))
```

```{r balance, results='asis'}
balance <- read_csv("cohort_balance.csv")
show_table(balance)
if(!is.null(balance) && any(abs(balance$smd[balance$population=="JM_complete_case"])>0.1,na.rm=TRUE)) cat("**Balance warning:** selected-cohort absolute SMD exceeds 0.1 for one or more characteristics. Inspect overlap and adjustment before interpreting arm contrasts.\n\n")
```

## Relative-time support and mortality extrapolation

```{r support, results='asis'}
show_table(read_csv("support_by_time.csv"))
z <- read_csv("mortality_extrapolation.csv")
if(!is.null(z)) {
  zz <- aggregate(z[c("years_after_last_FI","years_before_first_FI","beyond_canonical_boundary","beyond_cohort_FI")],list(Group=z$Group),mean)
  show_table(zz)
  cat("Values above are mean years per participant. Positive values identify survival follow-up requiring FI extrapolation; inspect the saved participant-level audit before interpretation.\n\n")
}
```

## Component models and survival

```{r components, results='asis'}
show_table(read_csv("component_attempts.csv"))
components <- read_rds("components.rds")
if(!is.null(components)) {
  cat("**Accepted random effects:** ",components$random_structure,"\n\n",sep="")
  cat("**Longitudinal comparator covariance:** ",components$covariance,"\n\n",sep="")
  if(!is.null(components$zph$table)) show_table(as.data.frame(components$zph$table)) else cat("Cox PH diagnostics unavailable.\n\n")
}
```

```{r survival}
if(!is.null(components$km) && length(components$km$time)) {
  km <- components$km
  z <- data.frame(time=km$time,survival=km$surv,Group=rep(names(km$strata),km$strata))
  print(ggplot(z,aes(time,survival,color=Group))+geom_step()+theme_minimal()+labs(x="Years since index",y="Estimated survival",title="Descriptive Kaplan–Meier curves in the selected run cohort"))
} else {
  cat("Kaplan–Meier output is unavailable for this run.\n")
}
```

## Longitudinal-only and joint-posterior comparisons

```{r comparisons, results='asis'}
longitudinal <- read_rds("longitudinal_only_summaries.rds")
if(!is.null(longitudinal)) show_table(longitudinal$theta)
all_curves <- list()
if(!is.null(longitudinal) && nrow(longitudinal$curves)) {
  z <- longitudinal$curves; z$model <- "Longitudinal only"; all_curves[[1]] <- z
}
for(nm in c("value","value_slope")) {
  st <- read_rds(file.path(nm,"status.rds"))
  cat("\n\n### Association: ",nm,"\n\n",sep="")
  if(is.null(st)) {cat("Not fitted.\n\n"); next}
  cat("**Status:** ",st$status,"\n\n",sep="")
  if(!is.null(st$error)) cat(st$error,"\n\n")
  sums <- read_rds(file.path(nm,"posterior_summaries.rds"))
  if(is.null(sums)) {cat("No posterior inference released. Inspect fit/diagnostic status.\n\n"); next}
  cat("**Interpretation status:** ",sums$inference_status,"\n\n",sep="")
  show_table(sums$theta); show_table(sums$associations); show_table(sums$group_mortality_hr)
  cat("Marginal likelihood criteria below compare eligible models within this run; no automatic model selection is performed.\n\n")
  show_table(read_csv(file.path(nm,"marginal_fit_statistics.csv")))
  if(nrow(sums$curves)) {z <- sums$curves; z$model <- paste("Joint",nm); all_curves[[length(all_curves)+1]] <- z}
}
curves <- dplyr::bind_rows(all_curves)
if(nrow(curves)) print(ggplot(curves,aes(time,estimate,color=model,fill=model))+
  geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.12,color=NA)+geom_line()+
  geom_vline(xintercept=0,linetype="dashed")+facet_wrap(~quantity,scales="free_y")+
  theme_minimal()+labs(x="Years relative to index",y="FI (derivative: FI/year)",
  title="Same-cohort component and joint-model comparisons",
  subtitle=if(status$profile=="pilot") "NON-INFERENTIAL PILOT" else "Pointwise intervals; review diagnostics and cohort balance"))
```

## Chain diagnostics and trace review

All monitored free regression, association, residual-scale, covariance,
baseline-hazard, and smoothing parameters require R-hat <1.05 and bulk/tail
ESS >400 for full-run numerical eligibility. Structural covariance zeros are
verified separately. Numerical eligibility still requires human trace review.

```{r traces, results='asis', fig.height=9}
for(nm in c("value","value_slope")) {
  cat("\n\n### ",nm,"\n\n",sep="")
  show_table(read_csv(file.path(nm,"mcmc_diagnostics.csv")))
  show_table(read_csv(file.path(nm,"constrained_parameters.csv")))
  trace <- read_rds(file.path(nm,"trace_data.rds"))
  if(is.null(trace)) next
  trace$label <- paste(trace$block,trace$parameter,sep=": ")
  chunks <- split(unique(trace$label),ceiling(seq_along(unique(trace$label))/6))
  for(labels in chunks) print(ggplot(trace[trace$label %in% labels,],aes(iteration,value,color=factor(chain)))+
    geom_line(alpha=0.65,linewidth=0.25)+facet_wrap(~label,scales="free_y",ncol=2)+theme_minimal()+
    labs(color="Chain",title=paste("Trace review:",nm),x="Retained draw within chain"))
}
```

## Provenance

```{r provenance}
if(!is.null(config)) print(config[c("schema_version","alignment_model_id","profile","seed","admin_date","date_policy","covariates","random_formula","canonical_scaling","baseline_hazard","prediction_window","mcmc","input_hashes","script_hashes","priors","software_versions")])
```
