---
title: "Frailty Index Pre/Post Cancer Trajectory Exploratory Data Analysis"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 1.1_EDA.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Performs exploratory data analysis for frailty trajectories before and after cancer, including data preparation, descriptive summaries, and initial visualizations.
```

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Required Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(gtsummary)
library(lubridate)
```

# 1. Data Preparation

```{r data_prep}
# Load the Imputed Data
fi_long <- readRDS("/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")

# 1. Prepare continuous time variables using exact months since Jan 1900
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    # Ensure variables are numeric for math
    worked_rtmnyr = as.numeric(worked_rtmnyr),
    cancer_dateca = as.numeric(cancer_dateca),
    
    # Calculate exact age at this cycle
    age_at_cycle = (worked_rtmnyr - as.numeric(dbmy09)) / 12,
    
    # Create readable Year for cross-sectional plots
    Year = case_when(
      as.numeric(cycle) > 50 ~ as.numeric(paste0("19", cycle)),
      TRUE ~ as.numeric(paste0("20", cycle))
    )
  )

# 2. Identify Baseline vs Incident Cancer 
# Group by ID to establish the participant's specific entry timeline
participant_timelines <- fi_prep %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  summarise(
    First_Survey_Month = first(worked_rtmnyr),
    Last_Survey_Month = max(worked_rtmnyr),
    Cancer_Month = first(cancer_dateca),
    Has_Cancer = first(cancer_cancer) == 1,
    .groups = "drop"
  ) %>%
  mutate(
    # Baseline Cancer: Cancer occurred BEFORE or DURING the month of their first survey
    Is_Baseline_Cancer = Has_Cancer & (!is.na(Cancer_Month) & Cancer_Month <= First_Survey_Month),
    
    # Incident Cancer: Cancer occurred AFTER their first survey
    Is_Incident_Cancer = Has_Cancer & (!is.na(Cancer_Month) & Cancer_Month > First_Survey_Month),
    
    # Plot Group
    Plot_Group = case_when(
      Is_Incident_Cancer ~ "Incident Cancer Sub-cohort",
      !Has_Cancer ~ "Censored/No-Cancer Population",
      TRUE ~ NA_character_ # Baseline cases will be NA and filtered out
    )
  )

# 3. Apply Exclusions and merge cohort assignments back to main data
fi_clean <- fi_prep %>%
  inner_join(participant_timelines %>% select(id, Cancer_Month, Last_Survey_Month, Plot_Group), by = "id") %>%
  filter(!is.na(Plot_Group)) %>% # This safely drops Baseline Cancer cases
  mutate(
    # Force FI to NA if they answered fewer than 20 non-cancer items
    fi_score_nocancer = if_else(n_answered_nocancer < 20, NA_real_, fi_score_nocancer),
    frailty_cat_nocancer = if_else(is.na(fi_score_nocancer), NA_character_, frailty_cat_nocancer)
  ) %>%
  mutate(Plot_Group = factor(Plot_Group, levels = c("Incident Cancer Sub-cohort", "Censored/No-Cancer Population")))

# Filter for valid FI scores only for active plotting
fi_valid <- fi_clean %>% filter(participated == 1, !is.na(fi_score_nocancer))
```

------------------------------------------------------------------------

# 2. Baseline Characteristics (Table 1)

```{r baseline_table}
baseline_data <- fi_valid %>%
  group_by(id) %>%
  arrange(worked_rtmnyr) %>%
  slice(1) %>% 
  ungroup() %>%
  select(Plot_Group, frailty_cat_nocancer, fi_score_nocancer, age_at_cycle) 

baseline_data %>%
  tbl_summary(
    by = Plot_Group,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "no"
  ) %>%
  add_p() %>%
  modify_header(label = "**Baseline Characteristic**") %>%
  bold_labels()
```

------------------------------------------------------------------------

# 3. Data Quality & Distribution Over Calendar Year

```{r data_quality}
attrition_data <- fi_clean %>%
  filter(participated == 1) %>%
  group_by(Plot_Group, Year) %>%
  summarise(
    N_Total = n(),
    N_Valid_FI = sum(!is.na(fi_score_nocancer)),
    .groups = "drop"
  ) %>%
  mutate(Percent_Active = (N_Valid_FI / N_Total) * 100,
         Year = as.factor(Year))

ggplot(attrition_data, aes(x = Year, y = Percent_Active, fill = Plot_Group)) +
  geom_bar(stat = "identity", position = position_dodge(width=0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Active)), 
            position = position_dodge(width=0.8), vjust = -0.5, fontface = "bold", size = 3) +
  scale_y_continuous(limits = c(0, 115), labels = scales::label_percent(scale = 1)) +
  scale_fill_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", "Incident Cancer Sub-cohort" = "#E15759")) +
  facet_wrap(~ Plot_Group, ncol = 1) +
  theme_minimal() +
  labs(
    title = "Data Usability Rate per Cycle (Non-Cancer FI)",
    subtitle = "Percentage of active participants with >= 20 valid items.",
    x = "Study Year", y = "% Usable Data", fill = "Cohort"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 10),
        strip.text = element_text(face="bold", size=12),
        legend.position = "top")
```

------------------------------------------------------------------------

# 4. Event-Centered Trajectories (Backward Timescale)

```{r setup_event_centered}
# Calculate Exact Continuous Time (No Binning)
fi_event_all <- fi_valid %>%
  mutate(
    # Continuous Time to Event (Exact Years)
    Time_to_Event_Exact = case_when(
      Plot_Group == "Incident Cancer Sub-cohort" ~ (worked_rtmnyr - Cancer_Month) / 12,
      Plot_Group == "Censored/No-Cancer Population" ~ (worked_rtmnyr - Last_Survey_Month) / 12
    )
  )

# Filter for the specific time window you want to visualize (-16 to +12 years)
fi_plot_data <- fi_event_all %>%
  filter(Time_to_Event_Exact >= -16 & Time_to_Event_Exact <= 12)

# Extract just one row per participant who got incident cancer
diagnosis_data <- participant_timelines %>%
  filter(Is_Incident_Cancer == TRUE) %>%
  mutate(
    Year_of_Diagnosis = 1900 + floor((Cancer_Month - 1) / 12)
  )

total_incident_n <- nrow(diagnosis_data)

ggplot(diagnosis_data, aes(x = Year_of_Diagnosis)) +
  geom_histogram(binwidth = 1, fill = "#E15759", color = "white", alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Distribution of Incident Cancer Diagnosis Timeline",
    subtitle = paste0("Total Incident Cancer Sample Size: N = ", total_incident_n),
    x = "Year of Diagnosis",
    y = "Number of Participants"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 11))
```

## Mean Trajectory: Cancer vs. Control

```{r mean_trajectory_comparison}
ggplot(fi_plot_data, aes(x = Time_to_Event_Exact, y = fi_score_nocancer, 
                         color = Plot_Group, fill = Plot_Group)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  
  # geom_smooth creates a smooth curve and shaded 95% CI area automatically
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), linewidth = 1.2, alpha = 0.2) +
  
  scale_x_continuous(breaks = seq(-16, 12, by = 4)) +
  scale_color_manual(values = c("Incident Cancer Sub-cohort" = "#E15759")) +
  scale_fill_manual(values = c("Incident Cancer Sub-cohort" = "#E15759")) +
  theme_minimal() +
  labs(
    title = "Continuous Smooth Trajectory: Incident Cancer",
    subtitle = "Time 0 = Cancer Diagnosis OR Last Follow-up. Shaded areas represent 95% CIs.",
    x = "Exact Years Relative to Event (Diagnosis)",
    y = "Estimated Non-Cancer Frailty Index Score",
    color = "Cohort",
    fill = "Cohort"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 11), legend.position = "bottom")
```

------------------------------------------------------------------------

# 5. Cancer Sub-Type Analysis

```{r cancer_subtype}
# Map exact cancer indicators to a single Primary Site variable
# Note: Using 'cancer_prosnoa1' (or 'cancer_newag') as the proxy for Prostate since the basic generic 'prostate' col was removed in 2020.
fi_subtype <- fi_event_all %>%
  filter(Plot_Group == "Incident Cancer Sub-cohort") %>%
  mutate(
    primary_cancer_site = case_when(
      cancer_prosnoa1 == 1 | cancer_newag == 1 ~ "Prostate",
      cancer_colorect == 1                     ~ "Colorectal",
      cancer_mel == 1                          ~ "Melanoma",
      cancer_lung == 1                         ~ "Lung",
      TRUE                                     ~ "Other/Mixed"
    )
  )

top_cancers <- c("Prostate", "Colorectal", "Melanoma", "Lung")

# Filter data for smooth plotting
subtype_plot_data <- fi_subtype %>%
  filter(primary_cancer_site %in% top_cancers) %>%
  filter(Time_to_Event_Exact >= -16 & Time_to_Event_Exact <= 12)

ggplot(subtype_plot_data, aes(x = Time_to_Event_Exact, y = fi_score_nocancer, 
                              color = primary_cancer_site, fill = primary_cancer_site)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  
  # Smooth lines for subtypes
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), linewidth = 1, alpha = 0.15) +
  
  scale_x_continuous(breaks = seq(-16, 12, by = 4)) +
  facet_wrap(~ primary_cancer_site) +
  theme_minimal() +
  labs(
    title = "Continuous Frailty Trajectory by Incident Cancer Types",
    subtitle = "Shaded areas represent 95% CIs.",
    x = "Exact Years Relative to Diagnosis",
    y = "Estimated Non-Cancer FI"
  ) +
  theme(legend.position = "none")
```

```{r eda_spaghetti_plots, fig.width=12, fig.height=6}
# 1. Randomly sample 100 IDs from each cohort to prevent overplotting
set.seed(123) # For reproducibility
sampled_ids <- fi_event_all %>%
  group_by(Plot_Group) %>%
  distinct(id) %>%
  sample_n(size = 100, replace = FALSE) %>%
  pull(id)

# Filter the data for only these individuals
spaghetti_data <- fi_event_all %>%
  filter(id %in% sampled_ids) %>%
  filter(Time_to_Event_Exact >= -12 & Time_to_Event_Exact <= 12)

# Plot
ggplot(spaghetti_data, aes(x = Time_to_Event_Exact, y = fi_score_nocancer, 
                           color = Plot_Group, group = id)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  # Draw the individual lines (very transparent)
  geom_line(alpha = 0.3, linewidth = 0.5) +
  # Draw the individual data points
  geom_point(alpha = 0.4, size = 1) +
  
  # Overlay a thick LOESS smooth line to show the general trend of the sample
  geom_smooth(aes(group = Plot_Group), method = "loess", se = FALSE, 
              linewidth = 2, color = "black") +
              
  scale_color_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", 
                                "Incident Cancer Sub-cohort" = "#E15759")) +
  facet_wrap(~ Plot_Group) +
  theme_minimal() +
  labs(
    title = "Raw Individual Frailty Trajectories (Sample of N=100 per group)",
    subtitle = "Black line represents the LOESS smoothed trend for this sample.",
    x = "Exact Years Relative to Event (Diagnosis / Censor)",
    y = "Non-Cancer Frailty Index Score",
    color = "Cohort"
  ) +
  theme(legend.position = "none", 
        strip.text = element_text(face="bold", size=12))
```

***

### 2. Attrition and Informative Censoring Checks
If the frailest people die or drop out of the study immediately after getting cancer, your post-cancer trajectory line will artificially flatten out (because only the healthiest survivors remain). This check isolates the frailty score *just before or at* the cancer diagnosis and compares those who stayed in the study to those who dropped out.

```{r eda_attrition_check, fig.height=6, fig.width=8}
# 1. Identify individuals who have at least one survey AFTER their cancer diagnosis
retained_ids <- fi_event_all %>%
  filter(Plot_Group == "Incident Cancer Sub-cohort", Time_to_Event_Exact > 0) %>%
  distinct(id) %>%
  pull(id)

# 2. Get the closest frailty score leading up to (or exactly at) diagnosis
attrition_check <- fi_event_all %>%
  filter(Plot_Group == "Incident Cancer Sub-cohort") %>%
  # Look at the 4 years immediately leading up to diagnosis
  filter(Time_to_Event_Exact >= -4 & Time_to_Event_Exact <= 0) %>%
  group_by(id) %>%
  # If they have multiple scores in that window, take the one closest to Time 0
  arrange(desc(Time_to_Event_Exact)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    FollowUp_Status = if_else(id %in% retained_ids, 
                              "Survived / Retained in Study", 
                              "Dropped Out / Deceased")
  )

# 3. Plot the comparison
ggplot(attrition_check, aes(x = FollowUp_Status, y = fi_score_nocancer, fill = FollowUp_Status)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  # Add jittered points to see the distribution size
  geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
  scale_fill_manual(values = c("Dropped Out / Deceased" = "#d7191c", 
                               "Survived / Retained in Study" = "#abd9e9")) +
  theme_minimal() +
  labs(
    title = "Survivor Bias Check: Frailty at Time of Diagnosis",
    subtitle = "Comparing pre-diagnosis frailty based on whether they provided post-diagnosis data.",
    x = "Status in the Post-Diagnosis Period",
    y = "Frailty Index Score at (or just before) Diagnosis"
  ) +
  theme(legend.position = "none", 
        axis.text.x = element_text(face = "bold", size = 12))

# Optional: Run a quick T-test to see if the difference is statistically significant
# t.test(fi_score_nocancer ~ FollowUp_Status, data = attrition_check)
```


***

### 3. Timeline Alignments (Time-to-Event Distributions)

```{r eda_timeline_gap, fig.height=5, fig.width=8}
# 1. Calculate the time gap to the FIRST post-event survey
first_post_event_gap <- fi_event_all %>%
  filter(Plot_Group == "Incident Cancer Sub-cohort") %>%
  # Only look at post-event surveys
  filter(Time_to_Event_Exact > 0) %>%
  group_by(id) %>%
  # Grab the survey closest to the event
  arrange(Time_to_Event_Exact) %>%
  slice(1) %>%
  ungroup()

# Calculate the median wait time to print on the plot
median_wait <- median(first_post_event_gap$Time_to_Event_Exact, na.rm = TRUE)

# 2. Plot the distribution
ggplot(first_post_event_gap, aes(x = Time_to_Event_Exact)) +
  geom_histogram(binwidth = 0.5, fill = "#2c3e50", color = "white", alpha = 0.9) +
  geom_vline(xintercept = median_wait, color = "#E15759", linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = median_wait + 0.2, y = max(table(cut(first_post_event_gap$Time_to_Event_Exact, breaks=seq(0,20,0.5)))), 
           label = paste("Median:", round(median_wait, 2), "Years"), 
           color = "#E15759", fontface = "bold", hjust = 0) +
  theme_minimal() +
  labs(
    title = "Time Gap Between Cancer Diagnosis and Next FI Assessment",
    subtitle = "Visualizing the 'Short-Term' follow-up window.",
    x = "Years from Cancer Diagnosis to First Post-Event Survey",
    y = "Number of Participants"
  ) +
  scale_x_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(face = "bold", size = 11))
```

