---
title: "Frailty Index Pre/Post Cancer Trajectory Analysis"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 1_EDA-for-FI-traj.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Performs exploratory analyses specifically for frailty-index trajectory modeling, including initial data preparation, descriptive tables, and plots.
```

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Required Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(gtsummary) # Added for Baseline Characteristics Table 1
```

# 1. Data Preparation

```{r data_prep}
# Load the Imputed Data containing both FI versions
fi_long <-  
  readRDS("/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
year_levels <- c("1988", "1992", "1996", "2000", "2004", "2008", "2012", "2016", "2020")

# 1. Standardize Years & Convert to Numeric for Event-Time Math
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    Year = case_when(
      as.numeric(cycle) > 50 ~ paste0("19", cycle),
      TRUE ~ paste0("20", cycle)
    ),
    Year = factor(Year, levels = year_levels),
    Year_num = as.numeric(as.character(Year))
  )

# 2. Identify and Exclude Baseline Cancer
baseline_cancer_ids <- fi_prep %>%
  group_by(id) %>%
  arrange(Year_num) %>%
  slice(1) %>% # Get their first recorded visit
  filter(fi_cancer == 1) %>%
  pull(id)

# 3. Identify Timing of Incident Cancer
cancer_timing <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  filter(fi_cancer == 1) %>%
  group_by(id) %>%
  summarise(Year_Cancer = min(Year_num), .groups = "drop") # First occurrence

incident_cancer_ids <- cancer_timing$id

# 4. Identify Censoring Timing (For the Control Group)
# For plotting a backward timescale, we need an anchor for the no-cancer group
censor_timing <- fi_prep %>%
  filter(!id %in% c(baseline_cancer_ids, incident_cancer_ids)) %>%
  group_by(id) %>%
  summarise(Year_Last_FollowUp = max(Year_num), .groups = "drop")

# 5. Apply exclusions and the < 20 item rule
fi_clean <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  mutate(
    # Force FI to NA if they answered fewer than 20 non-cancer items
    fi_score_nocancer = if_else(n_answered_nocancer < 20, NA_real_, fi_score_nocancer),
    # Sync the category with the new NA scores
    frailty_cat_nocancer = if_else(is.na(fi_score_nocancer), NA_character_, frailty_cat_nocancer)
  )

# 6. Create Plotting Data with Group Labels
plot_data <- fi_clean %>%
  mutate(Plot_Group = if_else(id %in% incident_cancer_ids, 
                              "Incident Cancer Sub-cohort", 
                              "Censored/No-Cancer Population"),
         Plot_Group = factor(Plot_Group, levels = c("Incident Cancer Sub-cohort", "Censored/No-Cancer Population")))

# Filter for valid FI scores only
fi_valid <- plot_data %>% filter(participated == 1, !is.na(fi_score_nocancer))
```

***

# 2. Baseline Characteristics (Table 1)

Establishing baseline comparability is critical. [cite_start]Similar to analyses in major cohorts comparing incident disease groups[cite: 195], we compare the Incident Cancer and No-Cancer populations at their earliest valid study entry.

```{r baseline_table}
# NOTE: Update 'sex', 'age_at_baseline', 'education' to match your actual column names.
baseline_data <- fi_valid %>%
  group_by(id) %>%
  arrange(Year_num) %>%
  slice(1) %>% # Get first valid observation per participant
  ungroup() %>%
  select(Plot_Group, frailty_cat_nocancer, fi_score_nocancer) 
  # Add your demographics here: e.g., select(..., sex, age_at_baseline, education)

baseline_data %>%
  tbl_summary(
    by = Plot_Group,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "no"
  ) %>%
  add_p() %>%
  modify_header(label = "**Baseline Characteristic**") %>%
  bold_labels()
```

***

# 3. Data Quality & Distribution Over Calendar Year

```{r data_quality}
attrition_data <- plot_data %>%
  filter(participated == 1) %>%
  group_by(Plot_Group, Year) %>%
  summarise(
    N_Total = n(),
    N_Valid_FI = sum(!is.na(fi_score_nocancer)),
    .groups = "drop"
  ) %>%
  mutate(Percent_Active = (N_Valid_FI / N_Total) * 100)

ggplot(attrition_data, aes(x = Year, y = Percent_Active, fill = Plot_Group)) +
  geom_bar(stat = "identity", position = position_dodge(width=0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Active)), 
            position = position_dodge(width=0.8), vjust = -0.5, fontface = "bold", size = 3) +
  scale_y_continuous(limits = c(0, 115), labels = scales::label_percent(scale = 1)) +
  scale_fill_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", "Incident Cancer Sub-cohort" = "#E15759")) +
  facet_wrap(~ Plot_Group, ncol = 1) +
  theme_minimal() +
  labs(
    title = "Data Usability Rate per Cycle (Non-Cancer FI)",
    subtitle = "Percentage of active participants with >= 20 valid items.",
    x = "Study Year", y = "% Usable Data", fill = "Cohort"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 10),
        strip.text = element_text(face="bold", size=12),
        legend.position = "top")
```

```{r cross_sectional_boxplot}
medians_df <- fi_valid %>%
  group_by(Plot_Group, Year) %>%
  summarise(Median_Val = median(fi_score_nocancer, na.rm=TRUE), .groups="drop")

ggplot(fi_valid, aes(x = Year, y = fi_score_nocancer, fill = Plot_Group)) +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.2) +
  scale_fill_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", "Incident Cancer Sub-cohort" = "#E15759")) +
  facet_wrap(~ Plot_Group, ncol = 1) +
  theme_minimal() +
  labs(title = "Distribution of Non-Cancer Frailty Scores (1988 - 2020)",
       x = "Year", y = "Non-Cancer Frailty Index Score") +
  theme(axis.text.x = element_text(face = "bold", size = 11),
        strip.text = element_text(face="bold", size=12), legend.position = "none")
```

***

# 4. Event-Centered Trajectories (Backward Timescale)

[cite_start]To capture the acceleration of frailty preceding an event, studies recommend modeling frailty index scores using a backward timescale[cite: 682, 706]. Here, Time 0 is the cycle of Cancer Diagnosis. For the control group, Time 0 is their final follow-up (censoring).

```{r setup_event_centered}
# Create Event-Centered Dataset for Both Groups
fi_event_cancer <- fi_valid %>%
  filter(Plot_Group == "Incident Cancer Sub-cohort") %>%
  inner_join(cancer_timing, by = "id") %>%
  mutate(Time_to_Event = Year_num - Year_Cancer)

fi_event_censor <- fi_valid %>%
  filter(Plot_Group == "Censored/No-Cancer Population") %>%
  inner_join(censor_timing, by = "id") %>%
  mutate(Time_to_Event = Year_num - Year_Last_FollowUp)

fi_event_all <- bind_rows(fi_event_cancer, fi_event_censor)
plot_window <- seq(-16, 12, by = 4)

fi_plot_data <- fi_event_all %>%
  filter(Time_to_Event %in% plot_window)
```

## Mean Trajectory: Cancer vs. Control

```{r mean_trajectory_comparison}
mean_data_all <- fi_plot_data %>%
  group_by(Plot_Group, Time_to_Event) %>%
  summarise(
    Mean_FI = mean(fi_score_nocancer),
    SD_FI = sd(fi_score_nocancer),
    N = n(),
    .groups = "drop"
  ) %>%
  mutate(
    SE = SD_FI / sqrt(N),
    Lower_CI = Mean_FI - (1.96 * SE),
    Upper_CI = Mean_FI + (1.96 * SE)
  )

ggplot(mean_data_all, aes(x = Time_to_Event, y = Mean_FI, color = Plot_Group, group = Plot_Group)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.5, alpha = 0.5) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = plot_window) +
  scale_color_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", "Incident Cancer Sub-cohort" = "#E15759")) +
  theme_minimal() +
  labs(
    title = "Mean Trajectory: Incident Cancer vs Censored Population",
    subtitle = "Time 0 = Cancer Diagnosis OR Last Follow-up (Censored). Error bars = 95% CIs.",
    x = "Years Relative to Event (Diagnosis / Censor)",
    y = "Mean Non-Cancer Frailty Index Score",
    color = "Cohort"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 11), legend.position = "bottom")
```

***

# 5. Subgroup Stratifications

[cite_start]Frailty changes following major health events are frequently modified by baseline demographics such as age and sex[cite: 212, 213, 214]. 

```{r stratification_analysis, fig.height=8}
# NOTE: Update 'sex' to match your data column. Assuming it is a factor (e.g., 'Male', 'Female').
# If 'sex' does not exist yet, this code will need modification.

if("sex" %in% names(fi_plot_data)) {
  mean_data_stratified <- fi_plot_data %>%
    group_by(Plot_Group, Time_to_Event, sex) %>%
    summarise(
      Mean_FI = mean(fi_score_nocancer),
      N = n(),
      .groups = "drop"
    )

  ggplot(mean_data_stratified, aes(x = Time_to_Event, y = Mean_FI, color = Plot_Group)) +
    geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 0.8) +
    geom_point(size = 2.5) +
    geom_line(linewidth = 1) +
    facet_wrap(~ sex) +
    scale_x_continuous(breaks = plot_window) +
    scale_color_manual(values = c("Censored/No-Cancer Population" = "#76B7B2", "Incident Cancer Sub-cohort" = "#E15759")) +
    theme_minimal() +
    labs(
      title = "Frailty Trajectories Stratified by Sex",
      x = "Years Relative to Event",
      y = "Mean Non-Cancer Frailty Index",
      color = "Cohort"
    ) +
    theme(legend.position = "bottom")
} else {
  print("Variable 'sex' not found in dataset. Update code to run stratification.")
}
```

***

# 6. Cancer Sub-Type Analysis

[cite_start]Event severity and physiological impact can vary significantly by condition type[cite: 341]. If specific cancer site data is available, we evaluate the top specific pathways.

```{r cancer_subtype}
# NOTE: Replace 'cancer_site' with your variable identifying the type of incident cancer.
# We filter only the Cancer cohort for this.

if("cancer_site" %in% names(fi_event_cancer)) {
  
  # Identify top 4 most common incident cancers
  top_cancers <- fi_event_cancer %>%
    filter(Time_to_Event == 0) %>%
    count(cancer_site, sort = TRUE) %>%
    head(4) %>%
    pull(cancer_site)
  
  mean_subtype <- fi_event_cancer %>%
    filter(cancer_site %in% top_cancers) %>%
    filter(Time_to_Event %in% plot_window) %>%
    group_by(cancer_site, Time_to_Event) %>%
    summarise(Mean_FI = mean(fi_score_nocancer), N=n(), .groups="drop")
  
  ggplot(mean_subtype, aes(x = Time_to_Event, y = Mean_FI, color = cancer_site)) +
    geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
    geom_point(size = 2.5) +
    geom_line(linewidth = 1) +
    scale_x_continuous(breaks = plot_window) +
    facet_wrap(~ cancer_site) +
    theme_minimal() +
    labs(
      title = "Frailty Trajectory by Top 4 Incident Cancer Types",
      x = "Years Relative to Diagnosis",
      y = "Mean Non-Cancer FI"
    ) +
    theme(legend.position = "none")
} else {
  print("Variable 'cancer_site' not found. Update code to run subtype analysis.")
}
```

***

# 7. Check for Informative Censoring (Survivor Bias)

To ensure post-cancer trajectories are not artificially flattened by the most frail individuals dropping out or dying, we assess the "Time 0" frailty score stratified by whether the participant remained in the study at the subsequent cycle.

```{r survivor_bias_check}
# Determine if a participant has data at Time_to_Event == 4
retained_at_plus4 <- fi_event_cancer %>%
  filter(Time_to_Event == 4) %>%
  pull(id) %>%
  unique()

# Label their status at Time_to_Event == 0
survivor_check_data <- fi_event_cancer %>%
  filter(Time_to_Event == 0) %>%
  mutate(
    Status_at_Next_Cycle = if_else(id %in% retained_at_plus4, 
                                   "Remained in Study", 
                                   "Dropped Out / Deceased")
  )

ggplot(survivor_check_data, aes(x = Status_at_Next_Cycle, y = fi_score_nocancer, fill = Status_at_Next_Cycle)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  scale_fill_manual(values = c("Dropped Out / Deceased" = "#d7191c", "Remained in Study" = "#abd9e9")) +
  labs(
    title = "Informative Censoring Check at Diagnosis (Time = 0)",
    subtitle = "Are those who drop out at the next cycle systematically more frail at diagnosis?",
    x = "Status at Next Study Cycle (Time = +4)",
    y = "Frailty Index at Diagnosis (Time = 0)"
  ) +
  theme(legend.position = "none", axis.text.x = element_text(face = "bold", size = 11))
```
```