---
title: "Frailty Index Pre/Post Cancer Trajectory Analysis"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 1_EDA.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Performs early exploratory data analysis for frailty trajectories around cancer diagnosis, including demographic summaries and trajectory visualizations.
```

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(gtsummary) # For demographic tables
```

# 1. Data Preparation

```{r data_prep}
# Load the Imputed Data
fi_long <-  
  readRDS("/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
year_levels <- c("1988", "1992", "1996", "2000", "2004", "2008", "2012", "2016", "2020")

# 1. Standardize Years, Convert to Numeric, & Calculate Exact Age
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    Year = case_when(
      as.numeric(cycle) > 50 ~ paste0("19", cycle),
      TRUE ~ paste0("20", cycle)
    ),
    Year = factor(Year, levels = year_levels),
    Year_num = as.numeric(as.character(Year)),
    
    # Calculate exact Age using your specific timeline logic
    Age = (worked_rtmnyr - dbmy09) / 12
  )

# 2. Identify Baseline Cancer (To Exclude)
baseline_cancer_ids <- fi_prep %>%
  group_by(id) %>%
  arrange(Year_num) %>%
  slice(1) %>% 
  filter(fi_cancer == 1) %>%
  pull(id)

# 3. Identify Timing of Incident Cancer
cancer_timing <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  filter(fi_cancer == 1) %>%
  group_by(id) %>%
  summarise(Year_Cancer = min(Year_num), .groups = "drop") 

incident_cancer_ids <- cancer_timing$id

# 4. Filter for Incident Cancer Cohort & Apply validity rules
fi_incident_base <- fi_prep %>%
  filter(id %in% incident_cancer_ids) %>%
  mutate(
    # Force FI to NA if they answered fewer than 20 non-cancer items
    fi_score_nocancer = if_else(n_answered_nocancer < 20, NA_real_, fi_score_nocancer),
    frailty_cat_nocancer = if_else(is.na(fi_score_nocancer), NA_character_, frailty_cat_nocancer)
  )

# 5. Create the purely valid dataset for trajectory plotting
fi_valid <- fi_incident_base %>%
  filter(participated == 1, !is.na(fi_score_nocancer))
```

***

# 2. Descriptive Characteristics

## 2.1 At Baseline (Study Entry)

This table summarizes the incident cancer cohort at their very first valid study cycle (well before their cancer diagnosis).

```{r baseline_table}
baseline_data <- fi_valid %>%
  group_by(id) %>%
  arrange(Year_num) %>%
  slice(1) %>% 
  ungroup() %>%
  select(Age, frailty_cat_nocancer, fi_score_nocancer)

baseline_data %>%
  tbl_summary(
    label = list(
      Age ~ "Age at Baseline (years)",
      frailty_cat_nocancer ~ "Frailty Category",
      fi_score_nocancer ~ "Frailty Index Score"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "no"
  ) %>%
  modify_header(label = "**Baseline Characteristic**") %>%
  bold_labels()
```

## 2.2 At Cancer Diagnosis (Time = 0)

This table summarizes the cohort exactly at the cycle they were first diagnosed with incident cancer. 

```{r cancer_diagnosis_descriptives}
diagnosis_data <- fi_valid %>%
  inner_join(cancer_timing, by = "id") %>%
  filter(Year_num == Year_Cancer) %>%
  ungroup() %>%
  select(Age, frailty_cat_nocancer, fi_score_nocancer) 

diagnosis_data %>%
  tbl_summary(
    label = list(
      Age ~ "Age at Diagnosis (years)",
      frailty_cat_nocancer ~ "Frailty Category",
      fi_score_nocancer ~ "Frailty Index Score"
    ),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "no"
  ) %>%
  modify_header(label = "**Characteristics at Diagnosis**") %>%
  bold_labels()
```

***

# 3. Cross-Sectional Descriptive Analysis (Calendar Year)

```{r attrition_plot}
# Denominator = active participants (participated == 1) in that specific cycle
attrition_data <- fi_incident_base %>%
  filter(participated == 1) %>%
  group_by(Year) %>%
  summarise(
    N_Eligible = n(),
    N_Valid_FI = sum(!is.na(fi_score_nocancer)),
    .groups = "drop"
  ) %>%
  mutate(Percent_Active = (N_Valid_FI / N_Eligible) * 100)

ggplot(attrition_data, aes(x = Year, y = Percent_Active)) +
  geom_bar(stat = "identity", fill = "#E15759", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Active)), 
            vjust = -0.5, fontface = "bold", size = 3) +
  geom_text(aes(label = paste0("(n=", N_Valid_FI, "/", N_Eligible, ")")), 
            vjust = 1.5, color = "white", size = 2.5) +
  scale_y_continuous(limits = c(0, 115), labels = scales::label_percent(scale = 1)) +
  theme_minimal() +
  labs(title = "Data Usability Rate per Cycle (Incident Cancer Cohort)", 
       subtitle = "Percentage of active participants in a given cycle providing >= 20 valid items.",
       x = "Study Year", y = "% Usable Data")
```

```{r cross_sectional_boxplot}
medians_df <- fi_valid %>%
  group_by(Year) %>%
  summarise(Median_Val = median(fi_score_nocancer, na.rm=TRUE), .groups="drop")

ggplot(fi_valid, aes(x = Year, y = fi_score_nocancer)) +
  geom_boxplot(fill = "#E15759", outlier.shape = 1, outlier.alpha = 0.2, alpha = 0.6) +
  geom_text(data = medians_df, aes(x = Year, y = Median_Val, label = sprintf("%.3f", Median_Val)),
            vjust = -1.2, color = "black", fontface = "bold", size = 3.5) +
  theme_minimal() +
  labs(title = "Distribution of Frailty Scores by Calendar Year", 
       x = "Year", y = "Non-Cancer FI Score")
```

```{r category_prevalence}
cat_data <- fi_valid %>%
  filter(!is.na(frailty_cat_nocancer)) %>%
  mutate(frailty_cat_nocancer = factor(frailty_cat_nocancer, 
                                       levels = c("Non-frail", "Pre-frail", "Mildly frail", "Mod/Sev frail"))) %>%
  group_by(Year, frailty_cat_nocancer) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)

ggplot(cat_data, aes(x = Year, y = Percentage, fill = frailty_cat_nocancer)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  scale_fill_manual(values = c("Non-frail" = "#2c7bb6", "Pre-frail" = "#abd9e9", 
                               "Mildly frail" = "#fdae61", "Mod/Sev frail" = "#d7191c")) +
  geom_text(aes(label = ifelse(Percentage > 4, paste0(round(Percentage, 0), "%"), "")), 
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  theme_minimal() +
  labs(title = "Clinical Frailty Category Prevalence by Calendar Year", 
       x = "Year", y = "Percentage (%)", fill = "Category") +
  theme(legend.position = "top")
```

***

# 4. Event-Centered Trajectories (Time = 0 at Diagnosis)

```{r setup_event_centered}
plot_window <- seq(-16, 12, by = 4)

# Create Event-Centered Dataset for Availability (Includes NAs to calculate denominator)
fi_event_base <- fi_incident_base %>%
  inner_join(cancer_timing, by = "id") %>%
  mutate(Time_to_Cancer = Year_num - Year_Cancer) %>%
  filter(Time_to_Cancer %in% plot_window)

# Create purely valid Event-Centered Dataset for Trajectories
fi_event_cancer <- fi_valid %>%
  inner_join(cancer_timing, by = "id") %>%
  mutate(Time_to_Cancer = Year_num - Year_Cancer) %>%
  filter(Time_to_Cancer %in% plot_window)
```

## Data Availability & Distribution

```{r data_availability}
n_data_event <- fi_event_base %>%
  filter(participated == 1) %>%
  group_by(Time_to_Cancer) %>%
  summarise(
    N_Eligible = n(),
    N_Valid_FI = sum(!is.na(fi_score_nocancer)),
    .groups = "drop"
  ) %>%
  mutate(Percent_Active = (N_Valid_FI / N_Eligible) * 100)

ggplot(n_data_event, aes(x = factor(Time_to_Cancer), y = Percent_Active)) +
  geom_bar(stat = "identity", fill = "#E15759", width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Active)), 
            vjust = -0.5, fontface = "bold", size = 3) +
  geom_text(aes(label = paste0("(n=", N_Valid_FI, "/", N_Eligible, ")")), 
            vjust = 1.5, color = "white", size = 2.5) +
  scale_y_continuous(limits = c(0, 115), labels = scales::label_percent(scale = 1)) +
  theme_minimal() +
  labs(title = "Data Availability Relative to Cancer Diagnosis",
       subtitle = "Percentage of active participants at each relative timepoint providing valid data.",
       x = "Years Relative to Cancer Diagnosis (0 = Diagnosis Year)", y = "% Usable Data")
```

```{r event_centered_boxplot}
label_data <- fi_event_cancer %>%
  group_by(Time_to_Cancer) %>%
  summarise(Median_FI = median(fi_score_nocancer), N = n(), .groups = "drop")

ggplot(fi_event_cancer, aes(x = factor(Time_to_Cancer), y = fi_score_nocancer)) +
  geom_vline(xintercept = which(levels(factor(fi_event_cancer$Time_to_Cancer)) == "0"), 
             color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_boxplot(fill = "#E15759", alpha = 0.6, outlier.alpha = 0.3) +
  geom_text(data = label_data, aes(y = Median_FI, label = sprintf("%.2f", Median_FI)),
            vjust = -0.8, color = "black", fontface = "bold", size = 3.5) +
  geom_text(data = label_data, aes(y = -0.02, label = paste0("N=", N)),
            color = "black", size = 3.5, fontface = "italic") +
  theme_minimal() +
  labs(title = "Distribution of Frailty Relative to Cancer Diagnosis", 
       x = "Years Relative to Diagnosis", y = "Non-Cancer FI Score")
```

## Individual & Mean Trajectories

```{r spaghetti_plot}
set.seed(123)
sample_ids <- sample(unique(fi_event_cancer$id), 100) 

fi_spaghetti <- fi_event_cancer %>% filter(id %in% sample_ids)

ggplot(fi_spaghetti, aes(x = Time_to_Cancer, y = fi_score_nocancer, group = id)) +
  geom_vline(xintercept = 0, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_line(alpha = 0.2, color = "#E15759") +
  geom_point(alpha = 0.4, size = 1, color = "#E15759") +
  geom_smooth(aes(group = 1), method = "loess", color = "black", linewidth = 1.5, se = FALSE) +
  scale_x_continuous(breaks = plot_window) +
  theme_minimal() +
  labs(title = "Individual Trajectories (Random Subsample N=100)",
       subtitle = "Black line represents the LOESS smoothed average.",
       x = "Years Relative to Cancer Diagnosis", y = "Non-Cancer FI Score")
```

```{r mean_trajectory}
mean_data_all <- fi_event_cancer %>%
  group_by(Time_to_Cancer) %>%
  summarise(Mean_FI = mean(fi_score_nocancer), SD_FI = sd(fi_score_nocancer), N = n(), .groups = "drop") %>%
  mutate(SE = SD_FI / sqrt(N), Lower_CI = Mean_FI - (1.96 * SE), Upper_CI = Mean_FI + (1.96 * SE))

ggplot(mean_data_all, aes(x = Time_to_Cancer, y = Mean_FI)) +
  geom_vline(xintercept = 0, color = "darkred", linetype = "dashed", linewidth = 1) +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.3, alpha = 0.5) +
  geom_point(size = 3, color = "#d7191c") +
  geom_line(linewidth = 1, color = "#d7191c") +
  scale_x_continuous(breaks = plot_window) +
  theme_minimal() +
  labs(title = "Unadjusted Mean Frailty Trajectory",
       subtitle = "Error bars represent 95% CIs. Notice the slope changes around Time 0.",
       x = "Years Relative to Diagnosis", y = "Mean Non-Cancer FI Score")
```

***

# 5. Check for Informative Censoring (Survivor Bias)

```{r survivor_bias_check}
# Determine if a participant has data at Time_to_Cancer == 4
retained_at_plus4 <- fi_event_cancer %>%
  filter(Time_to_Cancer == 4) %>%
  pull(id) %>%
  unique()

survivor_check_data <- fi_event_cancer %>%
  filter(Time_to_Cancer == 0) %>%
  mutate(Status_at_Next_Cycle = if_else(id %in% retained_at_plus4, "Remained in Study", "Dropped Out / Deceased"))

ggplot(survivor_check_data, aes(x = Status_at_Next_Cycle, y = fi_score_nocancer, fill = Status_at_Next_Cycle)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  scale_fill_manual(values = c("Dropped Out / Deceased" = "#d7191c", "Remained in Study" = "#abd9e9")) +
  labs(title = "Informative Censoring Check at Diagnosis (Time = 0)",
       subtitle = "Are those who drop out at the next cycle systematically more frail at diagnosis?",
       x = "Status at Next Study Cycle (Time = +4)", y = "FI Score at Diagnosis (Time = 0)") +
  theme(legend.position = "none")
```
```