---
title: "Frailty Trajectories Before and After Incident Cancer: Case only cohort"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 2.1_case only trajectory.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Analyzes frailty trajectories among cancer cases only, focusing on within-case frailty patterns before and after incident cancer diagnosis.
```

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Required Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(lmerTest) # For p-values in lmer
library(ggeffects)
library(patchwork)
```

# 1. Data Preparation

**Inclusion & Exclusion Criteria:**
1. Available covariates at baseline.
2. Participants need to have cancer diagnosis during the follow-up.
3. We remove prevalent cancers at baseline.
4. We ensure all participants have at least 2 longitudinal observations with FI.
5. FI is considered missing if with less than 20 out of 27 frailty items measured.

```{r data_prep}
# Define data path
data_path <- "/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
fi_long <- readRDS(data_path)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")

# 1A. Extract Baseline Covariates (Cycle 86)
baseline_covs <- fi_long %>%
  filter(cycle == "86") %>%
  select(
    id,
    age_months   = age_enrollment, 
    base_race    = race,
    base_marital = marital_status,
    base_living  = living_arr,
    base_pckgr   = pckgr
  ) %>%
  mutate(base_age = age_months / 12) %>% 
  select(-age_months) %>%
  drop_na()

# 1B. Prepare Continuous Time Variables
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    worked_rtmnyr = as.numeric(worked_rtmnyr),
    cancer_dateca = as.numeric(cancer_dateca),
    dbmy09        = as.numeric(dbmy09)
  )

# 1C. Establish Participant Timelines & Filter to Incident Cancer Only
participant_timelines <- fi_prep %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  summarise(
    First_Survey_Month = first(worked_rtmnyr),
    Cancer_Month = first(cancer_dateca),
    Has_Cancer = first(cancer_cancer) == 1,
    .groups = "drop"
  ) %>%
  mutate(
    # Incident Cancer: Diagnosed strictly AFTER their first survey
    Is_Incident_Cancer = Has_Cancer & (!is.na(Cancer_Month) & Cancer_Month > First_Survey_Month)
  )

incident_ids <- participant_timelines %>% filter(Is_Incident_Cancer) %>% pull(id)
# 1D. Create Piecewise Model Dataset (UPDATED)
piecewise_data <- fi_prep %>%
  filter(id %in% incident_ids) %>%
  inner_join(participant_timelines %>% select(id, Cancer_Month), by = "id") %>%
  left_join(baseline_covs, by = "id") %>%
  filter(!is.na(base_age), participated == 1, !is.na(fi_score_nocancer)) %>%
  
  mutate(
    Time_to_Event_Exact = (worked_rtmnyr - Cancer_Month) / 12,
    TimePre = if_else(Time_to_Event_Exact < 0, Time_to_Event_Exact, 0),
    TimePost = if_else(Time_to_Event_Exact >= 0, Time_to_Event_Exact, 0),
    CancerShift = if_else(Time_to_Event_Exact >= 0, 1, 0),
    
    # Center Age (this is still valid because age is numeric)
    Age_Centered = base_age - mean(base_age, na.rm = TRUE),
    
    # Treat pack-years as a factor and safely handle the original NAs
    base_pckgr = as.factor(if_else(is.na(base_pckgr), "Missing", as.character(base_pckgr)))
  ) %>%
  
  # Ensure other covariates are factors
  mutate(
    base_race = as.factor(base_race),
    base_marital = as.factor(base_marital),
    base_living = as.factor(base_living)
  )
```

------------------------------------------------------------------------

# 2. Exploratory Data Analysis (EDA)

Before modeling, it is essential to visualize the raw trajectories to ensure the linear piecewise assumption is appropriate for this data.

```{r eda_spaghetti}
set.seed(123)
sample_ids <- sample(unique(piecewise_data$id), 100)

spaghetti_data <- piecewise_data %>%
  filter(id %in% sample_ids, Time_to_Event_Exact >= -16, Time_to_Event_Exact <= 12)

ggplot(spaghetti_data, aes(x = Time_to_Event_Exact, y = fi_score_nocancer, group = id)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  geom_line(alpha = 0.3, color = "#E15759") +
  geom_point(alpha = 0.4, size = 1, color = "#E15759") +
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "black", linewidth = 1.5) +
  theme_minimal() +
  labs(
    title = "Raw Individual Frailty Trajectories (Sample of N=100)",
    subtitle = "Black line represents the LOESS smoothed trend.",
    x = "Exact Years Relative to Cancer Diagnosis",
    y = "Non-Cancer Frailty Index Score"
  )
```

------------------------------------------------------------------------

# 4. Piecewise Linear Mixed-Effects Models

We fit two models to evaluate the pre-diagnosis slope, the immediate shift at diagnosis, and the post-diagnosis slope. We use a random intercept model `(1 | id)`.

### 4A. Crude Model

Unadjusted for baseline demographic covariates (only centering age for mathematical stability).

```{r model_crude}
model_crude <- lmer(
  fi_score_nocancer ~ TimePre + CancerShift + TimePost + Age_Centered + (1 + TimePre + TimePost | id), 
  data = piecewise_data,
  REML = TRUE
)

summary(model_crude)
```

### 4B. Fully Adjusted Model

Adjusted for baseline age, race, marital status, living arrangement, and pack-years of smoking.

```{r model_adjusted}
model_adj <- lmer(
  fi_score_nocancer ~ TimePre + CancerShift + TimePost + 
    Age_Centered + base_race + base_marital + base_living + base_pckgr + 
    (1 + TimePre + TimePost | id), 
  data = piecewise_data,
  REML = TRUE
)

summary(model_adj)
```

------------------------------------------------------------------------

# 5. Visualizing the Model-Predicted Trajectory

Using the fully adjusted LME model, we predict and plot the expected trajectory of frailty for an "average" participant (holding covariates at their reference levels or sample means).

```{r plot_predicted_model}
# Update the dummy data generation for the plot
dummy_data <- data.frame(Time_to_Event_Exact = seq(-16, 16, by = 0.5)) %>%
  mutate(
    TimePre = if_else(Time_to_Event_Exact < 0, Time_to_Event_Exact, 0),
    TimePost = if_else(Time_to_Event_Exact >= 0, Time_to_Event_Exact, 0),
    CancerShift = if_else(Time_to_Event_Exact >= 0, 1, 0),
    Age_Centered = 0,             
    
    # Set factors to their reference levels
    base_race = levels(piecewise_data$base_race)[1],          
    base_marital = levels(piecewise_data$base_marital)[1],     
    base_living = levels(piecewise_data$base_living)[1],
    base_pckgr = "Never" # Or whatever level you want as the reference!
  )

# Predict and plot as normal
dummy_data$Predicted_FI <- predict(model_adj, newdata = dummy_data, re.form = NA)

# 3. Plot
ggplot(dummy_data, aes(x = Time_to_Event_Exact, y = Predicted_FI)) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1) +
  geom_line(color = "#E15759", linewidth = 1.5) +
  scale_x_continuous(breaks = seq(-16, 16, by = 4)) +
  theme_minimal() +
  labs(
    title = "Model-Predicted Frailty Trajectory (Incident Cancer Cohort)",
    subtitle = paste("Adjusted for Baseline Age, Race (", dummy_data$base_race[1], 
                     "), Marital Status (", dummy_data$base_marital[1], 
                     "), \nLiving Arrangement (", dummy_data$base_living[1], 
                     "), and Pack-years.", sep=""),
    x = "Exact Years Relative to Cancer Diagnosis",
    y = "Predicted Non-Cancer Frailty Index"
  ) +
  theme(axis.text.x = element_text(face = "bold", size = 11))
```
