---
title: "Sensitivity Analyses: Frailty Trajectories Before and After Incident Cancer"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 2.2.2_sensitivity analysis.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Runs sensitivity analyses for frailty trajectories before and after incident cancer, including alternative modeling or analytic restrictions intended to assess robustness.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 11, fig.height = 6
)

library(dplyr)
library(tidyr)
library(lubridate)
library(lme4)
library(lmerTest)
library(ggplot2)
library(knitr)
library(broom.mixed)
```

## 1. Data Preparation

We integrate the time variables you generated (specifically `years_to_death`) to create a `terminal_decline` indicator.

```{r data_prep}
# Load your pre-processed dataset containing the calculated time variables
data_path <- "/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
fi_long <- readRDS(data_path)

# 1. Re-establish Baseline Covariates (Cycle 86)
baseline_covs <- fi_long %>%
  filter(cycle == "86") %>%
  transmute(
    id,
    base_race    = as.factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = as.factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = as.factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = as.factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )

# 2. Build the Analytic Dataset & Add Terminal Decline Indicator
analysis_long <- fi_long %>%
  filter(participated == 1, !is.na(fi_score_nocancer)) %>%
  left_join(baseline_covs, by = "id") %>%
  
  # Ensure we have baseline age Z-score (using age_at_cycle for the first observation)
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  mutate(base_age = first(age_at_cycle)) %>%
  ungroup() %>%
  mutate(base_age_z = as.numeric(scale(base_age))) %>%
  
  # Remove Prevalent Cancers (cancer happened before or exactly at their first survey)
  group_by(id) %>%
  mutate(is_prevalent = if_else(!is.na(years_to_cancer) & first(years_to_cancer) >= 0, 1, 0)) %>%
  filter(is_prevalent == 0) %>%
  
  # Keep only those with at least 2 observations
  filter(n() >= 2) %>%
  ungroup() %>%
  
  # Generate Model Variables
  mutate(
    # Jump variable
    Cancer_Status = if_else(!is.na(years_to_cancer) & years_to_cancer >= 0, 1L, 0L),
    
    # Post-slope variable
    Years_Since_Cancer = if_else(Cancer_Status == 1L, years_to_cancer, 0),
    
    # NEW: Terminal Decline Indicator (Last year of life)
    # years_to_death is negative when the survey is taken BEFORE death.
    # Therefore, >= -1 and <= 0 captures the 12 months immediately preceding death.
    terminal_decline = if_else(!is.na(years_to_death) & years_to_death >= -1 & years_to_death <= 0, 1L, 0L)
  )

cat("Total observations:", nrow(analysis_long), "\n")
cat("Total unique participants:", length(unique(analysis_long$id)), "\n")
```

------------------------------------------------------------------------

## 2. Updated Main Model (Adjusted for Terminal Decline)

By adding the `terminal_decline` variable, the model "absorbs" the rapid spike in frailty that occurs right before death, preventing it from artificially steepening the `Years_Since_Cancer` slope (reducing survival bias).

```{r main_model}
mod_main <- lmer(
  fi_score_nocancer ~ years_since_baseline + Cancer_Status + Years_Since_Cancer + 
    terminal_decline + base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + years_since_baseline | id),
  data = analysis_long,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod_main)
```

------------------------------------------------------------------------

## 3. Sensitivity Analysis 1: Exclude Acute Phase (First 6 Months)

To ensure the "jump" isn't just acute surgical or chemo toxicity, we drop any frailty measurements taken within 0.5 years (6 months) after diagnosis.

```{r sen_acute}
# Filter out observations that occurred within 6 months post-cancer
data_sen1 <- analysis_long %>%
  filter(!(Cancer_Status == 1 & years_to_cancer <= 0.5))

mod_sen1 <- lmer(
  fi_score_nocancer ~ years_since_baseline + Cancer_Status + Years_Since_Cancer + 
    terminal_decline + base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + years_since_baseline | id),
  data = data_sen1,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod_sen1)
```

------------------------------------------------------------------------

## 4. Sensitivity Analysis 2: Exclude High Baseline Frailty (Ceiling Effect)

We restrict the cohort to participants who were relatively robust/pre-frail (FI \<= 0.25) at their very first measurement to ensure mathematical ceiling effects aren't compressing the cancer jump.

```{r sen_ceiling}
# Extract baseline FI for each person
baseline_fi <- analysis_long %>%
  group_by(id) %>%
  arrange(years_since_baseline) %>%
  slice(1) %>%
  select(id, base_fi = fi_score_nocancer) %>%
  ungroup()

# Filter dataset to only those with baseline FI <= 0.25
data_sen2 <- analysis_long %>%
  left_join(baseline_fi, by = "id") %>%
  filter(base_fi <= 0.25)

mod_sen2 <- lmer(
  fi_score_nocancer ~ years_since_baseline + Cancer_Status + Years_Since_Cancer + 
    terminal_decline + base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + years_since_baseline | id),
  data = data_sen2,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod_sen2)
```

------------------------------------------------------------------------

## 5. Sensitivity Analysis 3: Quadratic Time (Non-Linear Normal Aging)

We add a squared term for baseline time to allow the "normal aging" trajectory to curve upward, checking if the cancer jump still remains significant.

```{r sen_quad}
mod_sen3 <- lmer(
  fi_score_nocancer ~ years_since_baseline + I(years_since_baseline^2) + 
    Cancer_Status + Years_Since_Cancer + I(Years_Since_Cancer^2) +
    terminal_decline + base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + years_since_baseline | id),
  data = analysis_long,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod_sen3)
```

------------------------------------------------------------------------

## 6. Sensitivity Analysis 4: Age as the Timescale

We swap `years_since_baseline` out for `age_at_cycle` as the primary biological clock. *Note: `base_age_z` is removed from the fixed effects here, because current age is already accounting for it.*

```{r sen_age}
mod_sen4 <- lmer(
  fi_score_nocancer ~ age_at_cycle + Cancer_Status + Years_Since_Cancer + 
    terminal_decline + base_race + base_marital + base_living + base_pckgr +
    (1 + age_at_cycle | id),
  data = analysis_long,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod_sen4)
```

------------------------------------------------------------------------

## 7. Results Comparison Summary

We extract the main coefficients of interest (the sudden jump and the aftermath slope) across all models to see if the statistical story holds up.

```{r compare_results}
extract_coefs <- function(model, model_name, time_var) {
  res <- tidy(model)
  
  jump <- res %>% filter(term == "Cancer_Status") %>% pull(estimate)
  jump_p <- res %>% filter(term == "Cancer_Status") %>% pull(p.value)
  
  slope <- res %>% filter(term == "Years_Since_Cancer") %>% pull(estimate)
  slope_p <- res %>% filter(term == "Years_Since_Cancer") %>% pull(p.value)
  
  term_dec <- res %>% filter(term == "terminal_decline") %>% pull(estimate)
  
  data.frame(
    Model = model_name,
    `Acute Jump (Estimate)` = round(jump, 4),
    `Jump p-value` = format.pval(jump_p, digits = 3),
    `Post-Cancer Slope (Estimate)` = round(slope, 4),
    `Slope p-value` = format.pval(slope_p, digits = 3),
    `Terminal Decline Spike` = round(term_dec, 4),
    check.names = FALSE
  )
}

# Compile Table
results_table <- bind_rows(
  extract_coefs(mod_main, "Main (w/ Terminal Decline)", "years_since_baseline"),
  extract_coefs(mod_sen1, "Sen 1: Exclude <=6 months post-dx", "years_since_baseline"),
  extract_coefs(mod_sen2, "Sen 2: Exclude FI > 0.25", "years_since_baseline"),
  extract_coefs(mod_sen3, "Sen 3: Quadratic Time", "years_since_baseline"),
  extract_coefs(mod_sen4, "Sen 4: Age as Timescale", "age_at_cycle")
)

kable(results_table, caption = "Comparison of Cancer Impact Across Sensitivity Models")
```



---

## 8. Visualization: Comparing Sensitivity Models

To visually compare how our modeling assumptions alter the trajectory, we generate event-centered prediction grids. 
For models using `years_since_baseline`, we anchor "Time Zero" to the median time-to-diagnosis. 
For the age-based model, we anchor "Time Zero" to the median age-at-diagnosis.

```{r vis_sensitivity, fig.width=14, fig.height=10}
# Use the model frame to guarantee factor levels match what mod_sen3 expects
mf <- model.frame(mod_sen3)

ref_level <- function(x) {
  tab <- table(x)
  if ("Missing" %in% names(tab)) tab <- tab[names(tab) != "Missing"]
  names(which.max(tab))
}

# 1) Anchor diagnosis at the median time-to-diagnosis among cases
#    Identity: years_since_baseline - years_to_cancer = time from baseline to diagnosis (constant per id)
dx_time_by_id <- analysis_long %>%
  filter(!is.na(years_to_cancer)) %>%
  group_by(id) %>%
  summarise(
    everCancer_id = as.integer(any(!is.na(cancer_dateca))),
    tDx = min(years_since_baseline - years_to_cancer, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(everCancer_id == 1L)

tDx_median <- median(dx_time_by_id$tDx, na.rm = TRUE)

# 2) Event-centered window (paper-style)
min_rel <- -15
max_rel <-  25

# Prevent negative years_since_baseline in prediction grid
if (tDx_median + min_rel < 0) min_rel <- -tDx_median

# 3) Build prediction grids
grid_never <- tibble(tRel = seq(min_rel, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Never cancer (counterfactual)",
    years_since_baseline = tDx_median + tRel,
    Cancer_Status = 0L,
    Years_Since_Cancer = 0
  )

grid_cancer_pre <- tibble(tRel = seq(min_rel, 0, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    years_since_baseline = tDx_median + tRel,
    Cancer_Status = 0L,
    Years_Since_Cancer = 0
  )

grid_cancer_post <- tibble(tRel = seq(0, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    years_since_baseline = tDx_median + tRel,
    Cancer_Status = 1L,
    Years_Since_Cancer = tRel
  )

grid_rel <- bind_rows(grid_never, grid_cancer_pre, grid_cancer_post)

# 4) Set all remaining covariates exactly as required by mod_sen3
grid_rel <- grid_rel %>%
  mutate(
    # fixed effects only; set a non-terminal reference trajectory
    terminal_decline = 0L,

    # set mean age (z-score)
    base_age_z = 0,

    # set reference categories with factor levels exactly matching the fitted model
    base_race    = factor(ref_level(mf$base_race),    levels = levels(mf$base_race)),
    base_marital = factor(ref_level(mf$base_marital), levels = levels(mf$base_marital)),
    base_living  = factor(ref_level(mf$base_living),  levels = levels(mf$base_living)),
    base_pckgr   = factor(ref_level(mf$base_pckgr),   levels = levels(mf$base_pckgr)),

    # include an id (even though re.form = NA), using an existing model-frame level
    id = mf$id[1]
  )

# 5) Predict population-average (fixed-effect) values
grid_rel$pred_FI <- predict(mod_sen3, newdata = grid_rel, re.form = NA)

# Split for plotting
grid_never_plot <- grid_rel %>% filter(scenario == "Never cancer (counterfactual)")
grid_pre_plot   <- grid_rel %>% filter(scenario == "Incident cancer at t=0", Cancer_Status == 0L)
grid_post_plot  <- grid_rel %>% filter(scenario == "Incident cancer at t=0", Cancer_Status == 1L)

# Jump at t=0
y_pre_jump  <- grid_pre_plot  %>% filter(dplyr::near(tRel, 0)) %>% pull(pred_FI)
y_post_jump <- grid_post_plot %>% filter(dplyr::near(tRel, 0)) %>% pull(pred_FI)

# 6) Plot
ggplot() +
  geom_line(data = grid_never_plot, aes(x = tRel, y = pred_FI),
            color = "#E69F00", linetype = "dashed", linewidth = 1.2) +
  geom_line(data = grid_pre_plot, aes(x = tRel, y = pred_FI),
            color = "#0072B2", linewidth = 1.2) +
  geom_line(data = grid_post_plot, aes(x = tRel, y = pred_FI),
            color = "#0072B2", linewidth = 1.2) +
  geom_segment(aes(x = 0, xend = 0, y = y_pre_jump, yend = y_post_jump),
               color = "#0072B2", linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray30") +
  scale_x_continuous(breaks = seq(ceiling(min_rel), floor(max_rel), by = 1)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "bottom"
  ) +
  labs(
    title = "Predicted FI Trajectories Before/After Incident Cancer (Sensitivity Model: Quadratic Time)",
    subtitle = paste0(
      "Event-centered; diagnosis anchored at median time-to-diagnosis = ",
      round(tDx_median, 1), " years since baseline; terminal_decline fixed at 0."
    ),
    x = "Years relative to cancer diagnosis (0 = diagnosis)",
    y = "Predicted non-cancer FI"
  )
```


```{r, fig.width=14, fig.height=10}
# Use model.frame to guarantee factor levels match mod_sen4
mf <- model.frame(mod_sen4)

ref_level <- function(x) {
  tab <- table(x)
  if ("Missing" %in% names(tab)) tab <- tab[names(tab) != "Missing"]
  names(which.max(tab))
}

# 1) Choose an age anchor for diagnosis (age at cancer among cases)
#    Because the model time scale is age_at_cycle, an event-centered plot needs
#    an "age at diagnosis" anchor: age_at_dx = age_at_cycle - years_to_cancer
dx_age_by_id <- analysis_long %>%
  filter(!is.na(years_to_cancer) & !is.na(age_at_cycle)) %>%
  group_by(id) %>%
  summarise(
    everCancer_id = as.integer(any(!is.na(cancer_dateca))),
    age_at_dx = min(age_at_cycle - years_to_cancer, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(everCancer_id == 1L)

age_dx_median <- median(dx_age_by_id$age_at_dx, na.rm = TRUE)

# 2) Event-centered window (years relative to diagnosis)
min_rel <- -15
max_rel <-  25

# Guardrails: keep ages plausible (optional)
age_min_allowed <- max(0, min(mf$age_at_cycle, na.rm = TRUE))
age_max_allowed <- max(mf$age_at_cycle, na.rm = TRUE)

# 3) Build prediction grids
grid_never <- tibble(tRel = seq(min_rel, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Never cancer (counterfactual)",
    age_at_cycle = age_dx_median + tRel,
    Cancer_Status = 0L,
    Years_Since_Cancer = 0
  )

grid_cancer_pre <- tibble(tRel = seq(min_rel, 0, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    age_at_cycle = age_dx_median + tRel,
    Cancer_Status = 0L,
    Years_Since_Cancer = 0
  )

grid_cancer_post <- tibble(tRel = seq(0, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    age_at_cycle = age_dx_median + tRel,
    Cancer_Status = 1L,
    Years_Since_Cancer = tRel
  )

grid_rel <- bind_rows(grid_never, grid_cancer_pre, grid_cancer_post) %>%
  # (Optional) clip to observed age range to avoid extrapolation
  mutate(age_at_cycle = pmin(pmax(age_at_cycle, age_min_allowed), age_max_allowed))

# 4) Set remaining covariates to a reference profile with correct factor levels
grid_rel <- grid_rel %>%
  mutate(
    terminal_decline = 0L,  # set to 0 (change to 1 to visualize last-year-of-life scenario)
    base_race    = factor(ref_level(mf$base_race),    levels = levels(mf$base_race)),
    base_marital = factor(ref_level(mf$base_marital), levels = levels(mf$base_marital)),
    base_living  = factor(ref_level(mf$base_living),  levels = levels(mf$base_living)),
    base_pckgr   = factor(ref_level(mf$base_pckgr),   levels = levels(mf$base_pckgr)),
    id = mf$id[1]  # not used when re.form=NA, but harmless if id is in the model frame
  )

# 5) Predict fixed-effect trajectories (population-average)
grid_rel$pred_FI <- predict(mod_sen4, newdata = grid_rel, re.form = NA)

# 6) Jump at t=0
grid_pre  <- grid_rel %>% filter(scenario == "Incident cancer at t=0", Cancer_Status == 0L)
grid_post <- grid_rel %>% filter(scenario == "Incident cancer at t=0", Cancer_Status == 1L)

y_pre_jump  <- grid_pre  %>% filter(dplyr::near(tRel, 0)) %>% pull(pred_FI)
y_post_jump <- grid_post %>% filter(dplyr::near(tRel, 0)) %>% pull(pred_FI)

# 7) Plot
ggplot() +
  geom_line(data = grid_rel %>% filter(scenario == "Never cancer (counterfactual)"),
            aes(x = tRel, y = pred_FI),
            color = "#E69F00", linetype = "dashed", linewidth = 1.2) +
  geom_line(data = grid_pre,  aes(x = tRel, y = pred_FI),
            color = "#0072B2", linewidth = 1.2) +
  geom_line(data = grid_post, aes(x = tRel, y = pred_FI),
            color = "#0072B2", linewidth = 1.2) +
  geom_segment(aes(x = 0, xend = 0, y = y_pre_jump, yend = y_post_jump),
               color = "#0072B2", linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray30") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "bottom"
  ) +
  labs(
    title = "Predicted FI Trajectories Before/After Incident Cancer (Age Time Scale)",
    subtitle = paste0(
      "Event-centered; diagnosis anchored at median age at diagnosis = ",
      round(age_dx_median, 1), " years; terminal_decline fixed at 0."
    ),
    x = "Years relative to cancer diagnosis (0 = diagnosis)",
    y = "Predicted non-cancer FI"
  )
```
