---
title: "Piecewise Linear Frailty Trajectories Before and After Incident Cancer Diagnosis"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango    
    code_folding: hide
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 2.2_self reported cancer.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Fits piecewise linear frailty trajectory models before and after self-reported incident cancer diagnosis and compares cancer and control trajectories.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 11, fig.height = 6
)

library(dplyr)
library(tidyr)
library(lubridate)
library(lme4)
library(lmerTest)
library(ggplot2)
library(knitr)
library(gtsummary)
```

## 1. Study Overview

We use self reported cancer for cancer variable, and frailty index excluding cancer as the outcome. Therefore, there is a 4-year gap between the last data point before cancer diagnosis (t=-4) and the first data point after cancer diagnosis (t=0)

**Inclusion & Exclusion Criteria:**
1. Available covariates at baseline.
2. We remove prevalent cancers at baseline.
3. We ensure all participants have at least 2 longitudinal observations with FI.
4. FI is considered missing if with less than 20 out of 27 frailty items measured.

------------------------------------------------------------------------

## 2. Data Preparation & Eligibility

```{r data_prep}
# Load data (imputed longitudinal FI dataset)
fi_long <- readRDS("/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")

# Study cycles used for repeated frailty measurement
target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")

# 1) Baseline covariates (Cycle 86)
baseline_covs <- fi_long %>%
  filter(cycle == "86") %>%
  select(
    id,
    age_months   = age_enrollment,
    base_race    = race,
    base_marital = marital_status,
    base_living  = living_arr,
    base_pckgr   = pckgr
  ) %>%
  mutate(base_age = age_months / 12) %>%
  select(-age_months) %>%
  drop_na()

# 2) Prepare longitudinal repeated-measure dataset (target cycles)
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    # Convert cycle code to calendar year (e.g., 88 -> 1988; 20 -> 2020)
    Year_num = case_when(
      as.numeric(cycle) > 50 ~ 1900 + as.numeric(cycle),
      TRUE ~ 2000 + as.numeric(cycle)
    ),
    Age = (worked_rtmnyr - dbmy09) / 12
  )

# 3) Define each participant’s baseline observation year in the analytic window
baseline_year_by_id <- fi_prep %>%
  group_by(id) %>%
  summarise(Year_base = min(Year_num), .groups = "drop")

# 4) Exclude baseline cancer (cancer present at first analytic observation)
baseline_cancer_ids <- fi_prep %>%
  inner_join(baseline_year_by_id, by = "id") %>%
  filter(Year_num == Year_base) %>%
  filter(fi_cancer == 1) %>%
  pull(id) %>%
  unique()

# 5) Identify incident cancer timing among those without baseline cancer
cancer_timing <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  filter(fi_cancer == 1) %>%
  group_by(id) %>%
  summarise(Year_Cancer = min(Year_num), .groups = "drop")

# 6) Build analytic longitudinal dataset
analysis_long <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  inner_join(baseline_year_by_id, by = "id") %>%
  left_join(cancer_timing, by = "id") %>%
  inner_join(baseline_covs, by = "id") %>%
  mutate(
    # Outcome validity rule 
    fi_score_nocancer = if_else(n_answered_nocancer < 20, NA_real_, fi_score_nocancer),

    # Follow-up time since baseline observation (years)
    time = Year_num - Year_base,

    # Time-varying post-cancer indicator
    postCancer = if_else(!is.na(Year_Cancer) & Year_num >= Year_Cancer, 1L, 0L),

    # Time since diagnosis (0 before diagnosis; increases after)
    timeAfterCancer = if_else(postCancer == 1L, Year_num - Year_Cancer, 0),

    # Useful indicators for tables/plots
    everCancer = if_else(!is.na(Year_Cancer), 1L, 0L),

    # Factor covariates
    base_race    = factor(base_race),
    base_marital = factor(base_marital),
    base_living  = factor(base_living),
    base_pckgr   = factor(base_pckgr),

    # Scale baseline age for stability
    base_age_z = as.numeric(scale(base_age))
  ) %>%
  filter(participated == 1, !is.na(fi_score_nocancer)) %>%
  group_by(id) %>%
  filter(n() >= 2) %>%
  ungroup() %>%
  arrange(id, Year_num)
```

------------------------------------------------------------------------

## 3. Sample Flow & Baseline Descriptives

```{r descriptive_tables}
# Attrition logic
n_baseline_covs <- nrow(baseline_covs)
n_in_target_cycles <- fi_prep %>% distinct(id) %>% nrow()
n_excluded_baseline_cancer <- length(unique(baseline_cancer_ids))
n_ids_final <- analysis_long %>% distinct(id) %>% nrow()
n_obs_final <- nrow(analysis_long)

attrition_data <- data.frame(
  Step = c(
    "Participants with complete baseline (1986) covariates",
    "Participants observed in any target cycles (1988–2020)",
    "Excluded (baseline cancer at first analytic observation)",
    "Final eligible participants (>=2 FI observations)",
    "Total valid observations (rows) used in models"
  ),
  N = c(n_baseline_covs, n_in_target_cycles, n_excluded_baseline_cancer, n_ids_final, n_obs_final)
)

kable(attrition_data, caption = "Table: Study Sample Flow", align = c("l", "r"))

# Baseline Table by Future Cancer Status
baseline_for_table <- analysis_long %>%
  distinct(id, everCancer, base_age, base_age_z, base_race, base_marital, base_living, base_pckgr)

baseline_for_table %>%
  select(everCancer, base_age, base_race, base_marital, base_living, base_pckgr) %>%
  tbl_summary(
    by = everCancer,
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)")
  ) %>%
  add_p() %>%
  modify_header(label = "**Baseline characteristic (1986)**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Ever incident cancer during follow-up**")
```

------------------------------------------------------------------------

## 4. Main Piecewise Linear Mixed-Effects Models

```{r fit_models}
# Fully adjusted model
mod2_full <- lmer(
  fi_score_nocancer ~ time + postCancer + timeAfterCancer +
    base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + time | id),
  data = analysis_long,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod2_full)
```

```{r extract_effects}
b <- fixef(mod2_full)
data.frame(
  quantity = c("Pre-diagnosis slope (per year)", "Immediate jump after diagnosis", "Post-diagnosis slope (per year)"),
  estimate = c(b["time"], b["postCancer"], b["time"] + b["timeAfterCancer"])
) %>% 
  kable(digits = 4, caption = "Derived slopes and jump (point estimates)")
```

------------------------------------------------------------------------

## 5. Visualization: Predicted Trajectories (Bounded Event-Centered)

We visualize the trajectories based on the data's true follow-up span. To prevent the pre-cancer modeled `time` from becoming negative and generating a flatline, we bound the minimum relative plotting time against `-tDx_median`.

```{r visualization, fig.width=12, fig.height=7}
# Compute each cancer case's time-from-baseline to diagnosis
dx_time_by_id <- analysis_long %>%
  filter(everCancer == 1L) %>%
  distinct(id, Year_base, Year_Cancer) %>%
  mutate(tDx = Year_Cancer - Year_base)

tDx_median <- median(dx_time_by_id$tDx, na.rm = TRUE)

# Calculate the TRUE maximum and minimum relative follow-up times
analysis_long_cases <- analysis_long %>%
  filter(everCancer == 1L) %>%
  mutate(tRel = Year_num - Year_Cancer)

emp_min_rel <- floor(min(analysis_long_cases$tRel, na.rm = TRUE))
emp_max_rel <- ceiling(max(analysis_long_cases$tRel, na.rm = TRUE))

# BOUND min_rel to prevent negative 'time' values in the model
min_rel <- max(emp_min_rel, -floor(tDx_median))
max_rel <- emp_max_rel

cat("Plotting follow-up range from", min_rel, "to", max_rel, "years relative to diagnosis.\n")

# 1. Never Cancer Grid
grid_never <- data.frame(tRel = seq(min_rel, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Never cancer (counterfactual)",
    time = tDx_median + tRel, 
    postCancer = 0L,
    timeAfterCancer = 0
  )

# 2. Pre-Cancer Grid (Ends exactly at 0, no jump yet)
grid_cancer_pre <- data.frame(tRel = seq(min_rel, 0, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    time = tDx_median + tRel,
    postCancer = 0L,
    timeAfterCancer = 0
  )

# 3. Post-Cancer Grid (Starts exactly at 0, with the jump)
grid_cancer_post <- data.frame(tRel = seq(0, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    time = tDx_median + tRel,
    postCancer = 1L,
    timeAfterCancer = tRel
  )

# Combine and apply covariates
grid_rel <- bind_rows(grid_never, grid_cancer_pre, grid_cancer_post) %>%
  mutate(
    base_age_z = 0,
    base_race = levels(analysis_long$base_race)[1],
    base_marital = levels(analysis_long$base_marital)[1],
    base_living = levels(analysis_long$base_living)[1],
    base_pckgr = levels(analysis_long$base_pckgr)[1]
  )

# Predict marginal means
grid_rel$pred_FI <- predict(mod2_full, newdata = grid_rel, re.form = NA)

# Re-split for clean plotting
grid_never_plot <- grid_rel %>% filter(scenario == "Never cancer (counterfactual)")
grid_cancer_pre_plot <- grid_rel %>% filter(scenario == "Incident cancer at t=0", postCancer == 0L)
grid_cancer_post_plot <- grid_rel %>% filter(scenario == "Incident cancer at t=0", postCancer == 1L)

# Get precise y-values for the vertical jump
y_pre_jump <- grid_cancer_pre_plot$pred_FI[grid_cancer_pre_plot$tRel == 0]
y_post_jump <- grid_cancer_post_plot$pred_FI[grid_cancer_post_plot$tRel == 0]

ggplot() +
  # Plot Never Cancer trajectory
  geom_line(data = grid_never_plot, aes(x = tRel, y = pred_FI), 
            color = "#E69F00", linetype = "dashed", linewidth = 1.2) +
  
  # Plot Pre-Cancer trajectory
  geom_line(data = grid_cancer_pre_plot, aes(x = tRel, y = pred_FI), 
            color = "#0072B2", linewidth = 1.2) +
  
  # Plot Post-Cancer trajectory
  geom_line(data = grid_cancer_post_plot, aes(x = tRel, y = pred_FI), 
            color = "#0072B2", linewidth = 1.2) +
  
  # Add the vertical "jump" at Time 0
  geom_segment(aes(x = 0, xend = 0, 
                   y = y_pre_jump, 
                   yend = y_post_jump), 
               color = "#0072B2", linewidth = 1.2) +
  
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray30") +
  
  # Adjust x-axis breaks to accommodate long follow-up
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "bottom"
  ) +
  labs(
    title = "Predicted FI Trajectories Before/After Incident Cancer",
    subtitle = paste0("Event-centered display assuming diagnosis at median time-to-diagnosis (", round(tDx_median, 1), " years)."),
    x = "Years relative to cancer diagnosis (0 = Diagnosis)",
    y = "Predicted non-cancer FI"
  )
```

------------------------------------------------------------------------

## 6. Extended Model: Testing Pre-Diagnosis Differences (Optional)

```{r extended_model}
mod3_interactions <- lmer(
  fi_score_nocancer ~ time + everCancer + (time * everCancer) + 
    postCancer + timeAfterCancer +
    base_age_z + base_race + base_marital + base_living + base_pckgr +
    (1 + time | id),
  data = analysis_long,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod3_interactions)
```

## 7. Visualization: Extended Model (Pre-Diagnosis Differences)

```{r extended_visualization, fig.width=12, fig.height=7}
# Re-use min_rel and max_rel defined in Section 5
# Build independent grids, ensuring everCancer indicator is mapped correctly

# 1. Never Cancer Grid
grid_never_ext <- data.frame(tRel = seq(min_rel, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Never cancer (counterfactual)",
    time = tDx_median + tRel,
    everCancer = 0L,
    postCancer = 0L,
    timeAfterCancer = 0
  )

# 2. Pre-Cancer Grid (Ends exactly at 0)
grid_cancer_pre_ext <- data.frame(tRel = seq(min_rel, 0, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    time = tDx_median + tRel,
    everCancer = 1L,
    postCancer = 0L,
    timeAfterCancer = 0
  )

# 3. Post-Cancer Grid (Starts exactly at 0)
grid_cancer_post_ext <- data.frame(tRel = seq(0, max_rel, by = 0.25)) %>%
  mutate(
    scenario = "Incident cancer at t=0",
    time = tDx_median + tRel,
    everCancer = 1L,
    postCancer = 1L,
    timeAfterCancer = tRel
  )

# Combine and map covariates
grid_ext <- bind_rows(grid_never_ext, grid_cancer_pre_ext, grid_cancer_post_ext) %>%
  mutate(
    base_age_z = 0,
    base_race = levels(analysis_long$base_race)[1],
    base_marital = levels(analysis_long$base_marital)[1],
    base_living = levels(analysis_long$base_living)[1],
    base_pckgr = levels(analysis_long$base_pckgr)[1]
  )

# Predict marginal means using the EXTENDED model
grid_ext$pred_FI <- predict(mod3_interactions, newdata = grid_ext, re.form = NA)

# Re-split for clean plotting
grid_never_plot_ext <- grid_ext %>% filter(scenario == "Never cancer (counterfactual)")
grid_cancer_pre_plot_ext <- grid_ext %>% filter(scenario == "Incident cancer at t=0", postCancer == 0L)
grid_cancer_post_plot_ext <- grid_ext %>% filter(scenario == "Incident cancer at t=0", postCancer == 1L)

# Get precise y-values for the vertical jump
y_pre_jump_ext <- grid_cancer_pre_plot_ext$pred_FI[grid_cancer_pre_plot_ext$tRel == 0]
y_post_jump_ext <- grid_cancer_post_plot_ext$pred_FI[grid_cancer_post_plot_ext$tRel == 0]

# Generate Plot
ggplot() +
  # Plot Never Cancer trajectory (Unique intercept and slope)
  geom_line(data = grid_never_plot_ext, aes(x = tRel, y = pred_FI), 
            color = "#E69F00", linetype = "dashed", linewidth = 1.2) +
  
  # Plot Pre-Cancer trajectory
  geom_line(data = grid_cancer_pre_plot_ext, aes(x = tRel, y = pred_FI), 
            color = "#0072B2", linewidth = 1.2) +
  
  # Plot Post-Cancer trajectory
  geom_line(data = grid_cancer_post_plot_ext, aes(x = tRel, y = pred_FI), 
            color = "#0072B2", linewidth = 1.2) +
  
  # Add the vertical "jump" at Time 0
  geom_segment(aes(x = 0, xend = 0, 
                   y = y_pre_jump_ext, 
                   yend = y_post_jump_ext), 
               color = "#0072B2", linewidth = 1.2) +
  
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray30") +
  
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "bottom"
  ) +
  labs(
    title = "Predicted FI Trajectories: Testing Pre-Diagnosis Differences",
    subtitle = paste0("Extended model allowing unique baseline frailty and pre-diagnosis slopes (tDx = ", round(tDx_median, 1), ")."),
    x = "Years relative to cancer diagnosis (0 = Diagnosis)",
    y = "Predicted non-cancer FI"
  )
```
