---
title: "Piecewise Non-Linear Frailty Trajectory Before and After Cancer Diagnosis"
author: "XZ"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 2_model.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Fits piecewise nonlinear frailty trajectory models before and after cancer diagnosis, including spline-based model components and visual output.
```

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 7)

# Load required libraries
library(dplyr)
library(tidyr)
library(lubridate)
library(lme4)
library(lmerTest) 
library(splines)   # For natural cubic splines ns()
library(ggplot2)
library(patchwork) # For side-by-side plot layout
library(knitr)
library(gtsummary) 
```

## 1. Study Overview & Sample Eligibility

This analysis models the physiological trajectory of frailty (excluding cancer items) before and after an incident cancer diagnosis using **Piecewise Linear Mixed-Effects Models with Natural Cubic Splines**. 

By interacting the spline with a `Phase` indicator (Pre vs. Post diagnosis), the model captures both the non-linear physiological curves and the acute "jump" in frailty strictly associated with the cancer event at Time = 0.

---

## 2. Data Preparation & Inclusion Criteria

```{r data_prep}
# Load Imputed Data
fi_long <- readRDS("/n/home06/xyzhou/Frailty/2_data analysis/before and after cancer/data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds")
target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")

# 1. Extract Baseline Covariates (Cycle 86)
baseline_covs <- fi_long %>%
  filter(cycle == "86") %>%
  select(
    id,
    age_months   = age_enrollment, 
    base_race    = race,
    base_marital = marital_status,
    base_living  = living_arr,
    base_pckgr   = pckgr
  ) %>%
  mutate(base_age = age_months / 12) %>% # Convert baseline age to years
  select(-age_months) %>%
  drop_na()

# 2. Standardize Years for Target Cycles
fi_prep <- fi_long %>%
  filter(cycle %in% target_cycles) %>%
  mutate(
    Year_num = case_when(
      as.numeric(cycle) > 50 ~ 1900 + as.numeric(cycle),
      TRUE ~ 2000 + as.numeric(cycle)
    ),
    Age = (worked_rtmnyr - dbmy09) / 12 # Calculate precise age at specific cycle
  )

# 3. Identify and Exclude Baseline Cancer
baseline_cancer_ids <- fi_prep %>%
  group_by(id) %>%
  arrange(Year_num) %>%
  slice(1) %>%
  filter(fi_cancer == 1) %>%
  pull(id)

# 4. Identify Timing of Incident Cancer
cancer_timing <- fi_prep %>%
  filter(!id %in% baseline_cancer_ids) %>%
  filter(fi_cancer == 1) %>%
  group_by(id) %>%
  summarise(Year_Cancer = min(Year_num), .groups = "drop") 

incident_cancer_ids <- cancer_timing$id

# 5. Build Final Dataset for Modeling
final_population <- fi_prep %>%
  filter(id %in% incident_cancer_ids) %>%
  inner_join(cancer_timing, by = "id") %>%
  inner_join(baseline_covs, by = "id") %>%
  mutate(
    # Continuous time variable relative to diagnosis
    Time_to_Cancer = Year_num - Year_Cancer,
    
    # Phase indicator to force the piecewise "gap" at Time = 0
    Phase = if_else(Time_to_Cancer >= 0, "Post-Diagnosis", "Pre-Diagnosis"),
    Phase = factor(Phase, levels = c("Pre-Diagnosis", "Post-Diagnosis")),
    
    # Validity rules
    fi_score_nocancer = if_else(n_answered_nocancer < 20, NA_real_, fi_score_nocancer),
    frailty_cat_nocancer = if_else(is.na(fi_score_nocancer), NA_character_, frailty_cat_nocancer),
    
    base_race = as.factor(base_race),
    base_marital = as.factor(base_marital),
    base_living = as.factor(base_living),
    base_pckgr = as.factor(base_pckgr)
  ) %>%
  filter(participated == 1, !is.na(fi_score_nocancer)) %>%
  mutate(base_age_z = as.numeric(scale(base_age))) %>% # Standardize age for LMM convergence
  group_by(id) %>% filter(n() >= 2) %>% ungroup() %>%
  arrange(id, Time_to_Cancer)
```

```{r attrition_table}
attrition_data <- data.frame(
  Step = c(
    "Participants with complete 1986 baseline covariates",
    "Incident Cancer cohort (excluding baseline cancer)",
    "Final Eligible Participants (>= 2 measures)",
    "Total valid observations for model"
  ),
  N = c(
    nrow(baseline_covs),
    length(incident_cancer_ids),
    length(unique(final_population$id)),
    nrow(final_population)
  )
)
kable(attrition_data, caption = "Table: Study Sample Flow", align = c("l", "r"))
```

---

## 3. Spline Complexity Selection (df)

We determine the optimal natural cubic spline degrees of freedom (`df`) by minimizing the Bayesian Information Criterion (BIC).

```{r select_df, fig.width=8, fig.height=4.5}
df_values <- 2:6
fit_stats <- list()

for (i in df_values) {
  form <- as.formula(
    paste0("fi_score_nocancer ~ ns(Time_to_Cancer, df = ", i, ") * Phase + ",
           "base_age_z + base_race + base_marital + base_living + base_pckgr + ",
           "(1 + Time_to_Cancer | id)")
  )
  
  tmp_mod <- lmer(form, data = final_population, 
                  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
  
  fit_stats[[as.character(i)]] <- data.frame(df = i, AIC = AIC(tmp_mod), BIC = BIC(tmp_mod))
}

df_comparison <- bind_rows(fit_stats)
optimal_df <- df_comparison$df[which.min(df_comparison$BIC)]

# Print the comparison table and the optimal df
kable(df_comparison, caption = "Model Fit Statistics by Spline Degrees of Freedom")
cat("Optimal Spline df (Lowest BIC):", optimal_df, "\n")
```

---

## 4. Main Models

```{r fit_models}
# Model 1: Crude Piecewise Spline Trajectory
mod1_crude <- lmer(
  fi_score_nocancer ~ ns(Time_to_Cancer, df = optimal_df) * Phase + 
                      (1 + Time_to_Cancer | id),
  data = final_population, 
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

# Model 2: Fully Adjusted Piecewise Spline Trajectory
mod2_full <- lmer(
  fi_score_nocancer ~ ns(Time_to_Cancer, df = optimal_df) * Phase + 
                      base_age_z + base_race + base_marital + base_living + base_pckgr + 
                      (1 + Time_to_Cancer | id),
  data = final_population, 
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(mod2_full)
```

---

## 5. Visualization (Separated Models with Piecewise Gap)

```{r visualization, fig.width=14, fig.height=6}
# 1. Determine dynamic boundaries based on the actual dataset
min_time <- floor(min(final_population$Time_to_Cancer))
max_time <- ceiling(max(final_population$Time_to_Cancer))

cat("Plotting data-driven timeframe from", min_time, "to", max_time, "years.\n")

# 2. Create Split Prediction Grids to enforce the gap at Time 0
grid_pre <- expand.grid(
  Time_to_Cancer = seq(min_time, 0, by = 0.25),
  Phase = "Pre-Diagnosis",
  base_age_z = 0, # Median/Mean age profile
  base_race = levels(final_population$base_race)[1],
  base_marital = levels(final_population$base_marital)[1],
  base_living = levels(final_population$base_living)[1],
  base_pckgr = levels(final_population$base_pckgr)[1]
)

grid_post <- expand.grid(
  Time_to_Cancer = seq(0, max_time, by = 0.25),
  Phase = "Post-Diagnosis",
  base_age_z = 0, 
  base_race = levels(final_population$base_race)[1],
  base_marital = levels(final_population$base_marital)[1],
  base_living = levels(final_population$base_living)[1],
  base_pckgr = levels(final_population$base_pckgr)[1]
)

# Combine grids and lock factor levels
grid_all <- bind_rows(grid_pre, grid_post)
grid_all$Phase <- factor(grid_all$Phase, levels = c("Pre-Diagnosis", "Post-Diagnosis"))

# 3. Predict marginal effects 
grid_all$fi_crude <- predict(mod1_crude, newdata = grid_all, re.form = NA)
grid_all$fi_full <- predict(mod2_full, newdata = grid_all, re.form = NA)

# 4. Base Plotting Function
plot_spline <- function(data, y_var, title, color_val) {
  ggplot(data, aes(x = Time_to_Cancer, y = .data[[y_var]], group = Phase)) +
    geom_vline(xintercept = 0, color = "darkgray", linetype = "dashed", linewidth = 1) +
    geom_line(linewidth = 1.5, color = color_val) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
    theme_minimal() + 
    labs(
      title = title, 
      x = "Years Relative to Cancer Diagnosis", 
      y = "Predicted Non-Cancer FI"
    ) +
    theme(plot.title = element_text(face = "bold", size = 13))
}

# 5. Generate and Combine Separate Plots
p1 <- plot_spline(grid_all, "fi_crude", "Model 1: Crude Trajectory", "#4E79A7")
p2 <- plot_spline(grid_all, "fi_full", "Model 2: Fully Adjusted Trajectory", "#E15759")

# Use patchwork to display them side-by-side
p1 + p2 + plot_annotation(
  title = paste0("Piecewise Non-Linear Frailty Trajectories (Spline df = ", optimal_df, ")"),
  subtitle = "Time 0 represents the cycle of incident cancer diagnosis. Data-driven timeframe.",
  theme = theme(plot.title = element_text(size = 17, face = "bold"),
                plot.subtitle = element_text(size = 12, color = "gray30"))
)
```
```