---
title: "Frailty Trajectory Analysis"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 3.1_baseline_drawing.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Creates baseline-based frailty trajectory visualizations and mixed-model summaries using the longitudinal frailty dataset with cancer information.
```

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
# library(gtsummary) # install.packages("gtsummary") if needed
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id               = as.character(id),
    cycle            = as.character(cycle),
    cancer_dateca    = as.numeric(as.character(cancer_dateca)),
    dtdth            = as.numeric(as.character(dtdth)),
    dbmy09           = as.numeric(as.character(dbmy09)),
    worked_rtmnyr    = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated     = as.numeric(as.character(participated)),
    age_at_death     = (dtdth - dbmy09) / 12,
    age_at_cancer    = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build Longitudinal FI Trajectory Panel

Only keep FI from the planned 4-year waves starting in 1988.

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  mutate(
    years_since_baseline = (worked_rtmnyr - first(worked_rtmnyr)) / 12
  ) %>%
  ungroup()
```

## 4. Extract 1986 Baseline Covariates

Covariates stay anchored to cycle 1986.

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build the Analytic Longitudinal Dataset

```{r build-analysis-long}
analysis_long <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  mutate(
    baseline_time = first(worked_rtmnyr),
    base_age = first(age_at_cycle),
    true_event_time = (first(cancer_dateca) - baseline_time) / 12,
    is_prevalent = if_else(!is.na(true_event_time) & true_event_time <= 0, 1L, 0L),
    n_visits = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_followup = (max(worked_rtmnyr, na.rm = TRUE) - min(worked_rtmnyr, na.rm = TRUE)) / 12,
    Ever_Cancer = if_else(!is.na(true_event_time) & true_event_time > 0, 1L, 0L),
    Cancer_Status_TimeVarying = if_else(
      Ever_Cancer == 1L & years_since_baseline >= true_event_time,
      1L,
      0L
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup > 0) %>%
  ungroup()

base_age_scaling <- analysis_long %>%
  distinct(id, base_age) %>%
  mutate(base_age_z = as.numeric(scale(base_age))) %>%
  select(id, base_age_z)

analysis_long <- analysis_long %>%
  left_join(base_age_scaling, by = "id") %>%
  mutate(
    terminal_decline = if_else(
      !is.na(years_to_death) & years_to_death >= -1 & years_to_death <= 0,
      1L,
      0L
    ),
    Group_Label = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

## 6. Check Follow-Up Structure

```{r followup-checks}
zero_fu_check <- analysis_long %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_fu = max(years_since_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu == 0)

cat("analysis_long_n =", nrow(analysis_long), "\n")
cat("patients_zero_fu =", nrow(zero_fu_check), "\n")

if (nrow(zero_fu_check) > 0) {
  print(zero_fu_check)
} else {
  cat("No zero follow-up patients remain in analysis_long.\n")
}
```

## 7. Conditional Imputation of Index Time (T0) for Controls

Strict version: - same 5-year baseline age bin - donor event time must be within the control's observed follow-up - controls with no eligible donor are dropped

```{r impute-index-time}
set.seed(2026)

patient_summary <- analysis_long %>%
  group_by(id) %>%
  summarize(
    base_age = first(base_age),
    max_followup = first(max_followup),
    is_case = first(Ever_Cancer),
    true_event_time = first(true_event_time),
    .groups = "drop"
  ) %>%
  mutate(
    age_bin = floor(base_age / 5) * 5
  )

cases_summary <- patient_summary %>%
  filter(is_case == 1, !is.na(true_event_time), true_event_time > 0)

controls_summary <- patient_summary %>%
  filter(is_case == 0)

imputed_times <- rep(NA_real_, nrow(controls_summary))

for (i in seq_len(nrow(controls_summary))) {
  current_control <- controls_summary[i, ]

  eligible_cases <- cases_summary %>%
    filter(
      age_bin == current_control$age_bin,
      true_event_time <= current_control$max_followup
    )

  if (nrow(eligible_cases) > 0) {
    imputed_times[i] <- sample(eligible_cases$true_event_time, size = 1)
  } else {
    imputed_times[i] <- NA_real_
  }
}

controls_summary$assigned_event_time <- imputed_times
cases_summary$assigned_event_time <- cases_summary$true_event_time

assigned_times_df <- bind_rows(cases_summary, controls_summary) %>%
  filter(!is.na(assigned_event_time)) %>%
  select(id, assigned_event_time) %>%
  distinct(id, .keep_all = TRUE)

cat("Controls dropped for no eligible donor =", sum(is.na(imputed_times)), "\n")
```

## 8. Descriptive Analysis Before LME

```{r descriptive-data}
baseline_fi <- analysis_long %>%
  arrange(id, years_since_baseline) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient <- analysis_long %>%
  group_by(id) %>%
  summarize(
    Group_Label = first(Group_Label),
    Ever_Cancer = first(Ever_Cancer),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    true_event_time = first(true_event_time),
    num_obs = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_fu = max(years_since_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id") %>%
  left_join(assigned_times_df, by = "id") %>%
  mutate(
    Index_Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

### Table 1

```{r table1}
table1 <- followup_by_patient %>%
  select(
    Index_Group,
    base_age,
    fi_baseline,
    base_race,
    base_marital,
    base_living,
    base_pckgr
  ) %>%
  gtsummary::tbl_summary(
    by = Index_Group,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      base_age ~ "Baseline Age at First Valid FI (years)",
      fi_baseline ~ "Baseline Frailty Index"
    ),
    missing_text = "(Missing)"
  ) %>%
  gtsummary::add_p() %>%
  gtsummary::add_overall() %>%
  gtsummary::bold_labels()

table1
```

### Follow-Up Summary

```{r followup-summary}
followup_summary <- followup_by_patient %>%
  group_by(Index_Group) %>%
  summarize(
    Total_Patients = n(),
    Mean_Obs = round(mean(num_obs, na.rm = TRUE), 2),
    Median_Obs = median(num_obs, na.rm = TRUE),
    Mean_FollowUp_Yrs = round(mean(max_fu, na.rm = TRUE), 2),
    Median_FollowUp_Yrs = round(median(max_fu, na.rm = TRUE), 2),
    Min_FollowUp_Yrs = round(min(max_fu, na.rm = TRUE), 2),
    Max_FollowUp_Yrs = round(max(max_fu, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(as.data.frame(followup_summary))

followup_wilcox <- wilcox.test(max_fu ~ Index_Group, data = followup_by_patient)
print(followup_wilcox)
```

### Assigned Index Time Summary

```{r index-time-summary}
index_time_summary <- followup_by_patient %>%
  group_by(Index_Group) %>%
  summarize(
    N = n(),
    Mean_IndexTime_Yrs = round(mean(assigned_event_time, na.rm = TRUE), 2),
    SD_IndexTime_Yrs = round(sd(assigned_event_time, na.rm = TRUE), 2),
    Median_IndexTime_Yrs = round(median(assigned_event_time, na.rm = TRUE), 2),
    Q1_IndexTime_Yrs = round(quantile(assigned_event_time, 0.25, na.rm = TRUE), 2),
    Q3_IndexTime_Yrs = round(quantile(assigned_event_time, 0.75, na.rm = TRUE), 2),
    Min_IndexTime_Yrs = round(min(assigned_event_time, na.rm = TRUE), 2),
    Max_IndexTime_Yrs = round(max(assigned_event_time, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(as.data.frame(index_time_summary))

index_time_wilcox <- wilcox.test(assigned_event_time ~ Index_Group, data = followup_by_patient)
print(index_time_wilcox)

index_time_ttest <- t.test(assigned_event_time ~ Index_Group, data = followup_by_patient)
print(index_time_ttest)
```

### Index Time Sanity Checks

```{r index-time-checks}
control_t0_check <- followup_by_patient %>%
  filter(Index_Group == "Control") %>%
  summarize(
    n_controls = n(),
    n_bad_t0 = sum(assigned_event_time > max_fu, na.rm = TRUE),
    pct_bad_t0 = round(100 * mean(assigned_event_time > max_fu, na.rm = TRUE), 2)
  )

print(as.data.frame(control_t0_check))

case_t0_check <- followup_by_patient %>%
  filter(Index_Group == "Cancer Case") %>%
  summarize(
    n_cases = n(),
    n_mismatch = sum(abs(assigned_event_time - true_event_time) > 1e-8, na.rm = TRUE)
  )

print(as.data.frame(case_t0_check))
```

### Plots

```{r descriptive-plots, fig.width=8, fig.height=5}
p_followup_density <- ggplot(followup_by_patient, aes(x = max_fu, fill = Index_Group, color = Index_Group)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Distribution of Follow-Up Time",
    x = "Maximum Follow-Up Time (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  )

print(p_followup_density)

p_index_density <- ggplot(followup_by_patient, aes(x = assigned_event_time, fill = Index_Group, color = Index_Group)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Distribution of Assigned Index Time (T0)",
    x = "Assigned Index Time Since FI Baseline (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  )

print(p_index_density)

p_index_box <- ggplot(followup_by_patient, aes(x = Index_Group, y = assigned_event_time, fill = Index_Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.25) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Assigned Index Time by Group",
    x = NULL,
    y = "Assigned Index Time Since FI Baseline (years)"
  ) +
  theme(legend.position = "none")

print(p_index_box)
```

### Pre/Post Index Observation Counts

```{r pre-post-summary}
obs_relative_to_t0 <- analysis_long %>%
  inner_join(assigned_times_df, by = "id") %>%
  mutate(
    Index_Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case")),
    Post = if_else(years_since_baseline >= assigned_event_time, 1L, 0L)
  ) %>%
  group_by(id, Index_Group) %>%
  summarize(
    n_pre = sum(Post == 0, na.rm = TRUE),
    n_post = sum(Post == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Index_Group) %>%
  summarize(
    Mean_Pre_Obs = round(mean(n_pre, na.rm = TRUE), 2),
    Mean_Post_Obs = round(mean(n_post, na.rm = TRUE), 2),
    Median_Pre_Obs = median(n_pre, na.rm = TRUE),
    Median_Post_Obs = median(n_post, na.rm = TRUE),
    .groups = "drop"
  )

print(as.data.frame(obs_relative_to_t0))
```

## 9. Build Piecewise DiD Variables

```{r build-model-data}
model_data <- analysis_long %>%
  inner_join(assigned_times_df, by = "id") %>%
  mutate(
    Group = Ever_Cancer,
    Time = years_since_baseline,
    Post = if_else(Time >= assigned_event_time, 1L, 0L),
    TimeAfter = if_else(Post == 1L, Time - assigned_event_time, 0)
  )

zero_fu_model_check <- model_data %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_fu = max(Time, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu == 0)

cat("model_data_n =", nrow(model_data), "\n")
cat("model_zero_fu =", nrow(zero_fu_model_check), "\n")

if (nrow(zero_fu_model_check) > 0) {
  print(zero_fu_model_check)
} else {
  cat("No zero follow-up patients remain in model_data.\n")
}
```

## 10. Fit the Piecewise DiD LME Model

```{r fit-lme}
fi_model <- lmer(
  fi_score_nocancer ~
    Time + Group + Time:Group +
    Post + Post:Group +
    TimeAfter + TimeAfter:Group +
    terminal_decline + base_age_z + base_race + base_marital + base_pckgr +
    (1 + Time | id),
  data = model_data,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(fi_model)
```

```{R}
plot_grid <- expand.grid(
  Time_From_T0 = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    Time = 10 + Time_From_T0,
    assigned_event_time = 10,
    Post = if_else(Time_From_T0 >= 0, 1L, 0L),
    TimeAfter = if_else(Time_From_T0 >= 0, Time_From_T0, 0),
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data$base_race)[1], levels = levels(model_data$base_race)),
    base_marital = factor(levels(model_data$base_marital)[1], levels = levels(model_data$base_marital)),
    base_pckgr = factor(levels(model_data$base_pckgr)[1], levels = levels(model_data$base_pckgr))
  )

plot_grid$pred <- predict(fi_model, newdata = plot_grid, re.form = NA)

plot_grid$Group_Label <- factor(plot_grid$Group, levels = c(0, 1), labels = c("Control", "Cancer Case"))

ggplot(plot_grid, aes(x = Time_From_T0, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Model-Implied Piecewise Linear Frailty Trajectories",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )

```

```{r}
model_plot_t0_data <- model_data %>%
  mutate(
    pred_fi = predict(fi_model, newdata = ., re.form = NA),
    Group_Label = factor(Group, levels = c(0, 1), labels = c("Control", "Cancer Case")),
    Time_From_T0 = Time - assigned_event_time,
    Time_From_T0_Rounded = round(Time_From_T0)
  ) %>%
  group_by(Group_Label, Time_From_T0_Rounded) %>%
  summarize(
    mean_pred = mean(pred_fi, na.rm = TRUE),
    se_pred = sd(pred_fi, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 20)

p_lme_traj_t0 <- ggplot(model_plot_t0_data, aes(x = Time_From_T0_Rounded, y = mean_pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(ymin = mean_pred - 1.96 * se_pred, ymax = mean_pred + 1.96 * se_pred),
    width = 0.4
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Model-Based Frailty Trajectories Aligned at Index Time (T0)",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )

print(p_lme_traj_t0)

```

```{r}
model_data_rel <- model_data %>%
  mutate(
    TimeRel = Time - assigned_event_time,
    PreTime = pmin(TimeRel, 0),
    Post = if_else(TimeRel >= 0, 1L, 0L),
    PostTime = pmax(TimeRel, 0)
  )

```

```{r}
fi_model_rel <- lmer(
  fi_score_nocancer ~
    Group +
    PreTime + Group:PreTime +
    Post + Group:Post +
    PostTime + Group:PostTime +
    terminal_decline + base_age_z + base_race + base_marital + base_pckgr +
    (1 + PreTime + PostTime | id),
  data = model_data_rel,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(fi_model_rel)

```

```{r}
plot_grid <- expand.grid(
  TimeRel = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    PreTime = pmin(TimeRel, 0),
    Post = if_else(TimeRel >= 0, 1L, 0L),
    PostTime = pmax(TimeRel, 0),
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data_rel$base_race)[1], levels = levels(model_data_rel$base_race)),
    base_marital = factor(levels(model_data_rel$base_marital)[1], levels = levels(model_data_rel$base_marital)),
    base_pckgr = factor(levels(model_data_rel$base_pckgr)[1], levels = levels(model_data_rel$base_pckgr))
  )

plot_grid$pred <- predict(fi_model_rel, newdata = plot_grid, re.form = NA)
plot_grid$Group_Label <- factor(plot_grid$Group, levels = c(0, 1), labels = c("Control", "Cancer Case"))

ggplot(plot_grid, aes(x = TimeRel, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Model-Implied Piecewise Frailty Trajectories",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )

```

```{r}
fi_model_rel_3 <- lmer(
  fi_score_nocancer ~
    Group +
    PreTime + Group:PreTime +
    PostTime + Group:PostTime +
    terminal_decline + base_age_z + base_race + base_marital + base_pckgr +
    (1 + PreTime + PostTime | id),
  data = model_data_rel,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(fi_model_rel_3)


```

```{r}
plot_grid_rel_3 <- expand.grid(
  TimeRel = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    PreTime = pmin(TimeRel, 0),
    PostTime = pmax(TimeRel, 0),
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data_rel$base_race)[1], levels = levels(model_data_rel$base_race)),
    base_marital = factor(levels(model_data_rel$base_marital)[1], levels = levels(model_data_rel$base_marital)),
    base_pckgr = factor(levels(model_data_rel$base_pckgr)[1], levels = levels(model_data_rel$base_pckgr))
  )

plot_grid_rel_3$pred <- predict(
  fi_model_rel_3,
  newdata = plot_grid_rel_3,
  re.form = NA
)

plot_grid_rel_3$Group_Label <- factor(
  plot_grid_rel_3$Group,
  levels = c(0, 1),
  labels = c("Control", "Cancer Case")
)

p_model_rel_3 <- ggplot(plot_grid_rel_3, aes(x = TimeRel, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Model-Implied Frailty Trajectories",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )

print(p_model_rel_3)

```


```{r}
diff_curve_rel_3 <- plot_grid_rel_3 %>%
  select(TimeRel, Group, pred) %>%
  tidyr::pivot_wider(
    names_from = Group,
    values_from = pred,
    names_prefix = "Group_"
  ) %>%
  mutate(
    diff_case_minus_control = Group_1 - Group_0
  )

p_diff_rel_3 <- ggplot(diff_curve_rel_3, aes(x = TimeRel, y = diff_case_minus_control)) +
  geom_line(linewidth = 1.2, color = "#de2d26") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Estimated Cancer-Control Difference in Frailty",
    x = "Years Relative to Index Time",
    y = "Predicted FI Difference (Case - Control)"
  )

print(p_diff_rel_3)
```

```{r}
model_data_rel <- model_data_rel %>%
  mutate(
    PreTime10 = PreTime / 10,
    PostTime10 = PostTime / 10,
    PreTime10_2 = PreTime10^2,
    PostTime10_2 = PostTime10^2
  )

fi_model_rel_quad <- lmer(
  fi_score_nocancer ~
    Group +
    PreTime10 + PreTime10_2 +
    Group:PreTime10 + Group:PreTime10_2 +
    PostTime10 + PostTime10_2 +
    Group:PostTime10 + Group:PostTime10_2 +
    terminal_decline + base_age_z + base_race + base_marital + base_pckgr +
    (1 + PreTime10 + PostTime10 || id),
  data = model_data_rel,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(fi_model_rel_quad)
```
```{r}
plot_grid_rel_quad <- expand.grid(
  TimeRel = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    PreTime = pmin(TimeRel, 0),
    PostTime = pmax(TimeRel, 0),
    PreTime10 = PreTime / 10,
    PostTime10 = PostTime / 10,
    PreTime10_2 = PreTime10^2,
    PostTime10_2 = PostTime10^2,
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data_rel$base_race)[1], levels = levels(model_data_rel$base_race)),
    base_marital = factor(levels(model_data_rel$base_marital)[1], levels = levels(model_data_rel$base_marital)),
    base_pckgr = factor(levels(model_data_rel$base_pckgr)[1], levels = levels(model_data_rel$base_pckgr))
  )

plot_grid_rel_quad$pred <- predict(
  fi_model_rel_quad,
  newdata = plot_grid_rel_quad,
  re.form = NA
)

plot_grid_rel_quad$Group_Label <- factor(
  plot_grid_rel_quad$Group,
  levels = c(0, 1),
  labels = c("Control", "Cancer Case")
)

p_model_rel_quad <- ggplot(plot_grid_rel_quad, aes(x = TimeRel, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Quadratic Model-Implied Frailty Trajectories",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )

print(p_model_rel_quad)
```
