---
title: "Frailty Trajectory Analysis With Primary and Risk-Set Sensitivity Models"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 3_baseline_matching.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-28
# Purpose: Fits primary and risk-set sensitivity models for frailty trajectory analysis using baseline matching or baseline-aligned comparison structures.
```

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary)
```

## 1. Load Data

```{r load-data}
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Core Time Variables

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    participated      = as.numeric(as.character(participated)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build FI Trajectory Panel Using Only Target Waves

```{r build-fi-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  mutate(
    years_since_baseline = (worked_rtmnyr - first(worked_rtmnyr)) / 12
  ) %>%
  ungroup()
```

## 4. Extract 1986 Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Longitudinal Dataset

```{r build-analysis-long}
analysis_long <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, worked_rtmnyr) %>%
  group_by(id) %>%
  mutate(
    baseline_time = first(worked_rtmnyr),
    base_age = first(age_at_cycle),
    true_event_time = (first(cancer_dateca) - baseline_time) / 12,
    is_prevalent = if_else(!is.na(true_event_time) & true_event_time <= 0, 1L, 0L),
    n_visits = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_followup = (max(worked_rtmnyr, na.rm = TRUE) - min(worked_rtmnyr, na.rm = TRUE)) / 12,
    Ever_Cancer = if_else(!is.na(true_event_time) & true_event_time > 0, 1L, 0L),
    Cancer_Status_TimeVarying = if_else(
      Ever_Cancer == 1L & years_since_baseline >= true_event_time,
      1L,
      0L
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup > 0) %>%
  ungroup()

base_age_scaling <- analysis_long %>%
  distinct(id, base_age) %>%
  mutate(base_age_z = as.numeric(scale(base_age))) %>%
  select(id, base_age_z)

analysis_long <- analysis_long %>%
  left_join(base_age_scaling, by = "id") %>%
  mutate(
    terminal_decline = if_else(
      !is.na(years_to_death) & years_to_death >= -1 & years_to_death <= 0,
      1L,
      0L
    ),
    Group_Label = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

## 6. Check Follow-Up

```{r followup-check}
zero_fu_check <- analysis_long %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(worked_rtmnyr, na.rm = TRUE),
    max_fu = max(years_since_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu == 0)

cat("analysis_long_n =", nrow(analysis_long), "\n")
cat("patients_zero_fu =", nrow(zero_fu_check), "\n")
```
## 11. Risk-Set Matching Updated to Include Baseline FI

### 11a. Build Patient-Level Matching File

```{r risk-patient-summary}
patient_summary_risk <- analysis_long %>%
  group_by(id) %>%
  summarize(
    base_age = first(base_age),
    max_followup = first(max_followup),
    is_case = first(Ever_Cancer),
    true_event_time = first(true_event_time),
    base_age_z = first(base_age_z),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_pckgr = first(base_pckgr),
    .groups = "drop"
  ) %>%
  left_join(baseline_fi_by_id, by = "id") %>%
  mutate(
    age_bin = floor(base_age / 5) * 5,
    fi_bin = ntile(baseline_fi, 4)
  )

patient_summary_risk %>%
  summarize(
    n = n(),
    missing_baseline_fi = sum(is.na(baseline_fi)),
    missing_fi_bin = sum(is.na(fi_bin))
  )
```

### 11b. 1:1 Risk-Set Matching Without Replacement

Controls are matched on: - same 5-year baseline age bin - same baseline FI quartile - still under observation at the case event time

If no control is available in the same FI quartile, the fallback keeps the same age bin and then picks the nearest baseline FI.

```{r risk-matching}
set.seed(2026)

controls_per_case <- 1

cases_risk <- patient_summary_risk %>%
  filter(
    is_case == 1,
    !is.na(true_event_time),
    true_event_time > 0,
    !is.na(baseline_fi),
    !is.na(fi_bin)
  ) %>%
  arrange(desc(true_event_time))

controls_pool <- patient_summary_risk %>%
  filter(
    is_case == 0,
    !is.na(baseline_fi),
    !is.na(fi_bin)
  ) %>%
  mutate(used = FALSE)

matched_sets <- vector("list", nrow(cases_risk))

for (i in seq_len(nrow(cases_risk))) {
  current_case <- cases_risk[i, ]

  eligible_controls <- controls_pool %>%
    filter(
      used == FALSE,
      age_bin == current_case$age_bin,
      fi_bin == current_case$fi_bin,
      max_followup >= current_case$true_event_time
    ) %>%
    mutate(
      fu_distance = abs(max_followup - current_case$max_followup),
      fi_distance = abs(baseline_fi - current_case$baseline_fi)
    ) %>%
    arrange(fu_distance, fi_distance)

  if (nrow(eligible_controls) == 0) {
    eligible_controls <- controls_pool %>%
      filter(
        used == FALSE,
        age_bin == current_case$age_bin,
        max_followup >= current_case$true_event_time
      ) %>%
      mutate(
        fu_distance = abs(max_followup - current_case$max_followup),
        fi_distance = abs(baseline_fi - current_case$baseline_fi)
      ) %>%
      arrange(fi_distance, fu_distance)
  }

  if (nrow(eligible_controls) >= controls_per_case) {
    selected_controls <- eligible_controls %>%
      slice_head(n = controls_per_case)

    controls_pool$used[match(selected_controls$id, controls_pool$id)] <- TRUE

    matched_sets[[i]] <- tibble(
      set_id = i,
      case_id = current_case$id,
      control_id = selected_controls$id,
      assigned_event_time = current_case$true_event_time,
      case_age_bin = current_case$age_bin,
      case_fi_bin = current_case$fi_bin,
      case_baseline_fi = current_case$baseline_fi,
      control_baseline_fi = selected_controls$baseline_fi,
      case_max_followup = current_case$max_followup,
      control_max_followup = selected_controls$max_followup
    )
  }
}

matched_sets_df <- bind_rows(matched_sets)

n_matched_cases <- dplyr::n_distinct(matched_sets_df$case_id)
n_matched_controls <- dplyr::n_distinct(matched_sets_df$control_id)
n_unmatched_cases <- nrow(cases_risk) - n_matched_cases

cat("Matched cases =", n_matched_cases, "\n")
cat("Matched controls =", n_matched_controls, "\n")
cat("Unmatched cases =", n_unmatched_cases, "\n")
```

### 11c. Check Matching Balance

```{r risk-balance}
matched_cases_summary <- matched_sets_df %>%
  distinct(set_id, case_id, assigned_event_time) %>%
  transmute(id = case_id, Group_Label = "Cancer Case") %>%
  left_join(patient_summary_risk, by = "id")

matched_controls_summary <- matched_sets_df %>%
  distinct(set_id, control_id, assigned_event_time) %>%
  transmute(id = control_id, Group_Label = "Control") %>%
  left_join(patient_summary_risk, by = "id")

matched_summary <- bind_rows(matched_cases_summary, matched_controls_summary)

balance_summary <- matched_summary %>%
  group_by(Group_Label) %>%
  summarize(
    N = n(),
    Mean_Base_Age = round(mean(base_age, na.rm = TRUE), 2),
    Mean_Baseline_FI = round(mean(baseline_fi, na.rm = TRUE), 4),
    SD_Baseline_FI = round(sd(baseline_fi, na.rm = TRUE), 4),
    Mean_Max_FU = round(mean(max_followup, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(as.data.frame(balance_summary))

print(t.test(baseline_fi ~ Group_Label, data = matched_summary))
```

### 11d. Build Matched Person-Level File

```{r risk-person-file}
matched_cases_df <- matched_sets_df %>%
  distinct(set_id, case_id, assigned_event_time) %>%
  transmute(
    id = case_id,
    set_id,
    matched_role = "case",
    Group = 1L,
    assigned_event_time
  )

matched_controls_df <- matched_sets_df %>%
  transmute(
    id = control_id,
    set_id,
    matched_role = "control",
    Group = 0L,
    assigned_event_time
  )

matched_people <- bind_rows(matched_cases_df, matched_controls_df)
```

### 11e. Join Back to Longitudinal Data

```{r risk-model-data}
model_data_risk <- analysis_long %>%
  inner_join(matched_people, by = "id") %>%
  mutate(
    Time = years_since_baseline,
    TimeRel = Time - assigned_event_time,
    PreTime = pmin(TimeRel, 0),
    Post = if_else(TimeRel >= 0, 1L, 0L),
    PostTime = pmax(TimeRel, 0),
    Group_Label = factor(Group, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

### 11f. Require At Least One Pre and One Post Observation

```{r risk-eligibility}
eligible_ids_risk <- model_data_risk %>%
  group_by(id) %>%
  summarize(
    n_pre = sum(TimeRel < 0, na.rm = TRUE),
    n_post = sum(TimeRel >= 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_pre >= 1, n_post >= 1) %>%
  pull(id)

model_data_risk <- model_data_risk %>%
  filter(id %in% eligible_ids_risk)

cat("Eligible matched subjects for LME =", dplyr::n_distinct(model_data_risk$id), "\n")
```

### 11g. Check Matched Index Times

```{r risk-index-summary}
risk_index_plot_data <- model_data_risk %>%
  group_by(id, Group_Label) %>%
  summarize(
    assigned_event_time = first(assigned_event_time),
    max_fu = max(Time, na.rm = TRUE),
    .groups = "drop"
  )

risk_index_summary <- risk_index_plot_data %>%
  group_by(Group_Label) %>%
  summarize(
    N = n(),
    Mean_T0 = round(mean(assigned_event_time, na.rm = TRUE), 2),
    Median_T0 = round(median(assigned_event_time, na.rm = TRUE), 2),
    Mean_FollowUp = round(mean(max_fu, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(as.data.frame(risk_index_summary))
```

```{r risk-index-plot, fig.width=8, fig.height=5}
ggplot(risk_index_plot_data, aes(x = assigned_event_time, fill = Group_Label, color = Group_Label)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Assigned Index Time Under Risk-Set Matching",
    x = "Assigned Index Time Since FI Baseline (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  )
```

### 11h. Fit Risk-Set Matched LME Model

```{r risk-lme}
fi_model_risk <- lmer(
  fi_score_nocancer ~
    Group +
    PreTime + Group:PreTime +
    Post + Group:Post +
    PostTime + Group:PostTime +
    base_age_z + base_race + base_marital + base_pckgr +
    (1 + PreTime + PostTime || id) +
    (1 | set_id),
  data = model_data_risk,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(fi_model_risk)
```

### 11i. Plot Risk-Set Matched Model

```{r risk-model-plot, fig.width=8, fig.height=5}
plot_grid_risk <- expand.grid(
  TimeRel = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    PreTime = pmin(TimeRel, 0),
    Post = if_else(TimeRel >= 0, 1L, 0L),
    PostTime = pmax(TimeRel, 0),
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data_risk$base_race)[1], levels = levels(model_data_risk$base_race)),
    base_marital = factor(levels(model_data_risk$base_marital)[1], levels = levels(model_data_risk$base_marital)),
    base_pckgr = factor(levels(model_data_risk$base_pckgr)[1], levels = levels(model_data_risk$base_pckgr)),
    set_id = model_data_risk$set_id[1],
    id = model_data_risk$id[1]
  )

plot_grid_risk$pred <- predict(fi_model_risk, newdata = plot_grid_risk, re.form = NA)
plot_grid_risk$Group_Label <- factor(plot_grid_risk$Group, levels = c(0, 1), labels = c("Control", "Cancer Case"))

ggplot(plot_grid_risk, aes(x = TimeRel, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Risk-Set Matched Model-Implied Frailty Trajectories",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )
```

## 12. Compare Key Cancer Effects Across Primary and Sensitivity Models

```{r}
fi_model_risk_casejump <- lmer(
  fi_score_nocancer ~
    Group +
    PreTime + Group:PreTime +
    Group:Post +
    PostTime + Group:PostTime +
    base_age_z + base_race + base_marital + base_pckgr +
    (1 + PreTime + PostTime || id) +
    (1 | set_id),
  data = model_data_risk,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

```

```{r}
plot_grid_casejump <- expand.grid(
  TimeRel = seq(-20, 20, by = 0.1),
  Group = c(0, 1)
) %>%
  mutate(
    PreTime = pmin(TimeRel, 0),
    Post = if_else(TimeRel >= 0, 1L, 0L),
    PostTime = pmax(TimeRel, 0),
    terminal_decline = 0L,
    base_age_z = 0,
    base_race = factor(levels(model_data_risk$base_race)[1], levels = levels(model_data_risk$base_race)),
    base_marital = factor(levels(model_data_risk$base_marital)[1], levels = levels(model_data_risk$base_marital)),
    base_pckgr = factor(levels(model_data_risk$base_pckgr)[1], levels = levels(model_data_risk$base_pckgr)),
    set_id = model_data_risk$set_id[1],
    id = model_data_risk$id[1]
  )

plot_grid_casejump$pred <- predict(
  fi_model_risk_casejump,
  newdata = plot_grid_casejump,
  re.form = NA
)

plot_grid_casejump$Group_Label <- factor(
  plot_grid_casejump$Group,
  levels = c(0, 1),
  labels = c("Control", "Cancer Case")
)

ggplot(plot_grid_casejump, aes(x = TimeRel, y = pred, color = Group_Label)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Control" = "#3182bd", "Cancer Case" = "#de2d26")) +
  theme_minimal() +
  labs(
    title = "Risk-Set Matched Trajectories With Cancer-Specific Jump Only",
    x = "Years Relative to Index Time",
    y = "Predicted Frailty Index",
    color = "Group"
  )
```
