---
title: "Frailty Trajectories by Cancer Burden: Age Scale & Stochastic Drawing"
output: html_document
---
```{r script-documentation, include=FALSE}
# Project: Frailty Trajectories Before and After Incident Cancer in the Health Professionals Follow-up Study
# Script: 4.1_age_drawing_agressive cancer.Rmd
# Author: Nemo Zhou
# Date started: Unknown (pre-existing script before documentation standard was applied)
# Date last updated: 2026-06-29 (high-burden flag now read from 7.4_cancer_subtypes.R; cancer_lal no longer referenced)
# Purpose: Uses age-scale stochastic drawing to compare frailty trajectories by cancer burden, with emphasis on aggressive cancer classifications.
```

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(lme4)
library(lmerTest)
library(gtsummary) 
```

## 1. Load Data and Define Target FI Cycles

```{r load-data}
# Update path if necessary
fi_long_merged <- readRDS(
  "~/OneDrive - Harvard University/Research/Frailty Project/Data/FI_longitudinal_1986_2020_IMPUTED_Cancer.rds"
)

target_cycles <- c("88", "92", "96", "00", "04", "08", "12", "16", "20")
```

## 2. Prepare Time Variables (Age Scale)
Ensure that your raw cancer site variables are included here so they can be used in Step 5.

```{r prep-time}
fi_time_prepped <- fi_long_merged %>%
  mutate(
    id                = as.character(id),
    cycle             = as.character(cycle),
    cancer_dateca     = as.numeric(as.character(cancer_dateca)),
    dtdth             = as.numeric(as.character(dtdth)),
    dbmy09            = as.numeric(as.character(dbmy09)),
    worked_rtmnyr     = as.numeric(as.character(worked_rtmnyr)),
    fi_score_nocancer = as.numeric(as.character(fi_score_nocancer)),
    participated      = as.numeric(as.character(participated)),
    age_at_death      = (dtdth - dbmy09) / 12,
    age_at_cancer     = (cancer_dateca - dbmy09) / 12
  )
```

## 3. Build Longitudinal FI Trajectory Panel

```{r build-trajectory}
fi_trajectory <- fi_time_prepped %>%
  filter(
    participated == 1,
    cycle %in% target_cycles,
    !is.na(fi_score_nocancer),
    !is.na(worked_rtmnyr)
  ) %>%
  mutate(
    age_at_cycle = (worked_rtmnyr - dbmy09) / 12,
    years_to_death = (worked_rtmnyr - dtdth) / 12,
    years_to_cancer = (worked_rtmnyr - cancer_dateca) / 12
  ) %>%
  arrange(id, age_at_cycle) 
```

## 4. Extract Baseline Covariates

```{r baseline-covariates}
baseline_covs <- fi_time_prepped %>%
  filter(cycle == "86") %>%
  distinct(id, .keep_all = TRUE) %>%
  transmute(
    id,
    base_race    = factor(if_else(is.na(race), "Missing", as.character(race))),
    base_marital = factor(if_else(is.na(marital_status), "Missing", as.character(marital_status))),
    base_living  = factor(if_else(is.na(living_arr), "Missing", as.character(living_arr))),
    base_pckgr   = factor(if_else(is.na(pckgr), "Missing", as.character(pckgr)))
  )
```

## 5. Build Analytic Longitudinal Dataset (Stratified by Cancer Burden)

```{r build-analysis-long}
analysis_long_age <- fi_trajectory %>%
  left_join(baseline_covs, by = "id") %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  mutate(
    base_age = first(age_at_cycle),
    true_age_at_cancer = first(age_at_cancer), 
    
    # Prevalent filter
    is_prevalent = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer <= base_age, 1L, 0L),
    
    n_visits = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = min(age_at_cycle, na.rm = TRUE),
    max_age = max(age_at_cycle, na.rm = TRUE),
    max_followup_yrs = max_age - min_age, 
    
    Ever_Cancer = if_else(!is.na(true_age_at_cancer) & true_age_at_cancer > base_age, 1L, 0L),
    
    # High-burden flag precomputed in 1_data_cleaning/7.4_cancer_subtypes.R
    # (SEER 5-yr relative survival <50%: lung, pancreas, liver, esophagus,
    # stomach, brain); carried here, not re-derived.
    is_high_burden = coalesce(is_high_burden, 0L),
    
    # Construct the 3-Level Grouping Variable
    Cancer_Burden = case_when(
      Ever_Cancer == 0 ~ "Control",
      Ever_Cancer == 1 & is_high_burden == 1 ~ "High Burden",
      Ever_Cancer == 1 & is_high_burden == 0 ~ "Low/Moderate Burden"
    )
  ) %>%
  filter(is_prevalent == 0) %>%
  filter(n_visits >= 2, max_followup_yrs > 0) %>%
  ungroup() %>%
  mutate(
    # Set Control as the reference level
    Cancer_Burden = factor(Cancer_Burden, levels = c("Control", "Low/Moderate Burden", "High Burden"))
  )
```

## 6. Check Follow-Up Structure

```{r followup-checks}
zero_fu_check <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    num_obs = n(),
    n_times = n_distinct(age_at_cycle, na.rm = TRUE),
    max_fu_yrs = max(age_at_cycle, na.rm = TRUE) - min(age_at_cycle, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(max_fu_yrs == 0)

cat("Total observations in analysis_long_age =", nrow(analysis_long_age), "\n")
cat("Patients with zero follow-up years =", nrow(zero_fu_check), "\n")
```

## 7. Stochastic Random Drawing of Index Age
For each control, randomly sample an event age from the pool of cases in their baseline age cohort. 
The drawn age must fall within the control's observed follow-up window.

```{r impute-index-age}
set.seed(2026)

# Round ages to avoid floating point precision boundary errors
patient_summary_age <- analysis_long_age %>%
  group_by(id) %>%
  summarize(
    base_age = round(first(base_age), 4),
    min_age = round(first(min_age), 4),
    max_age = round(first(max_age), 4),
    is_case = first(Ever_Cancer),
    true_age_at_cancer = round(first(true_age_at_cancer), 4),
    .groups = "drop"
  ) %>%
  mutate(
    age_bin = floor(base_age / 5) * 5 
  )

cases_summary <- patient_summary_age %>%
  filter(is_case == 1, !is.na(true_age_at_cancer), true_age_at_cancer > min_age)

controls_summary <- patient_summary_age %>%
  filter(is_case == 0)

imputed_ages <- rep(NA_real_, nrow(controls_summary))

for (i in seq_len(nrow(controls_summary))) {
  current_control <- controls_summary[i, ]
  
  # CRITICAL CONSTRAINT: Donor's cancer age must fall within the control's observed age window
  eligible_cases <- cases_summary %>%
    filter(
      age_bin == current_control$age_bin,
      true_age_at_cancer >= current_control$min_age,
      true_age_at_cancer <= current_control$max_age
    )
  
  if (nrow(eligible_cases) > 0) {
    imputed_ages[i] <- sample(eligible_cases$true_age_at_cancer, size = 1)
  } else {
    imputed_ages[i] <- NA_real_
  }
}

controls_summary$assigned_event_age <- imputed_ages
cases_summary$assigned_event_age <- cases_summary$true_age_at_cancer

assigned_ages_df <- bind_rows(cases_summary, controls_summary) %>%
  filter(!is.na(assigned_event_age)) %>%
  select(id, assigned_event_age) %>%
  distinct(id, .keep_all = TRUE)

cat("Controls dropped due to no eligible donor in their age window =", sum(is.na(imputed_ages)), "\n")
```

## 8. Generate Time Scale & Spline Variables

```{r generate-splines}
model_data_age <- analysis_long_age %>%
  inner_join(assigned_ages_df, by = "id") %>%
  mutate(
    # Center age around the assigned event age
    Age_Centered = age_at_cycle - assigned_event_age,
    
    # Create the post-event knot
    Post = if_else(Age_Centered >= 0, 1L, 0L),
    Age_Post = if_else(Post == 1L, Age_Centered, 0),
    
    # Create Z-score of Index Age to adjust intercept in the LME model
    index_age_z = as.numeric(scale(assigned_event_age))
  )

cat("Final model_data_age row count =", nrow(model_data_age), "\n")
```

## 9. Descriptive Analysis Before LME

```{r descriptive-data}
baseline_fi <- analysis_long_age %>%
  arrange(id, age_at_cycle) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, fi_baseline = fi_score_nocancer)

followup_by_patient_age <- model_data_age %>% 
  group_by(id) %>%
  summarize(
    Cancer_Burden = first(Cancer_Burden),
    Ever_Cancer = first(Ever_Cancer),
    base_age = first(base_age),
    base_race = first(base_race),
    base_marital = first(base_marital),
    base_living = first(base_living),
    base_pckgr = first(base_pckgr),
    true_age_at_cancer = first(true_age_at_cancer),
    assigned_event_age = first(assigned_event_age),
    num_obs = n_distinct(age_at_cycle, na.rm = TRUE),
    min_age = first(min_age),
    max_age = first(max_age),
    max_fu_yrs = first(max_followup_yrs), 
    .groups = "drop"
  ) %>%
  left_join(baseline_fi, by = "id") %>%
  mutate(
    Index_Group = factor(Ever_Cancer, levels = c(0, 1), labels = c("Control", "Cancer Case"))
  )
```

### Table 1: Baseline Characteristics

```{r table1}
table1_age <- followup_by_patient_age %>%
  select(
    Cancer_Burden, # Using the 3-level burden factor
    base_age,
    assigned_event_age,
    fi_baseline,
    base_race,
    base_marital,
    base_living,
    base_pckgr
  ) %>%
  gtsummary::tbl_summary(
    by = Cancer_Burden,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      base_age ~ "Age at Study Entry (years)",
      assigned_event_age ~ "Age at Cancer / Assigned Index Age",
      fi_baseline ~ "Baseline Frailty Index"
    ),
    missing_text = "(Missing)"
  ) %>%
  gtsummary::add_p() %>%
  gtsummary::add_overall() %>%
  gtsummary::bold_labels()

table1_age
```

### Constraint Sanity Checks (Should be Zero)

```{r sanity-checks}
# Verify rounding fixed the boundary violations
control_age_check <- followup_by_patient_age %>%
  filter(Index_Group == "Control") %>%
  summarize(
    n_controls = n(),
    n_bad_age_early = sum(round(assigned_event_age, 4) < round(min_age, 4), na.rm = TRUE),
    n_bad_age_late = sum(round(assigned_event_age, 4) > round(max_age, 4), na.rm = TRUE)
  )

print(as.data.frame(control_age_check))
```

### Density Plot of Assigned Index Ages

```{r descriptive-plots, fig.width=8, fig.height=5}
p_index_density_age <- ggplot(followup_by_patient_age, aes(x = assigned_event_age, fill = Cancer_Burden, color = Cancer_Burden)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  scale_fill_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  scale_color_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  theme_minimal() +
  labs(
    title = "Distribution of Age at Index Event by Cancer Burden",
    x = "Assigned Index Age (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_index_density_age)
```
```{r}
# Bin the baseline ages for conditional visualization
plot_data_binned <- followup_by_patient_age %>%
  mutate(
    age_bin_factor = cut(base_age, 
                         breaks = c(40, 50, 60, 70, 80, 100), 
                         labels = c("40-49", "50-59", "60-69", "70-79", "80+"),
                         right = FALSE)
  ) %>%
  group_by(age_bin_factor, Cancer_Burden) %>%
  filter(n() > 10) %>% # Filter out sparse bins for plotting clarity
  ungroup()

p_index_density_stratified <- ggplot(plot_data_binned, aes(x = assigned_event_age, fill = Cancer_Burden, color = Cancer_Burden)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  facet_wrap(~ age_bin_factor, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  scale_color_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Distribution of Index Age Conditioned on Baseline Age",
    subtitle = "Within each age stratum, control index ages closely reflect the respective case distributions",
    x = "Assigned Index Age (years)",
    y = "Density",
    fill = "Group",
    color = "Group"
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f4f4f4", color = NA),
    strip.text = element_text(face = "bold")
  )

print(p_index_density_stratified)
```

## 10. Run the Stratified Piecewise LME Model

```{r lme-model}
model_piecewise_stratified <- lmer(
  fi_score_nocancer ~ Cancer_Burden * Age_Centered + 
                      Cancer_Burden * Age_Post + 
                      index_age_z + 
                      base_race + base_marital + base_pckgr +
                      (1 + Age_Centered | id), 
  data = model_data_age,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(model_piecewise_stratified)

# Extract fixed effects
model_results <- as.data.frame(summary(model_piecewise_stratified)$coefficients)
model_results$Term <- rownames(model_results)
rownames(model_results) <- NULL
print(model_results)
```

## 11. Generate Stratified Predictions

```{r generate-predictions}
plot_grid_strat <- expand.grid(
  Age_Centered = seq(-15, 15, by = 0.1),
  Cancer_Burden = factor(
    c("Control", "Low/Moderate Burden", "High Burden"), 
    levels = c("Control", "Low/Moderate Burden", "High Burden")
  )
) %>%
  mutate(
    Age_Post = if_else(Age_Centered >= 0, Age_Centered, 0),
    index_age_z = 0, # Hold index age at the sample mean
    
    # Hold covariates at reference levels
    base_race = factor(levels(model_data_age$base_race)[1], levels = levels(model_data_age$base_race)),
    base_marital = factor(levels(model_data_age$base_marital)[1], levels = levels(model_data_age$base_marital)),
    base_pckgr = factor(levels(model_data_age$base_pckgr)[1], levels = levels(model_data_age$base_pckgr))
  )

# Predict marginal effects
plot_grid_strat$pred <- predict(model_piecewise_stratified, newdata = plot_grid_strat, re.form = NA)
```

## 12. Plot the Stratified Trajectories

```{r plot-trajectories, fig.width=8, fig.height=6}
p_trajectory_strat <- ggplot(plot_grid_strat, aes(x = Age_Centered, y = pred, color = Cancer_Burden)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.6) +
  
  # 3-color palette to distinguish the groups
  scale_color_manual(values = c(
    "Control" = "#3182bd", 
    "Low/Moderate Burden" = "#fd8d3c", 
    "High Burden" = "#de2d26"
  )) +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Frailty Trajectories by Cancer Burden",
    subtitle = "Centered at Age of Cancer Diagnosis (or Randomly Assigned Index Age)",
    x = "Years Relative to Index Age",
    y = "Predicted Frailty Index"
  ) +
  scale_x_continuous(breaks = seq(-15, 15, by = 2), expand = expansion(mult = c(0, 0)))

print(p_trajectory_strat)
```
